### 2018-03-19 manuscript/main.tex
lots of work

\documentclass[11pt]{article}
\usepackage{amsmath,array}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{graphicx,subcaption}
\usepackage{todonotes} %[disable]
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{placeins} %FloatBarrier
\usetikzlibrary{arrows,shapes,calc}
\tikzset{line/.style = {draw, -latex'}}
\newcommand{\abs}[1]{\left|{#1}\right|}
\newcommand{\pde}[2]{{\frac{\partial {#1}}{\partial {#2}}}}
\newcommand{\pdde}[2]{\frac{\partial^2 {#1}}{\partial {#2}^2}}
\newcommand{\de}[2]{\frac{\mathrm{d} #1}{\mathrm{d} #2}}
\newcommand{\dde}[2]{\frac{\mathrm{d}^2 {#1}}{\mathrm{d} {#2}^2}}
\newcommand{\intdef}[4]{\int\limits_{#1}^{#2} #3\, \mathrm{d}{#4}}
\crefname{appendix}{Appendix}{appendices}
\crefname{section}{Section}{sections}
\crefname{table}{Table}{tables}
\Crefname{table}{Table}{Tables}
\crefname{algorithm}{Algorithm}{algorithms}
\Crefname{algorithm}{Algorithm}{Algorithms}
\crefname{equation}{Eq.}{eqs.}
\Crefname{equation}{Equation}{Equations}
\crefname{figure}{Fig.}{figs.}
\Crefname{figure}{Figure}{Figures}
\crefname{subfigure}{Fig.}{figs.}
\Crefname{subfigure}{Figure}{figures}
\captionsetup[subfigure]{labelformat=simple}
\renewcommand{\thesubfigure}{(\alph{subfigure})}
\renewcommand{\vec}[1]{\boldsymbol{#1}}
%opening
\title{Correctly inferring causal connectivity using optogenetics}
\author{Mikkel Elle Lepper\o{}d, Konrad Paul Kording}

\begin{document}

\maketitle

\begin{abstract}
\noindent Neurons interact through spikes and a central objective of neuroscience is measuring the way how neurons causally affect one another. To probe such interactions, scientists often use optogenetics which typically leads to stimulation effects on multiple cells. This then produces a so-called confounding problem - we cannot know which of the stimulated neurons affected the activity of a given target neuron. Here we show how the resulting biases are large and how causal inference techniques, in particular instrumental variables, from econometrics can ameliorate this problem. Our approach utilizes the fact that neurons have absolute refractory times where the stimulation has no influence, and the missing stimulation acts like an effectively random variable which can then be used to disentangle causality. If we can not apply truly random stimulation to a network of neurons then causal inference techniques are needed to identify mechanisms.
\end{abstract}

\section{Introduction}
% neuroscience experiments are observational which implicates a certain set of statistical tools
% spike timing cannot give a causal estimate of connectivity, thus we need to perturbe the system
%a quest for causality and optogenetics is a main tool

%set up the drive to causality
We want mechanism. Seriously. Neuroscience is to a large part experimentally driven.

%we generally use observational studies
We use observational studies. But causality is typically wrong. <nutshell version of the mehler paper>

%set up confounding
 The experiments are extremely hard to control, even with the most accurate measurements of environmental variables and minimal genetic variance. Actually, given the vast amount of neurons and their connections in primitive mammals experiments are to be considered an observational sort. In a quest for experimental control we are now able to measure from thousands of neurons at the same time with micro-second precision by electrophysiology or with high spatial accuracy by two photon imaging. However, the large amount of unmeasured activity makes estimates of functional connectivity confounded. By confounded we mean that correlations between neuron pairs are largely influenced by the background activity. One example of this confounding is when two neurons $ A,B $ are driven by a common input $ D $, while one wants to estimate connectivity between $ A $ and $ C $. However only $ B $ and C is connected such that B and C are strongly correlated, then since $ B $ and $ C $ is correlated then $ A $ and $ C $ is correlated and a regression $ C = \beta A $ will give an erroneous $ \beta > 0 $ due to the the $ A,B $ correlation given by their common input $ D $. In this case we say the regressor $ A $ is endogenous and the regression coefficient $ \beta $ estimates the magnitude of association rather than the magnitude of causation. To estimate causal relationships between neurons, stimulation is appropriate, and if able to stimulate single neurons the ability to estimate causal relationships by regression is within grasp. However, this is experimental challenging and yield very low cell count. Therefore an optogenetic approach is more than often preferable.

%optogenetics is totally nonlocal
Within the bounds of reachable tissue for photons emitted by a light source the decay of light intensity decays proportional to $ 1/r^2 $. Furthermore if the distribution of neurons is homogeneous the probability of affecting a neuron increases in distance by $ r^2 $. This essentially makes the distribution of affected neurons flat. When performing simultaneously recording with optogenetic stimulation we again get a confounding problem when estimating the functional relationships between neuron pairs. In this case the confounding is when several neurons see the same stimulus and thus become highly correlated during stimulation yielding statistically indistinguishable functional relationships, this case is similar to the above example where $ D $ now is the optogenetic stimulation. Flat distribution of V.

%causal inference is a thing, if not done well you say stupid things
The problem of endogenous regressors is common in microeconomics and in the social sciences\todo{cite}, where observed outcomes can be driven by a large number of unknown factors. It is thus largely recognized that special care has to be taken when attempting to infer causal relationships between a putative actor and an outcome.

%example
Let's say that you want to estimate the return $ \beta $ from education $ x $ to yearly income $ y $ with the regression $ y = \beta x + u $, where $ u $ is the factors other than schooling that contributes to yearly income. One of the factors in $ u $ is a persons ability, however this may also affect schooling and thus the regressor is correlated with the error term and a regression estimate will not estimate the magnitude of causation from schooling. In this case one may use the proximity to a college or university as an instrumental variable (IV) $ z $\todo{cite card 1995, found in cameron}. There are two criterion that must be fulfilled when choosing an instrumental variable, the instrument must be (1) uncorrelated with the error term, and (2) correlated with the regressor $ x $. Then one may for instance estimate $ \beta $ by a two stage ordinary least squares (OLS) where the first stage is an OLS of the regressor and the instrument which is further used in the second stage with the dependent variable $ y $\todo{cite TSOLS}. summary sentence.

%worry is this really correct? Yes, its provable. 
worry is this really correct? Yes, its . IVs are (under the right assumptions) actually correct.

%here we
The IV technique can also be employed if one seeks to estimate the causal connectivity between neuron pairs. In this study we begin by showing how a simple network of three leaky integrate and fire (LIF) neurons can be heavily confounded under the right circumstances when simulating optogenetic targeting as direct current pulses. We then show that using the refractory period as an instrumental variable we are able to produce good estimates of causal connectivity. Furthermore we show that the method works in a large (N=1250) network of LIF neurons. With this data at hand we compare the IV method with a logistic regression.

\section{Results}
\begin{figure}
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{\textwidth} \includegraphics[scale=1]{light-decay}
\caption{} \label{fig:concept:1}
\end{subfigure}

\medskip
\begin{subfigure}{\textwidth} \includegraphics[scale=1]{opto-powerlaw}
\caption{} \label{fig:concept:2}
\end{subfigure}
\caption{Sketch of the problem at hand \label{fig:concept}}
\end{figure}
\todo[inline]{perturbation should spread by 1/r2 and all perturbed connections should be tested, here logit will fail.}
\todo[inline]{figure showing stimulating of network}
\todo[inline]{figure comparing methods with real weights}
\todo[inline]{figure showing power (true false vs n stimulations)}
\todo[inline]{data showing false positive/negative comparing stimulation patterns, comparing stimulation durations}
\todo[inline]{data showing synchrounous regular does not work (can be put together with regular stim. pattern)}
\todo[inline]{Should we show that ccg or gml give lots of false positives? without considering stimulation?}
\todo[inline]{I think I would choose a couple of examples that emulate what an experimentalist would do
(a) periodic slow (bad power not enough data)
(b) periodic fast (good power but confounded)
(c) random fast (good everything)
But I would then also redo all of them with a "weak network" model where therefore the assumptions are pretty well matched an also a "strong network" model where neurons have massive influence on one another (maybe oscillate or something) and hence the weakness assumption that is behind the IV justification breaks down }
Consider three neurons (A,B,C) where A and B receives optogenetic stimulation and only B is connected to C \cref{fig:intro}\labelcref{fig:intro:1}. The stimulation induces a strong correlation between A and B \cref{fig:intro}\labelcref{fig:intro:1,fig:intro:2,fig:intro:3,fig:intro:4}, and thus the monosynaptic connection between B and C is thus spuriously inherited between A and C confounding the system by rendering the cross correlation histograms (CCHs) between BC and AC statistically indistinguishable as seen in \cref{fig:intro}\labelcref{fig:intro:5,fig:intro:6} respectively.
\begin{figure} 
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1.3]{simple}
\caption{} \label{fig:intro:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{traces}
\caption{} \label{fig:intro:2}
\end{subfigure}

\medskip
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{raster}
\caption{} \label{fig:intro:3}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{xcorr}
\caption{} \label{fig:intro:4}
\end{subfigure}
\caption{Sketch of a simple network containing three neurons depicted in red stimulated with blue laser light trough optogenetic activation. Invoked spikes due to stimulation is depicted in (b) upper panel with corresponding voltage traces in the lower panel. The neurons A and B are stimulated multiple trials and the  histogram of all trials are shown in (c) upper panel with the corresponding raster plot for each trial in the lower panel. Horizontal and vertical axes in correlograms represents respectively time lag in ms and counts of coincident spikes in bins of 1 ms for \labelcref{fig:intro:4} \label{fig:intro}}
\end{figure}

\begin{figure}
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{xcorr_highres_BC}
\caption{} \label{fig:cchvswald:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{xcorr_highres_AC}
\caption{} \label{fig:cchvswald:2}
\end{subfigure}

\medskip
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{wald_BC}
\caption{} \label{fig:cchvswald:3}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{wald_AC}
\caption{} \label{fig:cchvswald:4}
\end{subfigure}
\caption{Cross correlograms for all neurons are depicted in (d) and the high resolution zoom of BC and AC are visualized in (e) and (f) respectively. Horizontal and vertical axes in correlograms represents respectively time lag in ms and counts of coincident spikes in bins of 0.1 ms for \labelcref{fig:cchvswald:1,fig:cchvswald:2}. \label{fig:cchvswald}}
\end{figure}

\begin{figure}
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.2]{network_stim}
\caption{} \label{fig:network-power:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth}\includegraphics[scale=.2]{wald_network}
\caption{} \label{fig:network-power:2}
\end{subfigure}

\caption{Power analysis of IV estimator. \label{fig:network-power}}
\end{figure}

\begin{figure}
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{false_positive}
\caption{} \label{fig:network-class:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{false_negative}
\caption{} \label{fig:network-class:2}
\end{subfigure}

\medskip
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{fit_wald}
\caption{} \label{fig:network-class:3}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{fit_cch}
\caption{} \label{fig:network-class:4}
\end{subfigure}
\caption{Comparing CCH and logit with IV \labelcref{fig:network-class:1,fig:network-class:2}. \label{fig:network-class}}
\end{figure}

\section{Discussion}
Spike time methods works well when network is desynchronized, small amplitude stimulation with close range. Our method opens up for the possibility of strong perturbations giving larger yield.
We are making the network results quite hard as all synapses have the same time constant such that multiple neurons are affecting the same cell at the same time for each stimulation

%Correlated refractory periods is *the* failure mode. How to randomize them

% Maybe nonlinearities in stimulation rescue old school approaches
Discuss distribution of V. Discuss 2p stimulation.  Conclude that there is absolutely no reason to hope things will work with 1p.

%Maybe some biophysics pecularity will rescue us
No way. But cite Buszaki.

%We recommend the WALD 
easily usable in the context of many experiment. Potentials for future improvement.

%brain regions versus neurons
IVs on multiple scales.

%Causal inference techniques to fix experimental issues
Broad opportunity, cite RDD, cite diff-in-diff etc, broad overview of the promise of econometrics for neuro, matching

%Constructing good IVs
Multiple illuminations, designed sticky opsins

\FloatBarrier
\section{Methods}
\subsection{Instrumental variable estimation}
A simple approximation of the connectivity strength between putative sender and receiver neurons $ x $ and $ y $ respectively can be to ignore long distance excitation and simply calculate the relation between the spike times in $ x $ and $ y $ with a regression model given by

\begin{equation}
y = \beta x + u.
\label{eq:regress}
\end{equation}
Here $ y $ is the dependent variable, $ x $ is the explanatory variable, $ \beta $ is the slope of the $ x,y $ curve and $ u $ is an unknown error term. This system follows the causal path:
\begin{center}
\begin{tikzpicture}
    \node (u) {$ u $};
    \node [above = of u] (x) {$ x $};
    \node [above right = of u] (y) {$ y $};
    \path [line] (x) -- (y);
    \path [line] (u) -- (y);
\end{tikzpicture}
\end{center}
Assuming that changes in spike times $ y $ are described by $ \beta x $ i.e. $ \de{y}{x} = \beta $ for spike times $ x $. One problem with this idea is that in a confounded system, perfectly correlated neurons will give statistically indistinguishable $ \beta $. In the extreme case where two neurons are both made to fire every time they are stimulated, they will have the same weights according to \cref{eq:regress}, after all, during stimulation  $ y=1 $ for both, even if only one of them drives the downstream neuron. Another problem is if the network state affects both the probability of a neuron to fire and also the probability of downstream neurons to fire. In this case, the network state can induce a correlation which will make the estimation highly biased. Arguably, the network state will, in all realistic models, have a dramatic influence on all neurons. In general, if multiple neurons are stimulated synchronously our estimates will be off, potentially massively so as seen in the equation
\begin{equation}
y = \beta x + u(x).
\end{equation}
Following the causal path

\begin{center}
\begin{tikzpicture}
    \node (u) {$ u $};
    \node [above = of u] (x) {$ x $};
    \node [above right = of u] (y) {$ y $};
    \path [line] (x) -- (y);
    \path [line] (u) -- (x);
    \path [line] (u) -- (y);
\end{tikzpicture}
\end{center}
We now have the relation $ \de{y}{x} = \beta + \de{u}{x} $. To get at causality we thus require some stimulation that only highlight the activity in $ y $ caused by $ x $, however optogenetic stimulation is not specific enough as the photons will activate parts of the network activity $ u $. To have a remedy, we need something that can distinguish between different neurons that are stimulated. We thus require some instrument $ s_{r} $ which is (1) uncorrelated with the network $ u $ and (2) is correlated with the regressor $ x $\todo{cite the book or whatever}. The graph with stimulation is illustrated by

\begin{center}
\begin{tikzpicture}
    \node (u) {$ u $};
    \node [above = of u] (x) {$ x $};
    \node [above right = of u] (y) {$ y $};
    \node [left = of x] (s) {$ s $};
    \path [line] (x) -- (y);
    \path [line] (u) -- (x);
    \path [line] (u) -- (y);
    \path [line] (s) -- (x);
    \path [line] (s) -- (u);
\end{tikzpicture}
\end{center}
The previous graph represents a confounded stimulation, however we may use the fact that a neuron that has fired just before the stimulation will be in absolute refractory period and hence have $ x_{s}=0 $. This introduces times where the spike from one of the stimulated neurons is missing. Neurons have low firing probability once we use small bins and are very noisy at spiking at small time scales. Thus if we assume that (1) the stimulation pattern is random (2) the network activity $ u $ is asynchronous and irregular we may use the refractory period as an instrumental variable illustrated in the following graph

\begin{center}
\begin{tikzpicture}
    \node (u) {$ u $};
    \node [above = of u] (x) {$ x $};
    \node [above right = of u] (y) {$ y $};
    \node [left = of x] (sr) {$ s_r $};
    \path [line] (x) -- (y);
    \path [line] (u) -- (x);
    \path [line] (u) -- (y);
    \path [line] (sr) -- (x);
\end{tikzpicture}
\end{center}
\todo[inline]{this will not work if you stimulate non random since then you may correlate u with sr, u is still driving x but asynchronously so sr is still uncorrelated with u.}
Here $ s_r $ represents times where the sender neuron is refractory during stimulation. This is then an estimator that compares the downstream activity when a given neuron is non-refractory with the downstream activity when it is, thus removing the confounding. The true $ \beta $ is given by
\begin{equation}
\beta_{IV} = \de{y}{s_r} / \de{x}{s_r}
\end{equation}
Since our instrument $ s_r $ is binary we may use the Wald estimator \todo{cite wald} to estimate $ \beta_{IV} $ by
\begin{equation}
 \hat{\beta}_{IV} = \frac{\bar{y}_{s} - \bar{y}_{s_r}}{\bar{x}_{s} - \bar{x}_{s_r}} = \bar{y}_{s} - \bar{y}_{s_r}
 \label{eq:wald}
\end{equation}
Here $ \bar{y}_s $ is the average number of trials where stimulating $ x $ resulted in a response in $ y $ and $ \bar{y}_{s_r} $ is the average number of trials where an unsuccessful stimulation resulted in a response in $ y $.

However the estimate of the refractory period $ s_r $ may be biased, as the network state may differ between times when a neuron is refractory and times when it is not. This issue we can also correct for by calculating $ \beta $ for baseline bins with no stimulation and no stimulation during absolute refractory period. This will linearly correct for the differences in network state that are due to refractory periods while maintaining the causal validity of \cref{eq:wald}.

\todo[inline]{Note that if we assume that the stimulation is so strong that the network effect $ u_{ns} $ has a negligible effect on $ x_{s} $ and that due to synaptic and axonal delay the effect of $ u_{s} $ is unable to affect $ x_{s} $ in it's respective stimulus response time. Then a conventional method like logistic regression will work.}

\subsection{Statistical testing on cross correlation histogram}
The statistical tests giving the probabilities $ p_{diff} $ and $ p_{fast} $ were done according to \cite{Stark2009, English2017}. Briefly, to test if the cross correlation histogram (CCH) peak was significant we employed two tests. By using the Poisson distribution with a continuity correction \cite{Stark2009} given by \cref{eq:poissoncontcor} we calculated $ p_{diff} $ by comparing the peak in positive time lag with the maximum peak in negative time lag \cite{English2017} and the difference from the CCH convoluted with a hollow Gaussian \cite{Stark2009}.
\begin{equation}
p(N|\lambda(m)) = 1 - \sum_{k=0}^{N-1}\frac{e^{-\lambda(m)}\lambda(m)^k}{k!} - \frac{e^{-\lambda(m)}\lambda(m)^N}{2N!}
\label{eq:poissoncontcor}
\end{equation}
Here $ \lambda $ represents the counts at bin $ m $ and $ N $ is the number of bins considered.
\subsection{Simulated network}
The first simple network consists of three leaky integrate and fire (LIF) neurons given in \cref{eq:LIF}:
\begin{equation}
	\tau \dot{V}_{i}(t) = -(V_{i}(t) - E_{L}) + RI_{i}(t).
    \label{eq:LIF}
\end{equation}
When the membrane potential $ V_{i} $ of neuron $ i $ reaches a threshold $ \theta $ an action potential is emitted and the depolarization is reset to the leak potential $ E_{L} $ followed by an absolute refractory period $ \tau_{rp} $. The membrane time constant is represented by $ \tau $ and $ I_{i}(t) $ denotes the synaptic currents arriving at the soma modeled as a sum of alpha functions given by
\begin{equation}
	\label{eq:syn}
	RI_{i}(t) = \sum_{j}J_{ij}\sum_{k}t\exp(-(t-t_{j}^{k})/\tau_{syn}) - D.
\end{equation}\todo{make sure delay is correctly inserted}
Different synapses are denoted by $ j = 1,...,C+C_{ext} $ with post synaptic potential (PSP) amplitude (synaptic efficacy) denoted as $ J_{ij} $. When neuron $ j $ emits it's $ k $'th spike at time $ t_{j}^k $ it arrives at neuron $ i $ at $ t = t_{j}^k + D $ where $ D $ denotes a transmission delay. Furthermore all neurons are driven by an external Poisson drive and an Gaussian noise process given in \cref{eq:channelgauss}.
\begin{equation}
	I(t) = mean + std * N_j for t_0 + j dt \leq t < t_0 + (j-1) dt 
    \label{eq:channelgauss}
\end{equation}
where $N_j$ are Gaussian random numbers with unit standard deviation and $t_0$ is 
the device onset time.
\todo[inline]{small perturbations on noisy networks will not yield a correlation between AC because the variance in stimulus response time in A will not be explained by the variance in spike times of C. Small variations of onset times in A relative to C will over many trials will wash out correlations between AC and at the same time strengthen correlations between BC. The only way to get a strong correlation between AC is to impose a short and strong stimulus effectively making AB spike coincidentally.}

%-------------------------------------------------------------------------------
%                                 REFS
%-------------------------------------------------------------------------------
\pagestyle{empty}
\bibliographystyle{apalike}
{\footnotesize\linespread{1}
\bibliography{library}}
%-------------------------------------------------------------------------------
%                                 TODOS
%-------------------------------------------------------------------------------
\newpage
\listoftodos
\end{document}

### 2018-03-19 analyse.py

@@ -1,0 +1,118 @@
+import sys
+from method import IV
+import pandas as pd
+import numpy as np
+import os.path as op
+from tools import csv_append_dict, corrcoef, coef_var
+import quantities as pq
+from exana.statistics.tools import ccg_significance
+from tqdm import tqdm
+
+data_path = 'results'
+N = 10000
+winsize = 4
+binsize_corr = 5.
+min_stim = 1.
+
+if len(sys.argv) != 2:
+    raise IOError('Usage: "python run.py parameters"')
+param_module = sys.argv[1]
+jobname = param_module.replace('.py', '')
+
+dataa = np.load(op.join(data_path, jobname + '.npz'))['data'][()]
+
+latency = dataa['params']['delay'] + dataa['params']['tau_syn_ex']
+
+conn = dataa['connections']
+spiketrains = {s['sender']: {'pop': pop, 'times': s['times']}
+               for pop in ['in', 'ex']
+               for s in dataa['spiketrains'][pop]}
+sources = [s for s in dataa['stim_nodes']['ex'] if s in spiketrains.keys()]
+# targets = dataa['nodes']['ex'] + dataa['nodes']['in']
+targets = dataa['nodes']['ex']
+targets = [s for s in targets if s in spiketrains.keys()]
+pbar = tqdm(total=len(sources) * (len(targets) - len(sources)))
+for source in sources:
+    for target in targets: #NOTE different latency in inhibitory neurons
+        if target in sources:
+            continue
+        pbar.update(1)
+        stim_times = dataa['epoch']['times'][:N + 1]
+        period = np.min(np.diff(stim_times))
+        t_stop = stim_times[-1] + period
+        source_t = spiketrains[source]['times']
+        target_t = spiketrains[target]['times']
+        spike_trains = [source_t[source_t <= t_stop],
+                        target_t[target_t <= t_stop]]
+        iv = IV(*spike_trains, stim_times,
+                 winsize=winsize, latency=latency)
+        stim_amp = dataa['stim_amps'][spiketrains[source]['pop']]
+        try:
+            lr = iv.logreg
+            logreg = float(lr.coef_)
+            logreg_intercept = float(lr.intercept_)
+        except ValueError:
+            logreg = np.nan
+            logreg_intercept = np.nan
+        try:
+            lr = iv.logreg_ns
+            logreg_ns = float(lr.coef_)
+            logreg_ns_intercept = float(lr.intercept_)
+        except ValueError:
+            logreg_ns = np.nan
[…]

### 2018-03-19 method.py

@@ -1,0 +1,122 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from matplotlib.gridspec import GridSpec, GridSpecFromSubplotSpec
+from sklearn.linear_model import LogisticRegression
+import seaborn as sns
+
+def hist_stim(stim_times, source, target, period, winsize, latency):
+    hist = np.zeros([len(stim_times), 4])
+    src = np.searchsorted
+    for idx, t in enumerate(stim_times):
+        hist[idx,:] = (
+          # stim response
+          src(source, t, 'left') <
+          src(source, t + winsize, 'right'),
+          # stim synaptic response
+          src(target, t + latency, 'left') <
+          src(target, t + latency + winsize, 'right'),
+          # no stim response
+          src(source, t + period - winsize - latency, 'left') <
+          src(source, t + period - latency, 'right'),
+          # no stim synaptic response
+          src(target, t + period - winsize, 'left') <
+          src(target, t + period, 'right'))
+    return hist
+
+class IV:
+    def __init__(self, source, target, stim_times,
+                 winsize, latency):
+        '''
+        Parameters
+        ----------
+        source : array
+            putative sender neuron
+        target : array
+            putative receiver neuron
+        stim_times : array
+            stimulation times
+        '''
+        self.period = np.min(np.diff(stim_times))
+        self.stim_times = stim_times
+        self.source = source
+        self.target = target
+        self.latency = latency
+        self.winsize = winsize
+        self.lams = np.array(
+            hist_stim(stim_times, source, target,
+                      self.period, self.winsize, self.latency))
+        self.StimRef = self.lams[:, 0] == 0
+        self.Stim = self.StimRef == False
+        self.NoStimRef = self.lams[:, 2] == 0
+        self.NoStim = self.NoStimRef == False
+
+    @property
+    def wald(self):
+        ys = self.lams[self.Stim, 1]
+        ysr = self.lams[self.StimRef, 1]
+        return(ys.mean() - ysr.mean())
+
+    @property
+    def wald_ns(self):
+        ys = self.lams[self.NoStim, 3]
+        ysr = self.lams[self.NoStimRef, 3]
+        return(ys.mean() - ysr.mean())
+
+    @property
+    def logreg(self):
+        X, Y = self.lams[:, 0], self.lams[:, 1]
[…]

### 2018-03-19 params.py

@@ -1,72 +1,0 @@
-from quantities import s, ms, mV, pF, Hz, nS, pA
-import numpy as np
-
-g = 5.0  # ratio inhibitory weight/excitatory weight
-eta = 2.0  # external rate relative to threshold rate
-epsilon = 0.1  # connection probability
-
-N_neurons = 1000
-P_ex = 0.8
-
-N_ex = int(np.ceil(N_neurons * P_ex))
-N_in = int(np.ceil(N_neurons * (1 - P_ex)))
-C_ex = int(epsilon * N_ex)  # number of excitatory synapses per neuron
-C_in = int(epsilon * N_in)  # number of inhibitory synapses per neuron
-C_tot = int(C_in + C_ex)      # total number of synapses per neuron
-
-tau_m = 20.0  # time constant of membrane potential in ms
-theta = 20.0  # membrane threshold potential in mV
-J = 0.1   # postsynaptic amplitude in mV
-nr_ports = 100  # number of receptor types
-tau_syn = [0.1 + 0.01 * i for i in range(nr_ports)]
-C_m = 1.0
-t_ref = 2.0
-J_ex = J       # amplitude of excitatory postsynaptic current
-J_in = - g * J_ex  # amplitude of inhibitory postsynaptic current
-
-nu_th = theta / (J * C_ex * tau_m)
-nu_ex = eta * nu_th
-p_rate = 1000.0 * nu_ex * C_ex
-
-parameters = {
-    'neuron'         : 'iaf_psc_exp_multisynapse',
-    'record_from'    : ["V_m"],
-    'msd'            : 1234, # Master seed
-    'J_bg'           : J_ex, # Cortical input efficacy to neurons
-    'p_rate'         : p_rate * Hz, # Cortical Poisson input rate to neurons
-    'N_neurons'      : N_neurons, # Total number of neurons
-    'N_rec'          : N_neurons, # Number of neurons to record from
-    'res'            : 0.1 * ms, # Temporal resolution for simulation Delta t
-    'delay'          : 1.5 * ms, # Synaptic delay
-    'nr_ports'       : nr_ports,
-    'P_ex'           : P_ex,
-    'num_trials'     : 10,
-    'trial_start'    : 100 * ms,
-    'trial_duration' : 5 * ms,
-    'dc_amp'         : 10 * pA,
-    'N_stim'         : 100,
-    # Excitatory parameters
-    't_ref_ex'       : t_ref, # Duration of refractory period.
[…]

### 2018-03-19 params_1.py

@@ -1,0 +1,48 @@
+from quantities import s, ms, mV, pF, Hz, nS, pA
+import numpy as np
+
+parameters = {
+    'msd'             : 1234, # Master seed
+    'num_threads'     : 4,
+    'N_neurons'       : 1250,
+    'N_ex'            : 1000,
+    'N_in'            : 250,
+    'N_rec_spike_ex'  : None,
+    'N_rec_spike_in'  : None,
+    'N_rec_state_ex'  : None,
+    'N_rec_state_in'  : None,
+    'verbose'         : False,
+    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
+    'delay'           : 1.5, # Synaptic delay in ms
+    # Neuron parameters
+    't_ref'           : 2.0, # Duration of refractory period in ms
+    'V_m'             : 0.0, # Membrane potential, initial condition in mV
+    'E_L'             : 0.0, # Leak reversal potential in mV
+    'V_reset'         : 0.0, # Reset potential of the membrane in mV
+    'tau_m'           : 20.0, # Membrane timeconstant in ms
+    'C_m'             : 1.0, # Capacity of the membrane in pF
+    'V_th'            : 20.0, # Spike threshold in mV
+    'tau_syn_ex'      : 1., # Time constants of the excitatory synaptic exponential function in ms
+    'tau_syn_in'      : 1., # Time constants of the inhibitory synaptic exponential function in ms
+    # Connection parameters
+    'eta'             : .9, # external poisson rate in Hz
+    'J'               : .2, # mV
+    'g'               : 9.9,
+    'eps'            : 0.1, # connection prob
+    'J_high'         : 2.05, # max connection strength
+    'J_low'          : 0.05,
+    'p_var'          : .5, # percentage variation of mean in lognormal dist
+    # Stimulation parameters
+    'stim_nodes_ex'   : tuple(np.arange(1,602,1)),
+    'stim_nodes_in'   : (),
+    'stim_dist'       : 'poisson',
+    'stim_amp_ex'     : 10.0, # pA
+    'stim_amp_in'     : 0.0, # pA
+    'stim_period'     : 100.0, # ms
+    'stim_max_period' : 150, # only applies to poisson
+    'stim_duration'   : 2.0, # ms
+    'stim_N'          : 30000,
+    # analysis params
+    'latency': 6,
+    'winsize': 4
+}

### 2018-03-19 params_2.py

@@ -1,0 +1,48 @@
+from quantities import s, ms, mV, pF, Hz, nS, pA
+import numpy as np
+
+parameters = {
+    'msd'             : 1221, # Master seed
+    'num_threads'     : 4,
+    'N_neurons'       : 1250,
+    'N_ex'            : 1000,
+    'N_in'            : 250,
+    'N_rec_spike_ex'  : None,
+    'N_rec_spike_in'  : None,
+    'N_rec_state_ex'  : None,
+    'N_rec_state_in'  : None,
+    'verbose'         : False,
+    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
+    'delay'           : 1.5, # Synaptic delay in ms
+    # Neuron parameters
+    't_ref'           : 2.0, # Duration of refractory period in ms
+    'V_m'             : 0.0, # Membrane potential, initial condition in mV
+    'E_L'             : 0.0, # Leak reversal potential in mV
+    'V_reset'         : 0.0, # Reset potential of the membrane in mV
+    'tau_m'           : 20.0, # Membrane timeconstant in ms
+    'C_m'             : 1.0, # Capacity of the membrane in pF
+    'V_th'            : 20.0, # Spike threshold in mV
+    'tau_syn_ex'      : 1., # Time constants of the excitatory synaptic exponential function in ms
+    'tau_syn_in'      : 1., # Time constants of the inhibitory synaptic exponential function in ms
+    # Connection parameters
+    'eta'             : .9, # external poisson rate in Hz
+    'J'               : .2, # mV
+    'g'               : 4.4,
+    'eps'            : 0.1, # connection prob
+    'J_high'         : 2.05, # max connection strength
+    'J_low'          : 0.05,
+    'p_var'          : .5, # percentage variation of mean in lognormal dist
+    # Stimulation parameters
+    'stim_nodes_ex'   : tuple(np.arange(1,602,1)),
+    'stim_nodes_in'   : (),
+    'stim_dist'       : 'poisson',
+    'stim_amp_ex'     : 10.0, # pA
+    'stim_amp_in'     : 0.0, # pA
+    'stim_period'     : 100.0, # ms
+    'stim_max_period' : 150, # only applies to poisson
+    'stim_duration'   : 2.0, # ms
+    'stim_N'          : 30000,
+    # analysis params
+    'latency': 6,
+    'winsize': 4
+}

### 2018-03-19 params_3.py

@@ -1,0 +1,48 @@
+from quantities import s, ms, mV, pF, Hz, nS, pA
+import numpy as np
+
+parameters = {
+    'msd'             : 4321, # Master seed
+    'num_threads'     : 4,
+    'N_neurons'       : 1250,
+    'N_ex'            : 1000,
+    'N_in'            : 250,
+    'N_rec_spike_ex'  : None,
+    'N_rec_spike_in'  : None,
+    'N_rec_state_ex'  : None,
+    'N_rec_state_in'  : None,
+    'verbose'         : False,
+    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
+    'delay'           : 1.5, # Synaptic delay in ms
+    # Neuron parameters
+    't_ref'           : 2.0, # Duration of refractory period in ms
+    'V_m'             : 0.0, # Membrane potential, initial condition in mV
+    'E_L'             : 0.0, # Leak reversal potential in mV
+    'V_reset'         : 0.0, # Reset potential of the membrane in mV
+    'tau_m'           : 20.0, # Membrane timeconstant in ms
+    'C_m'             : 1.0, # Capacity of the membrane in pF
+    'V_th'            : 20.0, # Spike threshold in mV
+    'tau_syn_ex'      : 1., # Time constants of the excitatory synaptic exponential function in ms
+    'tau_syn_in'      : 1., # Time constants of the inhibitory synaptic exponential function in ms
+    # Connection parameters
+    'eta'             : .9, # external poisson rate in Hz
+    'J'               : .2, # mV
+    'g'               : 3.0,
+    'eps'            : 0.1, # connection prob
+    'J_high'         : 2.05, # max connection strength
+    'J_low'          : 0.05,
+    'p_var'          : .5, # percentage variation of mean in lognormal dist
+    # Stimulation parameters
+    'stim_nodes_ex'   : tuple(np.arange(1,602,1)),
+    'stim_nodes_in'   : (),
+    'stim_dist'       : 'poisson',
+    'stim_amp_ex'     : 10.0, # pA
+    'stim_amp_in'     : 0.0, # pA
+    'stim_period'     : 100.0, # ms
+    'stim_max_period' : 150, # only applies to poisson
+    'stim_duration'   : 2.0, # ms
+    'stim_N'          : 30000,
+    # analysis params
+    'latency': 6,
+    'winsize': 4
+}

### 2018-03-19 params_AI_longstim.py

@@ -1,0 +1,46 @@
+from quantities import s, ms, mV, pF, Hz, nS, pA
+import numpy as np
+
+parameters = {
+    'msd'             : 1234, # Master seed
+    'num_threads'     : 4,
+    'N_neurons'       : 1250,
+    'N_ex'            : 1000,
+    'N_in'            : 250,
+    'N_rec_spike_ex'  : None,
+    'N_rec_spike_in'  : None,
+    'N_rec_state_ex'  : None,
+    'N_rec_state_in'  : None,
+    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
+    'delay'           : 1.5, # Synaptic delay in ms
+    'eta'             : 1.0, # external poisson rate in Hz
+    # Neuron parameters
+    't_ref'           : 2.0, # Duration of refractory period in ms
+    'V_m'             : 0.0, # Membrane potential, initial condition in mV
+    'E_L'             : 0.0, # Leak reversal potential in mV
+    'V_reset'         : 0.0, # Reset potential of the membrane in mV
+    'tau_m'           : 20.0, # Membrane timeconstant in ms
+    'C_m'             : 1.0, # Capacity of the membrane in pF
+    'V_th'            : 20.0, # Spike threshold in mV
+    'tau_syn_ex'      : 4., # Time constants of the excitatory synaptic exponential function in ms
+    'tau_syn_in'      : 4., # Time constants of the inhibitory synaptic exponential function in ms
+    # Connection parameters
+    'J'               : .5, # mV
+    'g'               : 9.4,
+    'eps'            : 0.1, # connection prob
+    'J_high'         : 15., # max connection strength (low is 0)
+    'p_var'          : .5, # percentage variation of mean in lognormal dist
+    # Stimulation parameters
+    'stim_nodes_ex'   : tuple(np.arange(1,102,1)),
+    'stim_nodes_in'   : (),
+    'stim_dist'       : 'poisson',
+    'stim_amp_ex'     : 5.0, # pA
+    'stim_amp_in'     : 0.0, # pA
+    'stim_period'     : 30.0, # ms
+    'stim_max_period' : 100, # only applies to poisson
+    'stim_duration'   : 10.0, # ms
+    'stim_N'          : 30000,
+    # channelnoise
+    'gauss_mean'      : 0.,
+    'gauss_std'       : 0.
+}

### 2018-03-19 params_AI_reglongstim.py

@@ -1,0 +1,46 @@
+from quantities import s, ms, mV, pF, Hz, nS, pA
+import numpy as np
+
+parameters = {
+    'msd'             : 1234, # Master seed
+    'num_threads'     : 4,
+    'N_neurons'       : 1250,
+    'N_ex'            : 1000,
+    'N_in'            : 250,
+    'N_rec_spike_ex'  : None,
+    'N_rec_spike_in'  : None,
+    'N_rec_state_ex'  : None,
+    'N_rec_state_in'  : None,
+    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
+    'delay'           : 1.5, # Synaptic delay in ms
+    'eta'             : 1.0, # external poisson rate in Hz
+    # Neuron parameters
+    't_ref'           : 2.0, # Duration of refractory period in ms
+    'V_m'             : 0.0, # Membrane potential, initial condition in mV
+    'E_L'             : 0.0, # Leak reversal potential in mV
+    'V_reset'         : 0.0, # Reset potential of the membrane in mV
+    'tau_m'           : 20.0, # Membrane timeconstant in ms
+    'C_m'             : 1.0, # Capacity of the membrane in pF
+    'V_th'            : 20.0, # Spike threshold in mV
+    'tau_syn_ex'      : 4., # Time constants of the excitatory synaptic exponential function in ms
+    'tau_syn_in'      : 4., # Time constants of the inhibitory synaptic exponential function in ms
+    # Connection parameters
+    'J'               : .5, # mV
+    'g'               : 9.4,
+    'eps'            : 0.1, # connection prob
+    'J_high'         : 15., # max connection strength (low is 0)
+    'p_var'          : .5, # percentage variation of mean in lognormal dist
+    # Stimulation parameters
+    'stim_nodes_ex'   : tuple(np.arange(1,102,1)),
+    'stim_nodes_in'   : (),
+    'stim_dist'       : None,
+    'stim_amp_ex'     : 5.0, # pA
+    'stim_amp_in'     : 0.0, # pA
+    'stim_period'     : 30.0, # ms
+    'stim_max_period' : 100, # only applies to poisson
+    'stim_duration'   : 10.0, # ms
+    'stim_N'          : 30000,
+    # channelnoise
+    'gauss_mean'      : 0.,
+    'gauss_std'       : 0.
+}

### 2018-03-19 params_AI_regstim.py

@@ -1,0 +1,46 @@
+from quantities import s, ms, mV, pF, Hz, nS, pA
+import numpy as np
+
+parameters = {
+    'msd'             : 1234, # Master seed
+    'num_threads'     : 4,
+    'N_neurons'       : 1250,
+    'N_ex'            : 1000,
+    'N_in'            : 250,
+    'N_rec_spike_ex'  : None,
+    'N_rec_spike_in'  : None,
+    'N_rec_state_ex'  : None,
+    'N_rec_state_in'  : None,
+    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
+    'delay'           : 1.5, # Synaptic delay in ms
+    'eta'             : 1.0, # external poisson rate in Hz
+    # Neuron parameters
+    't_ref'           : 2.0, # Duration of refractory period in ms
+    'V_m'             : 0.0, # Membrane potential, initial condition in mV
+    'E_L'             : 0.0, # Leak reversal potential in mV
+    'V_reset'         : 0.0, # Reset potential of the membrane in mV
+    'tau_m'           : 20.0, # Membrane timeconstant in ms
+    'C_m'             : 1.0, # Capacity of the membrane in pF
+    'V_th'            : 20.0, # Spike threshold in mV
+    'tau_syn_ex'      : 4., # Time constants of the excitatory synaptic exponential function in ms
+    'tau_syn_in'      : 4., # Time constants of the inhibitory synaptic exponential function in ms
+    # Connection parameters
+    'J'               : .5, # mV
+    'g'               : 9.4,
+    'eps'            : 0.1, # connection prob
+    'J_high'         : 15., # max connection strength (low is 0)
+    'p_var'          : .5, # percentage variation of mean in lognormal dist
+    # Stimulation parameters
+    'stim_nodes_ex'   : tuple(np.arange(1,602,1)),
+    'stim_nodes_in'   : (),
+    'stim_dist'       : None,
+    'stim_amp_ex'     : 5.0, # pA
+    'stim_amp_in'     : 0.0, # pA
+    'stim_period'     : 30.0, # ms
+    'stim_max_period' : 100, # only applies to poisson
+    'stim_duration'   : 2.0, # ms
+    'stim_N'          : 30000,
+    # channelnoise
+    'gauss_mean'      : 0.,
+    'gauss_std'       : 0.
+}

### 2018-03-19 params_A~I.py

@@ -1,0 +1,48 @@
+from quantities import s, ms, mV, pF, Hz, nS, pA
+import numpy as np
+
+parameters = {
+    'msd'             : 1234, # Master seed
+    'num_threads'     : 4,
+    'N_neurons'       : 1250,
+    'N_ex'            : 1000,
+    'N_in'            : 250,
+    'N_rec_spike_ex'  : None,
+    'N_rec_spike_in'  : None,
+    'N_rec_state_ex'  : None,
+    'N_rec_state_in'  : None,
+    'verbose'         : False,
+    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
+    'delay'           : 1.5, # Synaptic delay in ms
+    # Neuron parameters
+    't_ref'           : 2.0, # Duration of refractory period in ms
+    'V_m'             : 0.0, # Membrane potential, initial condition in mV
+    'E_L'             : 0.0, # Leak reversal potential in mV
+    'V_reset'         : 0.0, # Reset potential of the membrane in mV
+    'tau_m'           : 20.0, # Membrane timeconstant in ms
+    'C_m'             : 1.0, # Capacity of the membrane in pF
+    'V_th'            : 20.0, # Spike threshold in mV
+    'tau_syn_ex'      : 4., # Time constants of the excitatory synaptic exponential function in ms
+    'tau_syn_in'      : .1, # Time constants of the inhibitory synaptic exponential function in ms
+    # Connection parameters
+    'eta'             : .7, # external poisson rate in Hz
+    'J'               : .5, # mV
+    'g'               : 3.9,
+    'eps'            : 0.1, # connection prob
+    'J_high'         : 2.05, # max connection strength
+    'J_low'          : 0.05,
+    'p_var'          : .5, # std of lognormal dist
+    # Stimulation parameters
+    'stim_nodes_ex'   : tuple(np.arange(1,602,1)),
+    'stim_nodes_in'   : (),
+    'stim_dist'       : 'poisson',
+    'stim_amp_ex'     : 10.0, # pA
+    'stim_amp_in'     : 0.0, # pA
+    'stim_period'     : 30.0, # ms
+    'stim_max_period' : 100, # only applies to poisson
+    'stim_duration'   : 2.0, # ms
+    'stim_N'          : 30000,
+    # analysis params
+    'latency': 6,
+    'winsize': 4
+}

### 2018-03-19 params_SI.py

@@ -1,0 +1,48 @@
+from quantities import s, ms, mV, pF, Hz, nS, pA
+import numpy as np
+
+parameters = {
+    'msd'             : 1234, # Master seed
+    'num_threads'     : 4,
+    'N_neurons'       : 1250,
+    'N_ex'            : 1000,
+    'N_in'            : 250,
+    'N_rec_spike_ex'  : None,
+    'N_rec_spike_in'  : None,
+    'N_rec_state_ex'  : None,
+    'N_rec_state_in'  : None,
+    'verbose'         : False,
+    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
+    'delay'           : 1.5, # Synaptic delay in ms
+    # Neuron parameters
+    't_ref'           : 2.0, # Duration of refractory period in ms
+    'V_m'             : 0.0, # Membrane potential, initial condition in mV
+    'E_L'             : 0.0, # Leak reversal potential in mV
+    'V_reset'         : 0.0, # Reset potential of the membrane in mV
+    'tau_m'           : 20.0, # Membrane timeconstant in ms
+    'C_m'             : 1.0, # Capacity of the membrane in pF
+    'V_th'            : 20.0, # Spike threshold in mV
+    'tau_syn_ex'      : .1, # Time constants of the excitatory synaptic exponential function in ms
+    'tau_syn_in'      : .1, # Time constants of the inhibitory synaptic exponential function in ms
+    # Connection parameters
+    'eta'             : .7, # external poisson rate in Hz
+    'J'               : .5, # mV
+    'g'               : 2.5,
+    'eps'            : 0.1, # connection prob
+    'J_high'         : 2.05, # max connection strength
+    'J_low'          : 0.05,
+    'p_var'          : .5, # percentage variation of mean in lognormal dist
+    # Stimulation parameters
+    'stim_nodes_ex'   : tuple(np.arange(1,602,1)),
+    'stim_nodes_in'   : (),
+    'stim_dist'       : 'poisson',
+    'stim_amp_ex'     : 2.0, # pA
+    'stim_amp_in'     : 0.0, # pA
+    'stim_period'     : 30.0, # ms
+    'stim_max_period' : 100, # only applies to poisson
+    'stim_duration'   : 2.0, # ms
+    'stim_N'          : 30000,
+    # analysis params
+    'latency': 3,
+    'winsize': 3
+}

### 2018-03-19 params_addnoise.py

@@ -1,0 +1,53 @@
+from quantities import s, ms, mV, pF, Hz, nS, pA
+import numpy as np
+
+parameters = {
+    'msd'             : 1234, # Master seed
+    'num_threads'     : 4,
+    'N_neurons'       : 1250,
+    'N_ex'            : 1000,
+    'N_in'            : 250,
+    'N_rec_spike_ex'  : None,
+    'N_rec_spike_in'  : None,
+    'N_rec_state_ex'  : None,
+    'N_rec_state_in'  : None,
+    'verbose'         : False,
+    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
+    'delay'           : 1.5, # Synaptic delay in ms
+    # Neuron parameters
+    't_ref'           : 2.0, # Duration of refractory period in ms
+    'V_m'             : 0.0, # Membrane potential, initial condition in mV
+    'E_L'             : 0.0, # Leak reversal potential in mV
+    'V_reset'         : 0.0, # Reset potential of the membrane in mV
+    'tau_m'           : 20.0, # Membrane timeconstant in ms
+    'C_m'             : 1.0, # Capacity of the membrane in pF
+    'V_th'            : 20.0, # Spike threshold in mV
+    'tau_syn_ex'      : .1, # Time constants of the excitatory synaptic exponential function in ms
+    'tau_syn_in'      : .1, # Time constants of the inhibitory synaptic exponential function in ms
+    # Connection parameters
+    'eta'             : .7, # external poisson rate in Hz
+    'J'               : .5, # mV
+    'g'               : 2.5,
+    'eps'            : 0.1, # connection prob
+    'J_high'         : 15., # max connection strength (low is 0)
+    'p_var'          : .5, # percentage variation of mean in lognormal dist
+    # Stimulation parameters
+    'stim_nodes_ex'   : tuple(np.arange(1,402,1)),
+    'stim_nodes_in'   : (),
+    'stim_dist'       : 'poisson',
+    'stim_amp_ex'     : 10.0, # pA
+    'stim_amp_in'     : 0.0, # pA
+    'stim_period'     : 30.0, # ms
+    'stim_max_period' : 100, # only applies to poisson
+    'stim_duration'   : 2.0, # ms
+    'stim_N'          : 30000,
+    # channelnoise
+    'gauss_mean'      : 0.,
+    'gauss_std'       : 0.,
+    # pulse packets
[…]

### 2018-03-19 run.py

@@ -1,16 +1,17 @@
+import sys
 from simulator import Simulator
-from params import parameters
-import time
+import imp
+import os.path as op
 
-working_name = 'cond'
+if len(sys.argv) != 2:
+    raise IOError('Usage: "python run.py parameters"')
 
-sim = Simulator(parameters, fname=working_name)
-sim.set_kernel()
-sim.set_neurons()
-sim.set_connections()
-sim.set_stimulation()
-sim.set_spike_rec()
-sim.set_state_rec()
-sim.simulate_trials()
-sim.save_data(overwrite=True)
-sim.save_raster()
+param_module = sys.argv[1]
+jobname = param_module.replace('.py', '')
+currdir = op.dirname(op.abspath(__file__))
+f, p, d = imp.find_module(jobname, [currdir])
+parameters = imp.load_module(jobname, f, p, d).parameters
+
+sim = Simulator(parameters, mpi=False, data_path='results',
+                fname=jobname)
+sim.simulate(save=True, progress_bar=True)

### 2018-03-19 simulator.py

@@ -2,23 +2,51 @@ import numpy as np
 import nest
 import pandas as pd
 import quantities as pq
-import neo
 import os
 import shutil
 import matplotlib.pyplot as plt
 from pprint import pprint
 import json
+import time
+import copy
+try:
+    from tqdm import tqdm
+    PBAR = True
+except ImportError:
+    PBAR = False
+try:
+    from mpi4py import MPI
+    HAS_MPI = True
+except ImportError:
+    HAS_MPI = False
+
+nest.set_verbosity(100)
+
+def hasattrs(o, *args):
+    if any(hasattr(o, a) for a in args):
+        return True
+    else:
+        return False
 
 
-def hasattrs(object, strings):
-    if isinstance(strings, str):
-        return hasattr(object, strings)
-    for string in strings:
-        try:
-            getattr(object, string)
-        except:
-            return False
-    return True
+def LambertWm1(x):
+    import nest
+    nest.sli_push(x)
+    nest.sli_run('LambertWm1')
+    y = nest.sli_pop()
+    return y
+
+
+def PSPnorm(tau_m, C_m, tau_syn):
+    a = (tau_m / tau_syn)
+    b = (1.0 / tau_syn - 1.0 / tau_m)
+
+    # time of maximum
+    t_max = 1.0 / b * (-LambertWm1(-np.exp(-1.0 / a) / a) - 1.0 / a)
+    # maximum of PSP for current of unit amplitude
+    return (np.exp(1.0) / (tau_syn * C_m * b) *
+            ((np.exp(-t_max / tau_m) - np.exp(-t_max / tau_syn)) / b -
+             t_max * np.exp(-t_max / tau_syn)))
 
 
 def convert_params(params):
@@ -39,12 +67,6 @@ def convert_params(params):
         raise ValueError('Parameters not sccessfully converted.')
 
 
-def dict_equal(d1, d2):
-    k1, k2 = [set(d.keys()) for d in (d1, d2)]
-    intersect = k1.intersection(k2)
-    return set(o for o in intersect if d2[o] != d1[o]) == set()
-
-
 def rtrim(val, trim):
     if not val.endswith(trim):
         return val
@@ -52,8 +74,25 @@ def rtrim(val, trim):
         return val[:len(val) - len(trim)]
 
 
+def poisson_clipped(N, period, low, high):
+    poisson = []
+    while len(poisson) < N:
+        p = - np.log(1 - np.random.uniform(0, 1)) * period
[…]

### 2018-03-19 sweep.py

@@ -1,0 +1,88 @@
+import pandas as pd
+import numpy as np
+import quantities as pq
+import os.path as op
+from tools import corrcoef, csv_append_dict
+from params_AI import parameters
+from tqdm import tqdm
+from simulator import Simulator
+
+
+parameters.update({
+    'stim_N': 9,
+    'stim_period': 500,
+    'stim_amp_ex': 0.0,
+    'stim_amp_in': 0.0,
+    'stim_dist': None,
+})
+
+simtime_s = (parameters['stim_N'] + 1) * parameters['stim_period'] / 1000
+rate_th = 1
+spike_n_th = rate_th * simtime_s
+binsize_corr = 5.
+N_corr = 100
+
+
+def sweep(a, b, fname, amin, amax, astep, bmin, bmax, bstep, **kwargs):
+    print('Sweeping ' + fname)
+    arange = np.arange(amin, amax, astep)
+    brange = np.arange(bmin, bmax, bstep)
+    pbar = tqdm(total=len(arange) * len(brange))
+    for _a in arange:
+        for _b in brange:
+            pbar.update(1)
+            par = {a: _a, b: _b}
+            if kwargs:
+                par.update(kwargs)
+            sim = Simulator(parameters, **par)
+            sim.simulate(save=False, raster=False)
+            p = sim.data['params']
+            t_stop = sim.data['params']['status']['time']
+            spiketrains = [s['times'] for s in sim.data['spiketrains']['ex']
+                           if len(s['times']) > spike_n_th]
+            if len(spiketrains) > 1:
+                cv = np.array([np.diff(t).std()/np.diff(t).mean()
+                               for t in spiketrains])
+                ccg = corrcoef(spiketrains[:N_corr], t_stop, binsize_corr)
+                m = np.triu(ccg, k=1)
+                p.update({
+                    'CC_mean': np.nanmean(m),
+                    'CC_min': np.nanmin(m),
+                    'CC_max': np.nanmax(m),
+                    'CV_mean': cv.mean(),
+                    'CV_min': cv.min(),
+                    'CV_max': cv.max()})
[…]

### 2018-03-19 tools.py

@@ -1,0 +1,175 @@
+import os.path as op
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+
+
+def fix_figure(padLegend=None, padHeight=0):
+    plt.draw() #to know size of legend
+    fig = plt.gcf()
+    ax = plt.gca()
+    dpi = fig.get_dpi()
+    if padLegend is None:
+        padLegend = ax.get_legend().get_frame().get_width() / dpi
+    else:
+        assert isinstance(padLegend, float)
+
+    widthAx, heightAx = fig.get_size_inches()
+    pos = ax.get_position()
+    padLeft   = pos.x0 * widthAx
+    padBottom = pos.y0 * heightAx
+    padTop    = (1 - pos.y0 - pos.height) * widthAx
+    padRight  = (1 - pos.x0 - pos.width) * heightAx
+
+    widthTot = widthAx + padLeft + padRight + padLegend
+    heightTot = heightAx + padTop + padBottom
+
+    # set figure size and ax position
+    fig.set_size_inches(widthTot, heightTot)
+    ax.set_position([padLeft / widthTot, padBottom / heightTot,
+                     widthAx / widthTot, heightAx / heightTot + padHeight])
+    plt.draw()
+
+def savefig(fig, fname, **kwargs):
+    fig.savefig(op.join('manuscript',fname + '.pdf'), **kwargs)
+
+def coef_var(spike_trains):
+    """
+    Calculate the coefficient of variation in inter spike interval (ISI)
+    distribution over several spike_trains
+
+    Parameters
+    ----------
+    spike_trains : list
+        contains spike trains where each spike train is an array of spike times
+
+    Returns
+    -------
+    out : list
+        Coefficient of variations for each spike_train, nan if len(spike_train) == 0
+
+    Examples
+    --------
+    >>> np.random.seed(12345)
+    >>> spike_trains = [np.arange(10), np.random.random((10))]
+    >>> print('{d[0]:.2f}, {d[1]:.2f}'.format(d=coeff_var(spike_trains)))
+    0.00, -9.53
+
+    """
+    cvs = []
+    for spike_train in spike_trains:
+        isi = np.diff(spike_train)
+        if len(isi) > 0:
+            cvs.append(np.std(isi) / np.mean(isi))
+        else:
+            cvs.append(np.nan)
+    return cvs
+
+
+def corrcoef(spike_trains, t_stop, binsize=10.):
+    """
[…]

### 2018-03-19 trun.py

@@ -1,0 +1,125 @@
+import matplotlib
+matplotlib.use('TkAgg')
+import matplotlib.pyplot as plt
+from simulator import Simulator
+import pandas as pd
+import numpy as np
+import quantities as pq
+import neo
+from tools import corr_coef
+from matplotlib.gridspec import GridSpec, GridSpecFromSubplotSpec
+
+par = {
+    'msd'             : 1234, # Master seed
+    'num_threads'     : 4,
+    'N_neurons'       : 1250,
+    'N_ex'            : 1000,
+    'N_in'            : 250,
+    'N_rec'           : None,
+    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
+    'delay'           : 1.5, # Synaptic delay in ms
+    'eta'             : 1., # external poisson rate in Hz
+    # Neuron parameters
+    't_ref'           : 2.0, # Duration of refractory period in ms
+    'V_m'             : 0.0, # Membrane potential, initial condition in mV
+    'E_L'             : 0.0, # Leak reversal potential in mV
+    'V_reset'         : 0.0, # Reset potential of the membrane in mV
+    'tau_m'           : 20.0, # Membrane timeconstant in ms
+    'C_m'             : 1.0, # Capacity of the membrane in pF
+    'V_th'            : 20.0, # Spike threshold in mV
+    'tau_syn_ex'      : 4., # Time constants of the excitatory synaptic exponential function in ms
+    'tau_syn_in'      : 4., # Time constants of the inhibitory synaptic exponential function in ms
+    # Connection parameters
+    'J'               : .1,
+    'g'               : 4,
+    'eps'            : 0.1, # connection prob
+    'J_high'         : 6., # max connection strength (low is 0)
+    'p_var'          : .5, # percentage variation of mean in lognormal dist
+    # Stimulation parameters
+    'stim_nodes_ex'   : tuple(np.arange(1,102,1)),
+    'stim_nodes_in'   : (),
+    'stim_dist'       : 'poisson',
+    'stim_amp_ex'     : 0.0, # pA
+    'stim_amp_in'     : 0.0, # pA
+    'stim_period'     : 30.0, # ms
+    'stim_max_period' : 100,
+    'stim_duration'   : 2.0, # ms
[…]

### 2018-03-25 manuscript/main.tex
updates

@@
 \section{Results}
-\begin{figure}
-\makeatletter
-\renewcommand\p@subfigure{}
-\makeatother
-\begin{subfigure}{\textwidth} \includegraphics[scale=1]{light-decay}
-\caption{} \label{fig:concept:1}
-\end{subfigure}
-
-\medskip
-\begin{subfigure}{\textwidth} \includegraphics[scale=1]{opto-powerlaw}
-\caption{} \label{fig:concept:2}
-\end{subfigure}
-\caption{Sketch of the problem at hand \label{fig:concept}}
+\begin{figure}\includegraphics[scale=1]{opto-powerlaw}
+\caption{Spatial extent of optogenetic stimulus.
+Due to scattering and geometric loss the light intensity (I cyan) plotted as percentage of intensity exiting the optogenetic fiber follows approximately an inverse square law $ z^{-2} $ where $ z $ is the distance from the fiber; see methods.
+If neurons are uniformly distributed the number of optogenetically affected neurons increase by $ z^{2} $ (N black) rendering the probability of activating a neuron approximately constant (IN blue).
+\label{fig:concept}}
 \end{figure}
-\todo[inline]{perturbation should spread by 1/r2 and all perturbed connections should be tested, here logit will fail.}
-\todo[inline]{figure showing stimulating of network}
-\todo[inline]{figure comparing methods with real weights}
-\todo[inline]{figure showing power (true false vs n stimulations)}
-\todo[inline]{data showing false positive/negative comparing stimulation patterns, comparing stimulation durations}
 \todo[inline]{data showing synchrounous regular does not work (can be put together with regular stim. pattern)}
-\todo[inline]{Should we show that ccg or gml give lots of false positives? without considering stimulation?}
 \todo[inline]{I think I would choose a couple of examples that emulate what an experimentalist would do
@@
 Consider three neurons (A,B,C) where A and B receives optogenetic stimulation and only B is connected to C \cref{fig:intro}\labelcref{fig:intro:1}.
-The stimulation induces a strong correlation between A and B \cref{fig:intro}\labelcref{fig:intro:1,fig:intro:2,fig:intro:3,fig:intro:4}, and thus the monosynaptic connection between B and C is thus spuriously inherited between A and C confounding the system by rendering the cross correlation histograms (CCHs) between BC and AC statistically indistinguishable as seen in \cref{fig:intro}\labelcref{fig:intro:5,fig:intro:6} respectively.
+The stimulation induces a strong correlation between A and B \cref{fig:intro}\labelcref{fig:intro:1,fig:intro:2,fig:intro:3}, and thus the monosynaptic connection between B and C is thus spuriously inherited between A and C confounding the system by rendering the cross correlation histograms (CCHs) between BC and AC statistically indistinguishable as seen in \cref{fig:intro}\labelcref{fig:intro:3} middle and lower right panel respectively.
 \begin{figure} 
@@
 \makeatother
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1.3]{simple}
+\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{simple}
 \caption{} \label{fig:intro:1}
 \end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{traces}
+\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{raster}
 \caption{} \label{fig:intro:2}
@@
 \medskip
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{raster}
+\begin{subfigure}{\textwidth}\centering \includegraphics[scale=1]{xcorr}
 \caption{} \label{fig:intro:3}
-\end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{xcorr}
-\caption{} \label{fig:intro:4}
 \end{subfigure}
-\caption{Sketch of a simple network containing three neurons depicted in red stimulated with blue laser light trough optogenetic activation.
-Invoked spikes due to stimulation is depicted in (b) upper panel with corresponding voltage traces in the lower panel.
-The neurons A and B are stimulated multiple trials and the  histogram of all trials are shown in (c) upper panel with the corresponding raster plot for each trial in the lower panel.
-Horizontal and vertical axes in correlograms represents respectively time lag in ms and counts of coincident spikes in bins of 1 ms for \labelcref{fig:intro:4} \label{fig:intro}}
-\end{figure}
-
-\begin{figure}
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{xcorr_highres_BC}
-\caption{} \label{fig:cchvswald:1}
-\end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{xcorr_highres_AC}
-\caption{} \label{fig:cchvswald:2}
-\end{subfigure}
-
-\medskip
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{wald_BC}
-\caption{} \label{fig:cchvswald:3}
-\end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{wald_AC}
-\caption{} \label{fig:cchvswald:4}
-\end{subfigure}
-\caption{Cross correlograms for all neurons are depicted in (d) and the high resolution zoom of BC and AC are visualized in (e) and (f) respectively.
-Horizontal and vertical axes in correlograms represents respectively time lag in ms and counts of coincident spikes in bins of 0.1 ms for \labelcref{fig:cchvswald:1,fig:cchvswald:2}.
-\label{fig:cchvswald}}
+\caption{Sketch of a simple network containing three neurons depicted in red stimulated with blue laser light due to optogenetic activation.
+The neurons A and B are stimulated 30000 trials and the  histogram of all trials are shown in \labelcref{fig:intro:2} upper panel with the corresponding raster plot for each trial in the lower panel.
+Horizontal and vertical axes in correlation histograms represents time lag in ms and counts of coincident spikes in bins of 1 ms in \labelcref{fig:intro:3} \label{fig:intro}}
 \end{figure}
@@
 \makeatother
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.2]{network_stim}
-\caption{} \label{fig:network-power:1}
+\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{wald_BC}
+\caption{} \label{fig:cchvswald:1}
 \end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth}\includegraphics[scale=.2]{wald_network}
-\caption{} \label{fig:network-power:2}
+\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{wald_AC}
+\caption{} \label{fig:cchvswald:2}
 \end{subfigure}
-
-\caption{Power analysis of IV estimator. \label{fig:network-power}}
+\caption{Wald estimator correctly estimates causal connectivity.
+The confounded network of three neurons (A,B,C) \cref{fig:intro}\labelcref{fig:intro:1} is disentangled with the Wald estimator quickly converging to $ BC \approx 0.2, AC \approx 0 $.
+Insets represent high resolution zoom of cross correlation histograms of BC and AC respectively where horizontal and vertical axes in correlation histograms represents time lag in ms and counts of coincident spikes in bins of 0.1 ms for \labelcref{fig:cchvswald:1,fig:cchvswald:2}.
+\label{fig:cchvswald}}
 \end{figure}
@@
 \begin{figure}
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{false_positive}
+\includegraphics[scale=1]{mse_pos}
+\caption{Mean square error (MSE) of Wald estimator in a recurrent neural network of three different parameter settings.
+The MSE as a function of number of trials shown on a logarithmic scale for asynchronous network activity when varying the relative inhibition yielding higher firing rate.
+\label{fig:error}}
+\end{figure}
+
+\begin{figure}
+\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{false_positive}
 \caption{} \label{fig:network-class:1}
 \end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{false_negative}
+\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{false_negative}
 \caption{} \label{fig:network-class:2}
@@
 \medskip
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{fit_wald}
+\begin{subfigure}{0.46\textwidth} \includegraphics[scale=1]{fit_wald_pos}
 \caption{} \label{fig:network-class:3}
 \end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{fit_cch}
+\begin{subfigure}{0.6\textwidth} \includegraphics[scale=1]{fit_cch_pos}
 \caption{} \label{fig:network-class:4}
 \end{subfigure}
-\caption{Comparing CCH and logit with IV \labelcref{fig:network-class:1,fig:network-class:2}. \label{fig:network-class}}
+\caption{False estimates and goodness of fit.
+Panel \labelcref{fig:network-class:1} show false positives for the cross correlation histogram (cch) method, logistic regression (logit) and the Wald IV estimator (wald).
+False negatives for cch and wald is shown in \labelcref{fig:network-class:2}.
+Positive estimates of weight as a function of true weight is scattered for the wald estimator in \labelcref{fig:network-class:3} and cch in \labelcref{fig:network-class:4} color coded by the size of perturbation intensity.
+\label{fig:network-class}}
 \end{figure}
@@
 
-\FloatBarrier
+%Wald is pairwise, GLMs require multiple neurons
+Pairwise Wald vs GLM multiple neurons
+
+%Inhibitory connections?
+Inhibitory connections may show up as negative Wald, but needs proper thinking
+
 \section{Methods}
@@
 Here $ \lambda $ represents the counts at bin $ m $ and $ N $ is the number of bins considered.
-\subsection{Simulated network}
-The first simple network consists of three leaky integrate and fire (LIF) neurons given in \cref{eq:LIF}:
+To estimate the connection weight between pairs we used the spike transmission probability defined in \cite{English2017} as
+\begin{align}
+P_{trans} = \frac{1}{n}\sum_{m=4ms}^{8} CCH(m) - \lambda_{Gauss}(m),
+\end{align}
+where $ n $ is the number of spikes detected in the presynaptic neuron.
+\subsection{Simulated network} \todo{refer to parameter table}
+To simulate a recurrent network of $ 1000 $ excitatory and $ 250 $ inhibitory neurons we used the leaky integrate and fire (LIF) model given by
 \begin{equation}
@@
 \end{equation}
-When the membrane potential $ V_{i} $ of neuron $ i $ reaches a threshold $ \theta $ an action potential is emitted and the depolarization is reset to the leak potential $ E_{L} $ followed by an absolute refractory period $ \tau_{rp} $.
+When the membrane potential $ V_{i} $ of neuron $ i $ reaches a threshold $ \theta $ an action potential is emitted and reset to the leak potential $ E_{L} $ followed by an absolute refractory period $ \tau_{rp} $.
 The membrane time constant is represented by $ \tau $ and $ I_{i}(t) $ denotes the synaptic currents arriving at the soma modeled as a sum of alpha functions given by
@@
 	\label{eq:syn}
-	RI_{i}(t) = \sum_{j}J_{ij}\sum_{k}t\exp(-(t-t_{j}^{k})/\tau_{syn}) - D.
+	RI_{i}(t) = \sum_{j}J_{ij}\sum_{k}t\exp(-(t-t_{j}^{k})/\tau_{syn}).
 \end{equation}\todo{make sure delay is correctly inserted}
@@
 When neuron $ j $ emits it's $ k $'th spike at time $ t_{j}^k $ it arrives at neuron $ i $ at $ t = t_{j}^k + D $ where $ D $ denotes a transmission delay.
-Furthermore all neurons are driven by an external Poisson drive and an Gaussian noise process given in \cref{eq:channelgauss}.
-\begin{equation}
-	I(t) = mean + std * N_j for t_0 + j dt \leq t < t_0 + (j-1) dt 
-    \label{eq:channelgauss}
-\end{equation}
-where $N_j$ are Gaussian random numbers with unit standard deviation and $t_0$ is 
-the device onset time.
+Furthermore all neurons are driven by an external Poisson drive.
+
+Synaptic efficacies were log normally distributed with a minimum and maximum amplitude at $ 0.05 mV $ and $ 2.05 mV $ respectively within bounds of experimental findings \cite{Sayer1990,Mason1991} as shown in \cref{fig:syndist}.
+\missingfigure{Figure showing synaptic distribution \label{fig:syndist}}
+
+To find suitable parameters yielding asynchronous activity as measured by the population correlation coefficient shown in \cref{fig:cchvswald} we performed several parameter sweeps and picked three parameter sets accordingly
+\missingfigure{parameter sweep \label{fig:sweep}}
+\missingfigure{CC distribution \label{fig:CC}}
+\missingfigure{CV distribution \label{fig:CV}}
+%and an Gaussian noise process given in \cref{eq:channelgauss}.
+%\begin{equation}
+%	I(t) = mean + std * N_j for t_0 + j dt \leq t < t_0 + (j-1) dt 
+%    \label{eq:channelgauss}
+%\end{equation}
+%where $N_j$ are Gaussian random numbers with unit standard deviation and $t_0$ is 
+%the device onset time.
 \todo[inline]{small perturbations on noisy networks will not yield a correlation between AC because the variance in stimulus response time in A will not be explained by the variance in spike times of C.
@@
 The only way to get a strong correlation between AC is to impose a short and strong stimulus effectively making AB spike coincidentally.}
+\subsection{Perturbation intensity}
+In order to replicate an optogenetic experiment we modeled transmission of light through brain tissue with the Kubelka-Munk model for diffuse scattering in planar, homogeneous, ideal diffusing media given by
+\begin{align}
+T = \frac{1}{Sz + 1}.
+\end{align}
+Here $ T $ denotes a transmisison fraction, $ S = 10.3 mm^{-1} $ is the scattering coefficient for mice \cite{Aravanis2007} and $ z $ is the distance from a light source \cite{Ho2017}.
+Further we combined diffusion with geometric loss assuming that absorption is negligible as in \cite{Aravanis2007} and computed the intensity as presented in \cref{fig:concept} by
+\begin{align}
+\frac{I(z)}{I(z=0)} = \frac{\rho^2}{(Sz + 1)(z + \rho)^2}
+\end{align}
+where
+\begin{align}
+\rho = r\sqrt{\left(\frac{n}{NA}\right)^2 - 1},
+\end{align}
+and $ r = 100\mu m $ is the radius of the optical fiber, $ NA = 0.37 $ is the numerical aperture of the optical fiber and $ n = 1.36 $ is the refraction index for gray matter \cite{Ho2017}.
+
+To estimate the distribution of light intensity on affected neurons we assumed a neuron density of $ 10^5 Nmm^{-3} $ and foo und the volume of a cut cone that could contain $ 950 $ ``source" excitatory neurons which were found to yield the depth $ z = 0.175 mm $.
+The last $ 50 $ of the excitatory neurons was used as the ``target" population which together with the inhibitory neurons were not perturbed directly by the light stimulus.
+
+In order to keep the stimulus model as simple as possible we used a perturbation strength of $ 10 pA $ which was found suitable by investigating the percentage of successful stimulations to be around $ 50\% $.
 

### 2018-03-25 analyse.py

@@ -9,7 +9,9 @@ from exana.statistics.tools import ccg_significance
 from tqdm import tqdm
 
 data_path = 'results'
-N = 10000
+# N_trials = 30000
+trials = [5000, 10000, 15000, 20000, 25000, 30000]
+N = 100
 winsize = 4
 binsize_corr = 5.
 min_stim = 1.
@@ -27,92 +29,95 @@ conn = dataa['connections']
 spiketrains = {s['sender']: {'pop': pop, 'times': s['times']}
                for pop in ['in', 'ex']
                for s in dataa['spiketrains'][pop]}
-sources = [s for s in dataa['stim_nodes']['ex'] if s in spiketrains.keys()]
-# targets = dataa['nodes']['ex'] + dataa['nodes']['in']
-targets = dataa['nodes']['ex']
-targets = [s for s in targets if s in spiketrains.keys()]
-pbar = tqdm(total=len(sources) * (len(targets) - len(sources)))
-for source in sources:
-    for target in targets: #NOTE different latency in inhibitory neurons
-        if target in sources:
-            continue
-        pbar.update(1)
-        stim_times = dataa['epoch']['times'][:N + 1]
-        period = np.min(np.diff(stim_times))
-        t_stop = stim_times[-1] + period
-        source_t = spiketrains[source]['times']
-        target_t = spiketrains[target]['times']
-        spike_trains = [source_t[source_t <= t_stop],
-                        target_t[target_t <= t_stop]]
-        iv = IV(*spike_trains, stim_times,
-                 winsize=winsize, latency=latency)
-        stim_amp = dataa['stim_amps'][spiketrains[source]['pop']]
-        try:
-            lr = iv.logreg
-            logreg = float(lr.coef_)
-            logreg_intercept = float(lr.intercept_)
-        except ValueError:
-            logreg = np.nan
-            logreg_intercept = np.nan
-        try:
-            lr = iv.logreg_ns
-            logreg_ns = float(lr.coef_)
-            logreg_ns_intercept = float(lr.intercept_)
-        except ValueError:
-            logreg_ns = np.nan
-            logreg_ns_intercept = np.nan
-        # cc, cv and stuff
-        w = conn[(conn.source==source) & (conn.target==target)].weight
-        n_syn = len(w)
-        weight = w.sum()
-        t_stop = dataa['params']['status']['time']
-        cc = corrcoef(spike_trains, t_stop,
[…]

### 2018-03-25 params_1.py

@@ -33,8 +33,8 @@ parameters = {
     'J_low'          : 0.05,
     'p_var'          : .5, # percentage variation of mean in lognormal dist
     # Stimulation parameters
-    'stim_nodes_ex'   : tuple(np.arange(1,602,1)),
-    'stim_nodes_in'   : (),
+    'stim_N_ex'       : 950,
+    'stim_N_in'       : 0,
     'stim_dist'       : 'poisson',
     'stim_amp_ex'     : 10.0, # pA
     'stim_amp_in'     : 0.0, # pA
@@ -42,7 +42,10 @@ parameters = {
     'stim_max_period' : 150, # only applies to poisson
     'stim_duration'   : 2.0, # ms
     'stim_N'          : 30000,
-    # analysis params
-    'latency': 6,
-    'winsize': 4
+    # Optogenetics
+    'r': 100e-3, # 100 um radius of fiber
+    'n': 1.36, # refraction index of gray matter
+    'NA': 0.37, # Numerical Aperture of fiber
+    'S': 10.3, # mm^-1 scattering index for rat, mouse = 11.2
+    'density': 10e4, # N/mm³
 }

### 2018-03-25 simulator.py

@@ -170,7 +170,7 @@ class Simulator:
                 {"weight": {'distribution': 'lognormal_clipped',
                             'mu': mu(j),
                             'sigma': sigma(j),
-                            'low': 0., 'high': high},
+                            'low': low, 'high': high},
                 "delay": self.p['delay']})
         self.vprint('Connecting excitatory neurons J = ', self.p['J_ex'], 'C = ',
               self.p['C_ex'])
@@ -261,19 +261,43 @@ class Simulator:
 
     def simulate_trials(self, progress_bar=False):
         progress_bar = progress_bar and PBAR
+
+        def intensity(z):
+            rho = self.p['r'] * np.sqrt((self.p['n'] / self.p['NA'])**2 - 1)
+            return rho**2 / ((self.p['S'] * z + 1) * (z + rho)**2)
+
+        def affected_neurons(z):
+            theta = np.arcsin(self.p['NA'] / self.p['n'])
+            lcorr = self.p['r'] / np.tan(theta)
+            rad = (z + lcorr) * np.tan(theta)
+            A = np.pi * rad**2
+            dz = z[1] - z[0]
+            dV = A * dz
+            N = dV * self.p['density']
+            return np.array([np.sum(N[:i+1]) for i in range(len(N))])
+
+        z = np.arange(0,1.01,.01)
+        neurons = affected_neurons(z)
+        mask = neurons <= self.p['stim_N_ex'] + self.p['stim_N_in']
+        z_N = z[mask][-1]
         # Simulate one period without stimulation
         nest.Simulate(self.stim_times[0])
 
         # Set dc stimulation
         stims = []
         amps = [self.p['stim_amp_ex'], self.p['stim_amp_in']]
+        if 'stim_nodes_ex' not in self.p:
+            self.p['stim_nodes_ex'] = self.nodes_ex[:self.p['stim_N_ex']]
+        if 'stim_nodes_in' not in self.p:
+            self.p['stim_nodes_in'] = self.nodes_ex[:self.p['stim_N_in']]
+
         assert all(np.in1d(self.p['stim_nodes_ex'], self.nodes_ex))
         assert all(np.in1d(self.p['stim_nodes_in'], self.nodes_in))
         stim_nodes = [self.p['stim_nodes_ex'], self.p['stim_nodes_in']]
         self.stim_amps = {'ex': [], 'in': []}
         for a, nodes, pop in zip(amps, stim_nodes, ['ex', 'in']):
             for n in nodes:
[…]

### 2018-03-25 tools.py

@@ -31,8 +31,11 @@ def fix_figure(padLegend=None, padHeight=0):
                      widthAx / widthTot, heightAx / heightTot + padHeight])
     plt.draw()
 
-def savefig(fig, fname, **kwargs):
-    fig.savefig(op.join('manuscript',fname + '.pdf'), **kwargs)
+def savefig(fig, fname, ext='.svg', **kwargs):
+    if ext[0] != '.':
+        ext = '.' + ext
+    fname = op.splitext(fname)[0]
+    fig.savefig(op.join('manuscript',fname + ext), **kwargs)
 
 def coef_var(spike_trains):
     """
@@ -114,11 +117,13 @@ def csv_append_dict(fname, dictionary):
         df.to_csv(fname, index=False, mode='a', header=False)
 
 
-def despine(ax=None, left=False, right=False, top=False, bottom=False,
-            xticks=True, yticks=True):
+def despine(ax=None, left=False, right=True, top=True, bottom=False,
+            xticks=True, yticks=True, all_sides=False):
     """
     Removes axis lines
     """
+    if all_sides:
+        left, right, top, bottom = [True] * 4
     if ax is None:
         ax = plt.gca()
     if not isinstance(ax, list):
@@ -169,7 +174,6 @@ def set_style(style='article', sns_style='white', w=1, h=1):
         }
     }
     rc = sdict[style]
-    sns.set_color_codes()
-    sns.set(rc=rc)
     plt.rcParams.update(rc)
-    sns.set_style(sns_style)
+    sns.set(rc=rc, style=sns_style,
+            color_codes=True)

### 2018-06-25 manuscript/main.tex
updates

--- 2018-03-19 manuscript/main.tex
@@
 \usepackage{amsthm}
+\usepackage{breakcites}
 \usepackage{graphicx,subcaption}
@@
 \usepackage{placeins} %FloatBarrier
+\usepackage{booktabs}
+\usepackage{natbib}
 \usetikzlibrary{arrows,shapes,calc}
-\tikzset{line/.style = {draw, -latex'}}
+\tikzset{line/.style = {draw, -latex', ultra thick}}
 \newcommand{\abs}[1]{\left|{#1}\right|}
@@
 \newcommand{\intdef}[4]{\int\limits_{#1}^{#2} #3\, \mathrm{d}{#4}}
+\newcommand{\mean}[1]{\left<{#1}\right>}
 \crefname{appendix}{Appendix}{appendices}
@@
 \Crefname{table}{Table}{Tables}
-\crefname{algorithm}{Algorithm}{algorithms}
-\Crefname{algorithm}{Algorithm}{Algorithms}
 \crefname{equation}{Eq.}{eqs.}
@@
 %opening
-\title{Correctly inferring causal connectivity using optogenetics}
+\title{Inferring causal connectivity from pairwise recordings and optogenetics}
+%\title{The need to combine optogenetics with neural recordings to estimate causal connectivity}
 \author{Mikkel Elle Lepper\o{}d, Konrad Paul Kording}
@@
 \begin{abstract}
-\noindent Neurons interact through spikes and a central objective of neuroscience is measuring the way how neurons causally affect one another.
-To probe such interactions, scientists often use optogenetics which typically leads to stimulation effects on multiple cells.
-This then produces a so-called confounding problem - we cannot know which of the stimulated neurons affected the activity of a given target neuron.
-Here we show how the resulting biases are large and how causal inference techniques, in particular instrumental variables, from econometrics can ameliorate this problem.
-Our approach utilizes the fact that neurons have absolute refractory times where the stimulation has no influence, and the missing stimulation acts like an effectively random variable which can then be used to disentangle causality.
-If we can not apply truly random stimulation to a network of neurons then causal inference techniques are needed to identify mechanisms.
+\noindent Neuroscientists often want to identify how neurons causally interact.
+Perturbations with optogenetics are a popular way of quantifying such interactions.
+However, even focused light will typically stimulate multiple cells, which produces a confound - we cannot know which of the stimulated neurons affected the activity of a given target neuron.
+Here we show how this confound produces large biases in estimates of connectivity, and how we can reduce these biases by a data analysis strategy that exploits the absolute refractory period of neurons.
+We use the instrumental variable (IV) framework from economics which estimates causal strengths by exploiting a variable that affects only one of the variables, here the presynaptic neuron.
+The interaction between optogenetic stimulation and the absolute refractory period produces a weak almost random signal (IV) that we exploit to solve the causal inference problem.
+In a simple simulation of integrate and fire neurons, we find that instrumental variable techniques estimate connectivity better than naive techniques (99.8\% vs 86\% correct).
+Our instrumental variable approach is robust and easy to implement in the context of current experimental paradigms.
 \end{abstract}
@@
 %a quest for causality and optogenetics is a main tool
+% We want mechanism. Seriously. Neuroscience is to a large part experimentally driven.
+%set up the drive to causality
+We want to understand the mechanisms or causal chains that give rise to activity in the brain, to perception, action, and cognition.
+For such an understanding it is not sufficient to know the correlations between variables or even be able to predict them.
+After all, there can be many ways how the same activities come about by distinct causal chains \citep{drton2011global, peters2017elements}.
+Identifiability of causal networks is complicated when it comes to brain data.
+Complex systems such as the brain are hard to understand because of the numerous ways the contributing elements may interact internally \citep{Jonas2017}.
+Thus, brain data which is almost always observational, will almost never satisfy the criteria needed for identifiability \citep{pearl2009causality}.
+While observing correlations within the system is usually relatively easy, transitioning from observed correlations to a causal or mechanistic understanding is hard or, more commonly, impossible.
+Getting at such an understanding in the human brain is nearly impossible as it contains approximately 86 billion neurons \citep{azevedo2009equal}, each of which influences many other neurons.
+Only under certain assumptions about nonlinearity or noise sources does a fully observed system become identifiable \citep{daniusis2012inferring,shimizu2006linear}.
+Even if we could record all neurons at the same time, estimating causality and producing a mechanistic understanding would be extremely challenging.
 
-%set up the drive to causality
-We want mechanism. Seriously. Neuroscience is to a large part experimentally driven.
-
-%we generally use observational studies
-We use observational studies. But causality is typically wrong. <nutshell version of the mehler paper>
+% we use observational studies. But causality is typically wrong. 
+Moreover, we generally only record from a small subset of all neurons.
+The data we obtain from typical recordings, e.g.
+from electrophysiology  or calcium imaging, is very low dimensional relative to the dimensionality of the brain.
+Moreover, it is observational, which means that it does not result from randomized perturbations.
+In such cases, we can never know to which level the observed activity was caused by other observed activity, or by the activity of the unobserved neurons.
+Such unobserved activity is then called confounders.
+If mechanisms are estimated from observational data in the presence of confounders, we will generally make large errors and draw incorrect conclusions \citep{angrist2008mostly}.
+Unobserved neural activity confounds estimates of causal interactions and makes it difficult to estimate underlying mechanisms.
 
 %set up confounding
- The experiments are extremely hard to control, even with the most accurate measurements of environmental variables and minimal genetic variance.
-Actually, given the vast amount of neurons and their connections in primitive mammals experiments are to be considered an observational sort.
-In a quest for experimental control we are now able to measure from thousands of neurons at the same time with micro-second precision by electrophysiology or with high spatial accuracy by two photon imaging.
-However, the large amount of unmeasured activity makes estimates of functional connectivity confounded.
-By confounded we mean that correlations between neuron pairs are largely influenced by the background activity.
-One example of this confounding is when two neurons $ A,B $ are driven by a common input $ D $, while one wants to estimate connectivity between $ A $ and $ C $.
-However only $ B $ and C is connected such that B and C are strongly correlated, then since $ B $ and $ C $ is correlated then $ A $ and $ C $ is correlated and a regression $ C = \beta A $ will give an erroneous $ \beta > 0 $ due to the the $ A,B $ correlation given by their common input $ D $.
+Confounding is the big threat to causal validity \citep{UpcomingMehlerPaper} irrespective of the use of simple regression techniques or advanced functional connectivity techniques \citep{stevenson2008inferring, honey2009predicting, aitchison2017or, pfau2013robust}.
+A much used method for estimating the output of single neurons is to perform multiple regression analyses \citep{pillow2008spatio}, modeling each neuron with a generalized linear model (GLM).
+Multiple regression may be perceived as a solution to confounding problems as they support the concept of ``explaining away" \citep{stevenson2008inferring}.
+However, in current work we will exclude approaches that look at multiple neurons at the same time.
+This is because, ``explaining away" can only be a good strategy if most neurons are included in the recordings but this is rarely the case for most experimental settings, especially in the mammalian brain.
+We will therefore focus on neuron pairs where stimulus and the pre- and post-synaptic neurons are known.
+
+To estimate connectivity it is first and foremost important that it stems from cause and effect e.g.
+B's action potential is causing an increase in C's membrane potential, therefore we use the term causal connectivity.
+For a given example we want to estimate causal connectivity between two neurons, $ A $ and $ C $ (\cref{fig:intro}\labelcref{fig:intro:1}).
+Two neurons, $ A $ and $ B $, are driven by a common input $ D $ and because $ B $ and $ C $ are connected they are strongly correlated.
+Consequently, $ A $ and $ C $ are also correlated and a regression  $ C = \beta A $ will misleadingly conclude that there is a direct interaction if it was causally interpreted.
 In this case we say the regressor $ A $ is endogenous and the regression coefficient $ \beta $ estimates the magnitude of association rather than the magnitude of causation.
-To estimate causal relationships between neurons, stimulation is appropriate, and if able to stimulate single neurons the ability to estimate causal relationships by regression is within grasp.
-However, this is experimental challenging and yield very low cell count.
-Therefore an optogenetic approach is more than often preferable.
+Na\"ive regressions in partially observed systems will generally not reveal causality.
+
+To estimate causal relationships between neurons, stimulating the pre-synaptic neuron is the gold standard.
+In fact, a common definition of causality is in terms of what would happen if one would change the value of one variable in the system, independently of changing other variables -- an intervention \citep{pearl2009causality}.
+If we stimulate single neurons, the ability to estimate causal relationships by regression is within reach.
+However, this is experimentally challenging and yield low cell count because it either requires intracellular, juxtacellular or two-photon stimulation \citep{pinault1996novel, lerman2017two, nikolenko2007two, emiliani2015all}.
+Because gold-standard perturbations are challenging, it would be highly desirable if causality could be obtained from optogenetic stimulation  \citep{boyden2005millisecond, zemelman2002selective}.
 
 %optogenetics is totally nonlocal
-Within the bounds of reachable tissue for photons emitted by a light source the decay of light intensity decays proportional to $ 1/r^2 $.
-Furthermore if the distribution of neurons is homogeneous the probability of affecting a neuron increases in distance by $ r^2 $.
-This essentially makes the distribution of affected neurons flat.
-When performing simultaneously recording with optogenetic stimulation we again get a confounding problem when estimating the functional relationships between neuron pairs.
-In this case the confounding is when several neurons see the same stimulus and thus become highly correlated during stimulation yielding statistically indistinguishable functional relationships, this case is similar to the above example where $ D $ now is the optogenetic stimulation.
-Flat distribution of V.
+Interpreting the results of optogenetic stimulation in terms of causal interactions is difficult.
+The expression of opsins are often restricted to a population of neurons, but regular optogenetic stimulation will still affect many neurons simultaneously.
+In \textit{in vivo} settings it is generally impossible to direct photons to only one neuron.
+Hence, the stimulus will produce a distributed pattern of activity.
+For example, if we focus a stimulation beam on one neuron, there will be a cone of light in front of and behind the neuron, which comes from the rays of light from the lens focused on the cell and again coming out on the backside..
+This distributed pattern of stimulation produces activity which then percolates through the network of neurons.
+Thus any post-synaptic activity induced by stimulation could in principle come from any of the stimulated neurons introducing problematic confounders.
 
 %causal inference is a thing, if not done well you say stupid things
-The problem of endogenous regressors is common in microeconomics and in the social sciences\todo{cite}, where observed outcomes can be driven by a large number of unknown factors.
-It is thus largely recognized that special care has to be taken when attempting to infer causal relationships between a putative actor and an outcome.
+The inference of causality from observational data is addressed in the fields of statistics \citep{pearl2009causality}, machine learning \citep{peters2017elements} and econometrics \citep{angrist2008mostly}.
+Within these fields, the problem of endogenous regressors is commonly addressed.
+We may thus look towards these fields for insights into how we may resolve the confounding problem induced by optogenetic stimulation.
 
 %example
-Let's say that you want to estimate the return $ \beta $ from education $ x $ to yearly income $ y $ with the regression $ y = \beta x + u $, where $ u $ is the factors other than schooling that contributes to yearly income.
-One of the factors in $ u $ is a persons ability, however this may also affect schooling and thus the regressor is correlated with the error term and a regression estimate will not estimate the magnitude of causation from schooling.
-In this case one may use the proximity to a college or university as an instrumental variable (IV) $ z $\todo{cite card 1995, found in cameron}.
-There are two criterion that must be fulfilled when choosing an instrumental variable, the instrument must be (1) uncorrelated with the error term, and (2) correlated with the regressor $ x $.
-Then one may for instance estimate $ \beta $ by a two stage ordinary least squares (OLS) where the first stage is an OLS of the regressor and the instrument which is further used in the second stage with the dependent variable $ y $\todo{cite TSOLS}.
-summary sentence.
+A commonly used approach towards causal inference in economics are instrumental variables.
+Let's say that we want to estimate the return $ \beta $ from education $ x $ to yearly wages $ y $ with the regression $ y = \beta x + u $.
+Here $ u $ are the factors other than education that contribute to yearly wages.
+One of the factors in $ u $ is a person's cognitive ability.
+However, a person's cognitive ability may also affect education and thus the regressor $ x $  is correlated with the error term $ u $.
+This will imply that the regression estimate $ \beta $ will not estimate the magnitude of causation from education on wages, but rather, its association.
+In this case one may use the proximity to a college or university as an instrumental variable (IV) \citep{card1993using}.
+Proximity is a good instrument since it can only affect wages through schooling, since we expect living in proximity of colleges and universities give higher probability to attend but not to affect other contributing factors to wages such as cognitive ability.
+Then, in order to attribute the causal effect of education on wages one may calculate the ratio of covariances $ \beta = cov(\mathrm{proximity},\mathrm{wages})/cov(\mathrm{proximity},\mathrm{education})$.
+This ratio corrects for the confounding factor.
 
-%worry is this really correct? Yes, its provable. 
-worry is this really correct? Yes, its . IVs are (under the right assumptions) actually correct.
+%worry is this really correct? Yes, its provable. IVs are (under the right assumptions) actually correct.
+The IV technique has been used extensively in econometrics and is provable causal given three assumptions \citep{angrist2008mostly}.
+First, the instrument must be uncorrelated with the error term.
+Second, the instrument must be correlated with the regressor.
+Third, there must be no direct influence of the instrument on the outcome variable but only an influence through the regressor variable.
+The validity of these assumptions is central when using the IV.
+
+% Refractory period as IV
+For an instrument to be good, it needs to be unaffected by other variables.
+In the brain, almost everything is affected by the network state.
+However, certain variables can be more or less affected.
+For example, the overall activity of the network is through slow and strongly nonrandom dynamics.
+In contrast, the temporal pattern of when a neuron is in a refractory state may be random.
+First, if neurons are spiking according to conditional Poisson distributions, their exact timing will, conditioned on the network state, be random.
+While refractoriness may not be perfectly random, the exact times of spiking are notoriously difficult to predict \citep{stevenson2008inferring} suggesting that refractoriness is likely quite random.
 
 %here we
-The IV technique can also be employed if one seeks to estimate the causal connectivity between neuron pairs.
-In this study we begin by showing how a simple network of three leaky integrate and fire (LIF) neurons can be heavily confounded under the right circumstances when simulating optogenetic targeting as direct current pulses.
-We then show that using the refractory period as an instrumental variable we are able to produce good estimates of causal connectivity.
-Furthermore we show that the method works in a large (N=1250) network of LIF neurons.
-With this data at hand we compare the IV method with a logistic regression.
+Here we show that the instrumental variable (IV) technique can be employed if one seeks to estimate the causal connectivity between neuron pairs.
+We begin by showing how confounding factors are introduced by regular optogenetic stimulations.
+We then simulate this confounding effect in a simple network of three leaky integrate and fire (LIF) neurons.
+With this simple model we show that by using the refractory period as an IV we are able to distinguish between connected and unconnected neuron pairs.
+We compare these estimates with a naive although widely used cross-correlation histogram (CCH) method that fails to distinguish respective pairs.
+We then turn to a simulated network of randomly recurrent connections of excitatory and inhibitory LIF neurons with distributed synaptic weights.
+With this data at hand we first calculate the mean squared errors of the IV method and show that it is robust to different simulated network states.
+Finally, we compare the amount and size of false positive and false negative estimates and goodness of fit on synaptic weights with pairwise assessments using CCH and logistic regression.
 
 \section{Results}
-\begin{figure}
-\makeatletter
-\renewcommand\p@subfigure{}
-\makeatother
-\begin{subfigure}{\textwidth} \includegraphics[scale=1]{light-decay}
-\caption{} \label{fig:concept:1}
-\end{subfigure}
+\subsection{Optogenetics is not local}
+Optogenetics is generally seen as a perturbation method that by and large affects neurons in close proximity of the light source.
+However, it can be questioned if this is the correct way of conceptualizing the spatial effect of stimulation.
+The stimulation effects depend on multiple factors.
+Firstly, light intensity and opsin density are important as more light and ion channels will cause a stronger effect on each cell.
+Secondly, the number of potentially stimulated neurons is critical as more neurons will have a larger impact on the overall population activity.
+Lastly, physiological properties of the cells are important as light may e.g.
+only have a strong effect on spiking activity when the membrane potential of the cell is sufficiently close to the firing threshold.
+The induced effect of optogenetic stimulation as a function of distance should be the product of four parameters:  light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and opsin distribution across neurons.
 
-\medskip
-\begin{subfigure}{\textwidth} \includegraphics[scale=1]{opto-powerlaw}
-\caption{} \label{fig:concept:2}
-\end{subfigure}
-\caption{Sketch of the problem at hand \label{fig:concept}}
-\end{figure}
-\todo[inline]{perturbation should spread by 1/r2 and all perturbed connections should be tested, here logit will fail.}
-\todo[inline]{figure showing stimulating of network}
-\todo[inline]{figure comparing methods with real weights}
-\todo[inline]{figure showing power (true false vs n stimulations)}
-\todo[inline]{data showing false positive/negative comparing stimulation patterns, comparing stimulation durations}
-\todo[inline]{data showing synchrounous regular does not work (can be put together with regular stim. pattern)}
-\todo[inline]{Should we show that ccg or gml give lots of false positives? without considering stimulation?}
-\todo[inline]{I think I would choose a couple of examples that emulate what an experimentalist would do
-(a) periodic slow (bad power not enough data)
-(b) periodic fast (good power but confounded)
-(c) random fast (good everything)
-But I would then also redo all of them with a "weak network" model where therefore the assumptions are pretty well matched an also a "strong network" model where neurons have massive influence on one another (maybe oscillate or something) and hence the weakness assumption that is behind the IV justification breaks down }
-Consider three neurons (A,B,C) where A and B receives optogenetic stimulation and only B is connected to C \cref{fig:intro}\labelcref{fig:intro:1}.
-The stimulation induces a strong correlation between A and B \cref{fig:intro}\labelcref{fig:intro:1,fig:intro:2,fig:intro:3,fig:intro:4}, and thus the monosynaptic connection between B and C is thus spuriously inherited between A and C confounding the system by rendering the cross correlation histograms (CCHs) between BC and AC statistically indistinguishable as seen in \cref{fig:intro}\labelcref{fig:intro:5,fig:intro:6} respectively.
-\begin{figure} 
-\makeatletter
-\renewcommand\p@subfigure{}
-\makeatother
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1.3]{simple}
-\caption{} \label{fig:intro:1}
-\end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{traces}
-\caption{} \label{fig:intro:2}
-\end{subfigure}
+To estimate the light intensity we calculated the spatial extent of laser light delivered by fiber-optics under plausible experimental conditions according to \cite{Aravanis2007}; see \cref{sec:method:opto}.
+This modeling of light intensity yield an approximately $ 1/r^2 $ reduction with distance $ r $ from the stimulation site \cref{fig:concept} (cyan line).
+This is explained by the surface of the 3D shell growing with $4\pi r^2$ and photons will be roughly isotropic beyond the scattering length \cref{fig:concept} (inset).
+The same number of photons has to cross each of the spheres around the stimulation location unless they are absorbed.
+The density of photons thus decreases rapidly with distance.
 
-\medskip
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{raster}
-\caption{} \label{fig:intro:3}
-\end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{xcorr}
-\caption{} \label{fig:intro:4}
-\end{subfigure}
-\caption{Sketch of a simple network containing three neurons depicted in red stimulated with blue laser light trough optogenetic activation.
-Invoked spikes due to stimulation is depicted in (b) upper panel with corresponding voltage traces in the lower panel.
-The neurons A and B are stimulated multiple trials and the  histogram of all trials are shown in (c) upper panel with the corresponding raster plot for each trial in the lower panel.
-Horizontal and vertical axes in correlograms represents respectively time lag in ms and counts of coincident spikes in bins of 1 ms for \labelcref{fig:intro:4} \label{fig:intro}}
+The number of illuminated neurons at a given distance will be larger the further away they are from the stimulation site given that neurons are uniformly distributed in brain tissue \cref{fig:concept} (black line).
+In fact, it will increase by approximately $ r^2 $ with distance.
+This derives from the same surface scaling as for the 3D shell.
+Thus the number of neurons that can be activated increases rapidly with distance.
+
+To estimate the effect of stimulation, the functionality underlying spiking activity needs to be considered.
+This can largely be characterized by the  distribution of membrane potentials across neurons.
+Surprisingly, this distribution has been observed to be symmetrically distributed and relatively flat \citep{pare1998impact,destexhe1999impact,rudolph2006use}.
+The response that is expected due to a pulse of light that induces a charge $Q$ should be proportional to the density of neurons whose membrane potential sit within a $Q/C$ range of the threshold ($C$ is the capacitance).
+The fact that the distribution of membrane potentials is relatively flat (the density close to the threshold is generally within an order of magnitude of the density of its mode) suggests that the response in membrane potential to a perturbation for any neuron is roughly proportional to the light intensity.
+Moreover, assuming that opsins are evenly distributed along the neuron, the opsin density in the neuron population will be relatively even.
+Based on this, we calculate the overall stimulation effect to be the product of neuron density and light intensity which is approximately constant in distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line).
+Thus, optogenetic stimulation utilizing single photon activation does not actually produce a localized effect.
+
+\begin{figure}\includegraphics[scale=1]{opto-powerlaw}
+\caption{{\bf Spatial extent of optogenetic stimulus}.
+Due to scattering and geometric loss the light intensity (I, cyan line) plotted as percentage of intensity exiting the optogenetic fiber follows approximately an inverse square law $ r^{-2} $ where $ r $ is the distance from the fiber.
+If neurons are uniformly distributed the number of affected neurons increase by $ r^{2} $ (N, black line) rendering the probability of activating a neuron approximately constant (IN, blue line).
+\label{fig:concept}}
 \end{figure}
 
-\begin{figure}
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{xcorr_highres_BC}
-\caption{} \label{fig:cchvswald:1}
-\end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{xcorr_highres_AC}
-\caption{} \label{fig:cchvswald:2}
-\end{subfigure}
+\subsection{Confounding as a problem for the estimation of causal effects}
+When we stimulate many neurons at the same time, and observe a post-synaptic neuron to be active after our stimulation, it is hard to know which of the stimulated neurons produced the activity.
+To illustrate such confounding effects we simulated a network comprised of three neurons (A,B, and C).
+The neurons receive Poisson spike trains and have added Gaussian white noise to the membrane potential.
+We also gave them interactions, where spikes of neuron B increase the probability of firing for neuron C but there are no other interactions \cref{fig:intro}\labelcref{fig:intro:1}.
+Finally, we allowed simulated optogenetic stimulation to affect neurons A and B (but not C). We thus have a simple system for exploring questions of causality.
 
-\medskip
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{wald_BC}
-\caption{} \label{fig:cchvswald:3}
-\end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{wald_AC}
-\caption{} \label{fig:cchvswald:4}
-\end{subfigure}
-\caption{Cross correlograms for all neurons are depicted in (d) and the high resolution zoom of BC and AC are visualized in (e) and (f) respectively.
-Horizontal and vertical axes in correlograms represents respectively time lag in ms and counts of coincident spikes in bins of 0.1 ms for \labelcref{fig:cchvswald:1,fig:cchvswald:2}.
-\label{fig:cchvswald}}
-\end{figure}
+After running the simulation, the peri-stimulus time histogram of the stimulated neurons show the result of both the stimulation itself (suppressed for visibility) and the neuron's refractory period \cref{fig:intro}\labelcref{fig:intro:2} (AA, BB).
+Since the stimulation affects A and B simultaneously, it induces a strong correlation between A and B \cref{fig:intro}\labelcref{fig:intro:3} (AB).
+This further generates a strong correlation between A and C, confounding the system by rendering the cross correlation histograms (CCHs) between BC and AC both statistically significant ($ p_{\mathrm{fast}} < 0.001, p_{\mathrm{diff}} < 0.001 $; see \cref{sec:method:cch}) shown in \cref{fig:intro}\labelcref{fig:intro:3}.
+Even though the correlation peak between B and C is larger than between A and C due to correlated spikes outside the periods with stimulation one may imagine a situation where only A and C is measured, giving rise to a false prediction that they are connected.
+If stimulation affects multiple neurons simultaneously, there is a real confounding problem.
 
@@
 \makeatother
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.2]{network_stim}
-\caption{} \label{fig:network-power:1}
+\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{simple}
+\caption{} \label{fig:intro:1}
 \end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth}\includegraphics[scale=.2]{wald_network}
-\caption{} \label{fig:network-power:2}
-\end{subfigure}
-
-\caption{Power analysis of IV estimator. \label{fig:network-power}}
-\end{figure}
-
-\begin{figure}
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{false_positive}
-\caption{} \label{fig:network-class:1}
-\end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{false_negative}
-\caption{} \label{fig:network-class:2}
+\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{raster}
+\caption{} \label{fig:intro:2}
 \end{subfigure}
@@
 \medskip
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{fit_wald}
+\begin{subfigure}{\textwidth}\centering \includegraphics[scale=1]{xcorr}
+\caption{} \label{fig:intro:3}
+\end{subfigure}
+\caption{{\bf A,B,C network}. A sketch of the simple network containing three neurons %depicted in \labelcref{fig:intro:3} 
+shows stimulation configuration with blue laser light and the connections with arrows (a).
+The neurons A and B are stimulated 1000 trials and the corresponding peristimulus time histogram are shown in \labelcref{fig:intro:2} upper panel with a raster plot in the lower panel.
+Cross correlation histograms (CCHs) are shown in \labelcref{fig:intro:3} where horizontal and vertical axes represents time lag in ms and counts of coincident spikes in bins of 1 ms.
+\label{fig:intro}}
+\end{figure}
+
+\subsection{Instrumental variables to resolve confounding}
+In order to estimate the actual influence of stimulation of a neuron on post-synaptic neurons we need something that can distinguish the influence of one stimulated neuron from the influence of another stimulated neuron.
+We would thus need something that affects the stimulation effect separately between neurons.
+Arguably, refractoriness is such a variable.
+If a neuron is in its absolute refractory period, then no amount of stimulation will make it spike.
+This gives us an interesting way of getting at causality, by comparing the network state between a time when a neuron is able to spike and a time where the neuron is unable to spike.
+
+Instrumental variables requires the existence of a random (or sufficiently random) variable that affects a variable of interest.
+This independent influence among neurons then allows quantifying the influence of the variable of interest on the rest of the network.
+In our case, the refractory states of a neuron is in good approximation independent on small time scales (see Discussion for caveats).
+It affects the influence of stimulation on the potentially pre-synaptic neuron (\cref{fig:cchvswald}\labelcref{fig:cchvswald:0}).
+The trials where a stimuli is unable to elicit a spike due to the refractory state can then be used to identify causal effect on the potentially post-synaptic neuron.
+
+We can now investigate if the use of an instrumental variable gives a better estimate of connectivity strength than simply analyzing the lagged correlations by means of the CCH; see \cref{eq:ptrans}.
+We thus use the IV estimator \cref{eq:wald} on the three neuron system (\cref{fig:cchvswald}\labelcref{fig:cchvswald:1,fig:cchvswald:2}).
+It converges to the correct causal conclusions that the weights $ w_{BC} = 0.2 $ and $ w_{AC} = 0 $.
+For such a simple system, it produces meaningful estimates of the causal interactions between neurons.
+
+\begin{figure}
+\makeatletter
+\renewcommand\p@subfigure{}
+\makeatother
+\begin{subfigure}{0.485\textwidth} \centering
+% * <tristan.stoeber@posteo.net> 2018-06-24T21:23:23.987Z:
+% What is S?
+% ^.
+\begin{tikzpicture}
+    \node (u) {$ S $};
+    \node [above = of u] (x) {$ A/B $};
+    \node [above right = of u] (y) {$ C $};
+    \node [left = of x] (sr) {$ A_r/B_r $};
+    \path [line] (x) -- (y);
+    \path [line] (u) -- (x);
+    \path [line] (u) -- (y);
+    \path [line] (sr) -- (x);
+\end{tikzpicture}
+\caption{} \label{fig:cchvswald:0}
+\end{subfigure}\hfill
+\begin{subfigure}{0.485\textwidth} \centering
+\includegraphics[scale=1]{wald_legend}
+\end{subfigure}\medskip
+
+\begin{subfigure}{0.485\textwidth} 
+\includegraphics[scale=1]{wald_BC}
+\caption{} \label{fig:cchvswald:1}
+\end{subfigure}\hfill
+\begin{subfigure}{0.485\textwidth} 
+\includegraphics[scale=1]{wald_AC}
+\caption{} \label{fig:cchvswald:2}
+\end{subfigure}
+\caption{{\bf Instrumental variable estimation (IV) of connectivity}.
+\labelcref{fig:cchvswald:0} In an instrumental variable estimation procedure we use a variable that is assumed to be random (here refractoriness) which influences a variable of interest (here spiking) and to use this influence to infer the causal interaction of that variable on other variables (here spiking of A or B onto C).
+\labelcref{fig:cchvswald:1} A popular estimation approach for IVs,  the Wald technique correctly estimates causal connectivity in the A,B,C neural network using the refractory period.
+Subfigure \labelcref{fig:cchvswald:0} shows a path diagram \citep{wright1921correlation} between the upstream neuron $ A $ or $ B $, the stimulation $ S $, the post-synaptic neuron $ C $ and the IV as $ A_{r} $ or $ B_{r} $.
+Arrows represent associations where $ S $ is associated with $ A,B $ and potentially also with $ C $ both directly and through $ A,B $.
+The IV estimator calculated by \cref{eq:wald} converges to $ \hat{\beta}_{BC} \approx 0.2, \hat{\beta}_{AC} \approx 0 $ after approximately $ 5000 $ trials as seen in \labelcref{fig:cchvswald:1,fig:cchvswald:2} respectively.
+% I would have put in a take home message / interpretation of the results in the figure text.
+Insets represent high resolution zoom of cross correlation histograms of BC and AC where horizontal and vertical axes represents time lag and counts of coincident spikes in bins of 0.1 ms respectively.
+\label{fig:cchvswald}}
+\end{figure}
+
+\subsection{Larger simulated networks}
+The interacting neurons in a biological network exhibit inhibition and interact in many kinds of ways.
+To evaluate the IV method in a more meaningful setting we thus simulated a recurrent neural network consisting of 1250 randomly connected leaky integrate and fire neurons where 250 had inhibitory synapses.
+The network was tuned to be in an asynchronous regime; see \cref{fig:state}\labelcref{fig:CC} with log-normally distributed synaptic weights according to patch clamp experiments \citep{Sayer1990,Mason1991}; see \cref{fig:state}\labelcref{fig:syn_dist} and \cref{tab:params} for parameters.
+Further, we selected 800 excitatory neurons for stimulation and gave each neuron a random spatial distance from the simulated optogenetic stimulus.
+The stimulus intensity was then set according to \cref{eq:intensity} with a maximum of 10 pA, and was constant throughout trials.
+The trial onset had a temporal Poisson distribution with period 100 ms and was further clipped between 100-150 ms.
+For weight estimates we randomly selected among the excitatory population 50 stimulated neurons and 50 unstimulated neurons.
+
+To evaluate how well the IV method estimates the weights as a function of number of trials we calculated the mean squared error of the weight connecting the 100 neuron pairs \cref{fig:error}.
+As seen here, the IV estimator's precision decreases similarly in three different settings with varying amounts of relative inhibition $ g $.
+\begin{figure}
+\makeatletter
+\renewcommand\p@subfigure{}
+\makeatother
+\begin{subfigure}{\textwidth} 
+\includegraphics[scale=1]{network-raster}
+\caption{} \label{fig:error:raster}
+\end{subfigure}\medskip
+
+\begin{subfigure}{\textwidth} 
+\includegraphics[scale=1]{mse_pos}
+\caption{} \label{fig:error:mse}
+\end{subfigure}
+\caption{{\bf Mean square error (MSE) of IV estimator in a large network}.
+\labelcref{fig:error:raster} Excitatory neurons (in blue raster) are stimulated with varying intensity indicated by the y-axis in lower panel, upper panel shows inhibitory neurons (red raster) where y-axis indicate random neuron-id.
+\labelcref{fig:error:mse} The IV- and CCH estimator is evaluated for the asynchronous recurrent neural network at three different amounts of relative inhibition $ g $.
+The MSE as a function of number of trials is shown on a logarithmic scale where the slopes was found to be $ -0.77, -0.48, -0.40 $ for $g = 9.9, 4.4, 3.0$ respectively.
+\label{fig:error}}
+\end{figure}
+
+We want to compare the IV estimator which exploits the refractory period with the CCH method given by \cref{eq:ptrans} which ignores network confounding.
+To indicate the amount of connections which are falsely attributed a non-zero weight we calculated the amount of false positives.
+This was given as the percentage of estimated synapses larger than $ 0.05 $ where the true weight was $ 0 $, finding $ 13.3\% $ for CCH and $ 0.2\% $ for the IV estimator; see \cref{fig:network-class}\labelcref{fig:network-class:1}.
+In addition we compared the size of the estimations at false positive instances and found that the CCH have slightly higher median.
+The IV approach, while not being perfect, thus outperforms the simple CCH approach.
+
+It might be that modeling refractory periods in the context of a naive regression estimates connectivity equally well as the IV method.
+We thus performed a logistic regression as seen in \cref{fig:network-class}\labelcref{fig:network-class:1} denoted LOGIT.
+As seen here, LOGIT performs even worse than CCH showing that it really helps to use the refractory period as an instrumental variable.
+To further evaluate the methods we calculated false negatives as instances where the true weight is non-zero but estimated to be equal to zero in \cref{fig:network-class}\labelcref{fig:network-class:2} showing that the CCH and IV estimators performs equally well.
+Finally we wanted to evaluate the estimated weights as a function of true weights shown in \cref{fig:network-class}\labelcref{fig:network-class:3,fig:network-class:4} after 30000 trials.
+The IV estimator yields a good prediction $ r^2 = 0.55 $, while the CCH estimator performs surprisingly bad with $ r^2 = 0.002 $.
+Utilizing refractory periods as an (imperfect) instrumental variable considerably improves estimates.
+
+\begin{figure}
+\makeatletter
+\renewcommand\p@subfigure{}
+\makeatother
+\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{false_positive}
+\caption{} \label{fig:network-class:1}
+\end{subfigure}\hfill
+\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{false_negative}
+\caption{} \label{fig:network-class:2}
+\end{subfigure}
+% * <tristan.stoeber@posteo.net> 2018-06-24T21:29:41.690Z:
+% 
+% Why is logit not considered for false negatives?
+% 
+% ^.
+
+\medskip
+\begin{subfigure}{0.46\textwidth} \includegraphics[scale=1]{fit_wald_pos}
 \caption{} \label{fig:network-class:3}
 \end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=.25]{fit_cch}
+\begin{subfigure}{0.6\textwidth} \includegraphics[scale=1]{fit_cch_pos}
 \caption{} \label{fig:network-class:4}
 \end{subfigure}
-\caption{Comparing CCH and logit with IV \labelcref{fig:network-class:1,fig:network-class:2}. \label{fig:network-class}}
+\caption{{\bf False estimates and goodness of fit}.
+False positives are shown in \labelcref{fig:network-class:1} for the cross correlation histogram (CCH) method, logistic regression (LOGIT) and the IV estimator.
+False negatives for CCH and IV is shown in \labelcref{fig:network-class:2}.
+Positive estimates of weight as a function of true weight is scattered for the IV estimator in \labelcref{fig:network-class:3} and CCH in \labelcref{fig:network-class:4} color coded by the size of perturbation intensity.
+Shaded area show the 95 \% confidence interval calculated by bootstrapping.
+\label{fig:network-class}}
 \end{figure}
@@
 \section{Discussion}
-Spike time methods works well when network is desynchronized, small amplitude stimulation with close range. Our method opens up for the possibility of strong perturbations giving larger yield.
-We are making the network results quite hard as all synapses have the same time constant such that multiple neurons are affecting the same cell at the same time for each stimulation
+%% maybe add somewhere that using other recorded neurons may be useful as well.
+
+Here we have asked if the refractory period of neurons can be used as an instrumental variable to reverse engineer the causal flow of activity in a network of simulated neurons.
+We have found that this approach performs considerably better than the naive method.
+We have found that neither naive linear nor naive logit models produce good estimates of weights.
+Our system effectively reverse engineers causality by looking at the response that is missing because of refractoriness which effectively allows better estimates of causal effects.
+
+%Wald is pairwise, GLMs require multiple neurons
+One popular way of estimating causal effects is fitting generalized linear models (GLMs) to simultaneously recorded neurons \citep{pillow2008spatio}.
+The GLMs are basically multiple regressions and require multiple neurons in order to perform well.
+In fact, if one recorded all neurons, GLMs could be sufficient to estimate causal connections.
+However, this is not the case for research in the mammalian brain, especially not for primates, where we only record a very small subset of the neurons that do the actual computation.
+The GLM field is very strong at modeling latency distributions and sequences of spikes in individual neurons.
+These ideas should, arguably, be merged with IV approaches.
+The main strength of the  IV estimator presented here compared with GLM methods is that we only require one pair because we can utilize the randomness in refractoriness.
+
+%We recommend the WALD, easily usable in the context of many experiment. Potentials for future improvement.
+%The IV estimator seems to be useful when using the refractory period.
+This method yields a direct estimation by computing the difference in mean activity of a putative post-synaptic neuron when the upstream neuron is responsive to the stimulus or not.
+Moreover, as this method only requires one pair it is easily usable in many types of experiments, with large or small recording counts.
+However, there would be many ways of improving the estimates, we could better model the temporal sequence of spikes, we could utilize latency in a more meaningful way.
+While there are many ways of improving the IV estimators, we have here represented the most basic implementation using the Wald estimator.
+
+% Maybe nonlinearities in stimulation rescue old school approaches
+% Discuss distribution of V. Discuss 2p stimulation.  Conclude that there is absolutely no reason to hope things will work with 1p.
+The main problem with optogenetics when using it to infer connectivity is its non-local property.
+This is due to the inverse relation between changes in light intensity and affected number of neurons combined with the observation that distribution of membrane potentials across neurons is flat \citep{destexhe1999impact,rudolph2006use,pare1998impact}.
+One could however imagine situations where optogenetic activation were more local.
+If membrane potential distributions were in general skewed with the mode far from threshold the optogenetic perturbation would be more local.
+This is because each neuron would require a very strong stimulus in order to  elicit spikes at all.
+But there could be other ways of making optogenetics more local.
+For example, if one could engineer opsins with far more absorbent wavelengths one could stimulate locally.
+How to engineer more localized stimulation is an important problem if one wants to causally interrogate systems.
+
+%Maybe weak stimulation will rescue us. No way, but cite Buszaki.
+Very weak laser pulses in noisy networks might only affect very few neurons each trial \citep{English2017}.
+However, the stimulus will still affect the many far away neurons by a tiny bit.
+Therefore, weak stimulation does not remove the principal problem of the CCH estimator.
+Moreover, the network still acts as a confounder and, if anything, the weak stimulation will dramatically reduce the statistical power of the approach.
+Lowering stimulation amplitudes does not appear to be a way of obtaining meaningful causal estimates.
 
 %Correlated refractory periods is *the* failure mode. How to randomize them
+For the refractory period to be a good instrument, it is necessary that it is not overly affected by the network activity.
+This clearly is going to be problematic in many cases, after all the network activity affects refractoriness.
+However, there are multiple scenarios where refractoriness will be a good instrument.
+For example, if we have balanced excitation and inhibition we may expect largely independent refractory states of individual neurons.
+If a neuron biophysically implements something like conditional Poisson spiking its refractory states will be random.
+Even if neurons refractory states are strongly correlated during normal network operation there may be ways of randomizing them.
+Giving one burst of stimulation which is strong enough to elicit multiple spikes from each neuron may effectively randomize the phase of each neuron \citep{ermentrout2008reliability}.
+Importantly, we may expect the phase of a neuron to be far more random than the activity of the network as a whole.
 
-% Maybe nonlinearities in stimulation rescue old school approaches
-Discuss distribution of V. Discuss 2p stimulation.  Conclude that there is absolutely no reason to hope things will work with 1p.
+%Still some troubles
+We found some negative values of the IV estimator which were suppressed as we knew that we only stimulated excitatory neurons.
+This happens because some neurons  have correlated refractory times which we hypothesize to be mainly due to synchrony induced by the stimulation \citep{ermentrout2008reliability}.
+This hypothesis was strengthened by observing that this negative values did not occur when the stimulation intensity was set to zero (data not shown).
+Furthermore, respective simulated neural network introduces much response overlap due to synapses having identical synaptic time constants and transfer delays.
+This makes inference quite hard since multiple neurons are affecting the same cell at the same time for each stimulation.
+However, this is less likely to occur in biological networks, which have high variability of synaptic properties and where firing patterns are sparser.
+These biological aspects would most likely work to the advantage of the IV method.
+Independence of refractoriness would be further improved, if in addition a clever stimulation routine was implemented such that the distribution of stimulation strength varies spatially from trial to trial.
 
-%Maybe some biophysics pecularity will rescue us
-No way. But cite Buszaki.
+% Importance of randomized stimulus times and short pulses 
+The independence of refractory times is the one factor which makes or breaks the IV approach.
+Therefore we may think of ways of making the refractoriness distribution more random.
+First, it would help to use a task and situation where neurons are as uncorrelated as possible.
+Second, we may use a set of conditioning pulses of stimulation to increase independence of refractory states.
+Third, we may utilize chemical, behavioral, or molecular perturbations to randomize refractoriness.
+There has not, to the authors knowledge, been any attempts in neuroscience to randomize times of refractory states, so there may be a lot of possibilities to improve.
 
-%We recommend the WALD 
-easily usable in the context of many experiment. Potentials for future improvement.
-
-%brain regions versus neurons
-IVs on multiple scales.
+Generalizing this idea, we may ask if there are ways of constructing good instrumental variables.
+One may assume a way of building molecular oscillators or otherwise pattern generators into neurons which affect their firing rate.
+Such modulatory activity would be observable in the extracellular activity of the neuron.
+Any neuron that then correlates to this modulation must be post-synaptic of the recorded neuron.
+Any kind of a signal that is local to a cell and not affected by network activity could produce a meaningful instrumental variable and it might be reasonably doable to link membrane channels to intracellular signal generation.
+Even if perfect IVs do not exist in brains, we may be able to make them.
 
 %Causal inference techniques to fix experimental issues
-Broad opportunity, cite RDD, cite diff-in-diff etc, broad overview of the promise of econometrics for neuro, matching
+There are many techniques for causal inference and most of them are largely unknown in neuroscience and are based on approximating randomness in a world that does not have it.
+In many cases, one could use regression discontinuity designs in a spiking system \citep{BenPaper, imbens2008regression}.
+Moreover, one could use a difference in difference approach \citep{abadie2005semiparametric}.
+Matching approaches \citep{stuart2010matching}, but see \citep{king2016propensity}, can be used where one compares similar network states and their evolution over time.
+In general, neuroscience is in a quest for causal interpretations, we should therefore be able to benefit considerably by utilizing techniques that are popular in the field of causal inference.
 
-%Constructing good IVs
-Multiple illuminations, designed sticky opsins
+%Constructing good IVs; Multiple illuminations, designed sticky opsins
+% Many trials needed here, good IV could possibly decrease this number
+%We note that we obtained low errors from the IV estimator after relatively few trials when estimating weights in the neural network of 1250 neurons.\todo{how would this be in the brain with more variable synaptic properties?} This amount of trials is not experimentally possible and we therefore foresee the need for improvements of the IV method.
+It can be improved in two ways, either with algorithms that uses more of the information available, such as the exact stimulus latencies and/or the amount of spikes in each trial.
+The other improvement could be to design experimental procedures that open for causal interpretation.
 
-\FloatBarrier
+%If we assume that the assumption of random refractory periods are violated somehow by the physiology, e.g.a very synchronized network.
+To ameliorate one could introduce multiple light sources that is activated in a randomized fashion such that the intensity distribution is randomized.
+This could be realized with e.g.
+laser holograms/filter or multiple micro LEDs This would in addition increase the amount of neurons targeted and thus give larger experimental yield.
+Another possibility is to design opsins that have a fluctuating baseline levels randomizing membrane potential of affected neurons, however this could also lead to a chronic deficit in respective neural network and would thus shorten the time span of the experimental procedure.
+But importantly, designing experiments so that causal inference is possible should be a major consideration for experimental design.
+
 \section{Methods}
 \subsection{Instrumental variable estimation}
-A simple approximation of the connectivity strength between putative sender and receiver neurons $ x $ and $ y $ respectively can be to ignore long distance excitation and simply calculate the relation between the spike times in $ x $ and $ y $ with a regression model given by
+A simple approximation of the connectivity strength between a pre-synaptic neuron $ x $ and post-synaptic neuron $ y $ can be to ignore external excitation and simply calculate the relation between the spike times in $ x $ and $ y $ with a regression model given by
 
@@
 \end{equation}
-Here $ y $ is the dependent variable, $ x $ is the explanatory variable, $ \beta $ is the slope of the $ x,y $ curve and $ u $ is an unknown error term. This system follows the causal path:
+Here $ y $ is the dependent variable, $ x $ is the explanatory variable, $ \beta $ is the effect of the $ x$ on $y $ and $ u $ is an unknown error term.
+\Cref{eq:regress} follows from the causal path diagram (invented by \cite{wright1921correlation})
 \begin{center}
@@
 One problem with this idea is that in a confounded system, perfectly correlated neurons will give statistically indistinguishable $ \beta $.
-In the extreme case where two neurons are both made to fire every time they are stimulated, they will have the same weights according to \cref{eq:regress}, after all, during stimulation  $ y=1 $ for both, even if only one of them drives the downstream neuron.
-Another problem is if the network state affects both the probability of a neuron to fire and also the probability of downstream neurons to fire.
+In the extreme case where two neurons are both made to fire every time they are stimulated, they will have the same weights according to \cref{eq:regress}.
+After all, during stimulation $ y=1 $ for both, even if only one of them drives the post-synaptic neuron.
+Another problem is if the network state affects both the probability of a neuron to fire and also the probability of post-synaptic neurons to fire.
 In this case, the network state can induce a correlation which will make the estimation highly biased.
-Arguably, the network state will, in all realistic models, have a dramatic influence on all neurons.
-In general, if multiple neurons are stimulated synchronously our estimates will be off, potentially massively so as seen in the equation
+Arguably, the network state will, in all realistic models, have a dramatic influence on all neurons and the regression model is better described by
 \begin{equation}
@@
 \end{equation}
-Following the causal path
+Corresponding to the following path diagram
 
@@
 \end{center}
-We now have the relation $ \de{y}{x} = \beta + \de{u}{x} $.
-To get at causality we thus require some stimulation that only highlight the activity in $ y $ caused by $ x $, however optogenetic stimulation is not specific enough as the photons will activate parts of the network activity $ u $.
-To have a remedy, we need something that can distinguish between different neurons that are stimulated.
-We thus require some instrument $ s_{r} $ which is (1) uncorrelated with the network $ u $ and (2) is correlated with the regressor $ x $\todo{cite the book or whatever}.
-The graph with stimulation is illustrated by
+Here we have the relation $ \de{y}{x} = \beta + \de{u}{x} $.
+To get at causality we thus require some stimulation that only highlight the activity in $ y $ caused by $ x $, disassociating $x$ from $u$.
+However, the optogenetic stimulation is not specific to $x$ and will activate parts of the network activity $ u $.
+Let us assume that the stimulus renders only a subset of $u$ correlated with $x$, namely $u_s$ ($s$ denotes stimulated).
+To disassociate $x$ from $u_s$ we need something that can distinguish between different neurons that are stimulated.
+We thus require some instrument $ x_{s_r} $ which is (1) uncorrelated with the network $ u $ and (2) is correlated with the regressor $ x $ \citep{angrist2008mostly}.
+We assume that the neurons are independent at small time scales and that stimulation additionally randomize membrane potential individually in neurons.
+We may thus use the fact that a neuron that has fired just before the stimulation will be in an absolute refractory state and hence have $ x_{s_r}=0 $ independently of $u$, where the subscript $s_r$ denotes stimulation during refractory state.
+This introduces times where the spike from one of the stimulated neurons are missing.
+Thus we may use the refractory period as an instrumental variable, as illustrated with the following path diagram
 
@@
 \begin{tikzpicture}
-    \node (u) {$ u $};
+    \node (u) {$ u_s $};
     \node [above = of u] (x) {$ x $};
     \node [above right = of u] (y) {$ y $};
-    \node [left = of x] (s) {$ s $};
-    \path [line] (x) -- (y);
-    \path [line] (u) -- (x);
-    \path [line] (u) -- (y);
-    \path [line] (s) -- (x);
-    \path [line] (s) -- (u);
-\end{tikzpicture}
-\end{center}
-The previous graph represents a confounded stimulation, however we may use the fact that a neuron that has fired just before the stimulation will be in absolute refractory period and hence have $ x_{s}=0 $.
-This introduces times where the spike from one of the stimulated neurons is missing.
-Neurons have low firing probability once we use small bins and are very noisy at spiking at small time scales.
-Thus if we assume that (1) the stimulation pattern is random (2) the network activity $ u $ is asynchronous and irregular we may use the refractory period as an instrumental variable illustrated in the following graph
-
-\begin{center}
-\begin{tikzpicture}
-    \node (u) {$ u $};
-    \node [above = of u] (x) {$ x $};
-    \node [above right = of u] (y) {$ y $};
-    \node [left = of x] (sr) {$ s_r $};
+    \node [left = of x] (sr) {$ x_{s_r} $};
     \path [line] (x) -- (y);
@@
 \end{center}
-\todo[inline]{this will not work if you stimulate non random since then you may correlate u with sr, u is still driving x but asynchronously so sr is still uncorrelated with u.}
-Here $ s_r $ represents times where the sender neuron is refractory during stimulation.
-This is then an estimator that compares the downstream activity when a given neuron is non-refractory with the downstream activity when it is, thus removing the confounding.
+Here $ x_r $ represent times where the pre-synaptic neuron is refractory during stimulation.
+This is then an estimator that compares the post-synaptic activity when a given neuron is non-refractory with the post-synaptic activity when it is refractory, thus removing the confounding.
 The true $ \beta $ is given by
+% * <tristan.stoeber@posteo.net> 2018-06-24T22:11:38.685Z:
+% Shouldnt it be  $ x_s_r $
+% ^.
 \begin{equation}
-\beta_{IV} = \de{y}{s_r} / \de{x}{s_r}
+\beta_{IV} = \de{y}{x_{s_r}} / \de{x}{x_{s_r}}
 \end{equation}
-Since our instrument $ s_r $ is binary we may use the Wald estimator \todo{cite wald} to estimate $ \beta_{IV} $ by
+Since our instrument $ x_{s_r} $ is binary we may use the IV (or more precisely Wald) estimator \citep{wald1940fitting} to estimate $ \beta_{IV} $ by
 \begin{equation}
@@
 \end{equation}
-Here $ \bar{y}_s $ is the average number of trials where stimulating $ x $ resulted in a response in $ y $ and $ \bar{y}_{s_r} $ is the average number of trials where an unsuccessful stimulation resulted in a response in $ y $.
+Here $ \bar{y}_s $ is the average number of trials where successfully stimulating $ x $ resulted in a response in $ y $ and $ \bar{y}_{s_r} $ is the average number of trials where an unsuccessful stimulation of $x$ resulted in a response in $ y $.
+The successful stimulations of $x$ are denoted $x_s$ and thus  $\bar{x}_s \equiv 1$.
+Conversely $x_{s_r}$ denotes unsuccessful stimulations of $x$ i.e.
+stimulations of $x$ during its refractory state and $\bar{x}_{s_r} \equiv 0$.
 
-However the estimate of the refractory period $ s_r $ may be biased, as the network state may differ between times when a neuron is refractory and times when it is not.
-This issue we can also correct for by calculating $ \beta $ for baseline bins with no stimulation and no stimulation during absolute refractory period.
-This will linearly correct for the differences in network state that are due to refractory periods while maintaining the causal validity of \cref{eq:wald}.
+To utilize the refractory period as an IV we first picked out one window of $ 4 $ ms for each of the pre-synaptic and post-synaptic neuron with a latency relative to stimulation time of $ 0 $ and $ \tau_{syn} + D $ ms (see \cref{eq:syn}) respectively.
+By classifying each window for each trial whether $x$ contained a spike we obtained the two binary arrays $ y_s $ and $ y_{sr} $.
 
-\todo[inline]{Note that if we assume that the stimulation is so strong that the network effect $ u_{ns} $ has a negligible effect on $ x_{s} $ and that due to synaptic and axonal delay the effect of $ u_{s} $ is unable to affect $ x_{s} $ in it's respective stimulus response time.
-Then a conventional method like logistic regression will work.}
-
-\subsection{Statistical testing on cross correlation histogram}
+\subsection{Cross correlation histogram}\label{sec:method:cch}
 The statistical tests giving the probabilities $ p_{diff} $ and $ p_{fast} $ were done according to \cite{Stark2009, English2017}.
 Briefly, to test if the cross correlation histogram (CCH) peak was significant we employed two tests.
-By using the Poisson distribution with a continuity correction \cite{Stark2009} given by \cref{eq:poissoncontcor} we calculated $ p_{diff} $ by comparing the peak in positive time lag with the maximum peak in negative time lag \cite{English2017} and the difference from the CCH convoluted with a hollow Gaussian \cite{Stark2009}.
+By using the Poisson distribution with a continuity correction \citep{Stark2009} given by \cref{eq:poissoncontcor} we calculated $ p_{diff} $ by comparing the peak in positive time lag with the maximum peak in negative time lag \citep{English2017}.
+The probability $ p_{fast} $ represents the difference between CCH and it's convolution with a hollow Gaussian kernel \citep{Stark2009}.
 \begin{equation}
@@
 Here $ \lambda $ represents the counts at bin $ m $ and $ N $ is the number of bins considered.
+To estimate the connection weight between pairs we used the spike transmission probability defined in \cite{English2017} as
+\begin{align}
+\label{eq:ptrans}
+P_{trans} = \frac{1}{n}\sum_{m=3ms}^{6ms} CCH(m) - \lambda_{Gauss}(m),
+\end{align}
+where $ n $ is the number of spikes detected in the presynaptic neuron and $\lambda_{Gauss}(m)$ is the CCH count convolved with a hollow Gaussian kernel at bin $m$.
+
+\subsection{Logistic regression}
+To utilize the refractory period without using it as an IV we estimated synaptic weights using a logistic regression.
+To do this we first picked out one window of $ 4 $ ms for each of the pre-synaptic and post-synaptic neuron with a latency relative to stimulation time of $ 0 $ and $ \tau_{syn} + D $ ms (see \cref{eq:syn} ) respectively.
+By classifying each window for each trial whether it contained a spike we obtained two binary arrays, the regressor $ x $ and the dependent variable $ y $ where we want to estimate the probability $ P(y = 1|x) $ by fitting the parameters $ \vec{\beta} $ such that
+\begin{align}
+y = \begin{cases}
+	1 & \text{if } \beta_0 + \beta_1x + u > 0\\
+    0 & \text{else}
+\end{cases}
+\end{align}
+where $ u $ is an error term. Further, we used the logit link function such that the the probability giving the proxy for synaptic weight is given by
+\begin{align}
+p(x) = \frac{1}{1 + e^{-(\beta_0 + \beta_1x)}}
+\end{align}
+The model was fitted using the python package scikit-learn \citep{scikit-learn}
+
 \subsection{Simulated network}
-The first simple network consists of three leaky integrate and fire (LIF) neurons given in \cref{eq:LIF}:
+To simulate a recurrent network of excitatory and inhibitory neurons we used the leaky integrate and fire (LIF) model given by
 \begin{equation}
-	\tau \dot{V}_{i}(t) = -(V_{i}(t) - E_{L}) + RI_{i}(t).
+	\de{V_m^i}{t} = - \frac{(V_m^i - E_L)}{\tau_m} + \frac{I_{syn}^i(t)}{C_m}.
     \label{eq:LIF}
 \end{equation}
-When the membrane potential $ V_{i} $ of neuron $ i $ reaches a threshold $ \theta $ an action potential is emitted and the depolarization is reset to the leak potential $ E_{L} $ followed by an absolute refractory period $ \tau_{rp} $.
-The membrane time constant is represented by $ \tau $ and $ I_{i}(t) $ denotes the synaptic currents arriving at the soma modeled as a sum of alpha functions given by
+When the membrane potential $ V_m^{i} $ of neuron $ i $ reaches a threshold $ V_{th} $ an action potential is emitted and $ V_m^{i} $ reset to the leak potential $ E_{L} $ followed by an absolute refractory period $ \tau_{ref} $.
+The membrane time constant is represented by $ \tau_{m} $ and $ I_{syn}^{i}(t) $ denotes the post synaptic current (PSC) for neuron $ i $ modeled as a sum of alpha functions given by
 \begin{equation}
 	\label{eq:syn}
-	RI_{i}(t) = \sum_{j}J_{ij}\sum_{k}t\exp(-(t-t_{j}^{k})/\tau_{syn}) - D.
-\end{equation}\todo{make sure delay is correctly inserted}
-Different synapses are denoted by $ j = 1,...,C+C_{ext} $ with post synaptic potential (PSP) amplitude (synaptic efficacy) denoted as $ J_{ij} $.
-When neuron $ j $ emits it's $ k $'th spike at time $ t_{j}^k $ it arrives at neuron $ i $ at $ t = t_{j}^k + D $ where $ D $ denotes a transmission delay.
-Furthermore all neurons are driven by an external Poisson drive and an Gaussian noise process given in \cref{eq:channelgauss}.
-\begin{equation}
-	I(t) = mean + std * N_j for t_0 + j dt \leq t < t_0 + (j-1) dt 
-    \label{eq:channelgauss}
+	I_{syn}^i(t) = \sum_{j=1}^C J_j \alpha(t - t_j - D),
 \end{equation}
-where $N_j$ are Gaussian random numbers with unit standard deviation and $t_0$ is 
-the device onset time.
-\todo[inline]{small perturbations on noisy networks will not yield a correlation between AC because the variance in stimulus response time in A will not be explained by the variance in spike times of C.
-Small variations of onset times in A relative to C will over many trials will wash out correlations between AC and at the same time strengthen correlations between BC.
-The only way to get a strong correlation between AC is to impose a short and strong stimulus effectively making AB spike coincidentally.}
+where $ t_j $ denotes an incoming spike through synapse $ j $ at delay $ D $ and $ C $ is the number of incoming synapses on neuron $ i $.
+The PSC amplitude is given by $ J_{j} $ and the alpha function is given by
+\begin{align}
+\tau_{syn}\alpha(t) = te^{-\frac{t}{\tau_{syn}}} H(t).
+\end{align}
+Here $ \tau_{syn} $ denotes the synaptic integration time constant and $ H $ is the Heaviside step function. All neurons were driven by an external Poisson process with rate $ rate_{p} $.
+
+Synaptic weights were log-normally distributed such that the increase in membrane potential $ V_m^i $ due to one spike were restricted to lie between $ V_{syn} = 0.05 mV $ and $ V_{syn} = 2.05 mV $ based on experimental findings \citep{Sayer1990,Mason1991}.
+The synaptic distribution is shown in \cref{fig:state}\labelcref{fig:syn_dist} where the inhibitory PSC amplitude is given by $ J_{in} = g J_{ex} $ where $ J_{ex} $ denotes the excitatory synaptic weight.
+
+To find suitable parameters yielding asynchronous activity we measured the population correlation coefficient given by
+
+\begin{align}
+\mean{CC}_{pop} = \mean{\mean{\frac{h_{i} - \mean{h_{i}}}{std(h_{i})}\frac{h_{j} - \mean{h_{j}}}{std(h_{j})}}}_{pop},
+\end{align}
+where $ h $ is the spike time histogram with binsize at $ 5ms $ for neuron $ i,j $ and $ \mean{\cdot} $ is the mean operator.
+The distribution of $ CC $ is shown in \cref{fig:cchvswald} which were found by performing several parameter sweeps picking three parameter sets which mainly differed in firing rate (data not shown).
+
+To further evaluate the network state we calculated the coefficient of variation (CV) of the population given by
+\begin{align}
+\mean{CV} = \mean{\frac{std(ISI_i)}{\mean{ISI_i}}}_{pop},
+\end{align}
+where $ ISI $ denotes the inter-spike interval of neuron $ i $.
+Due to the finite time synaptic integration time constant $ \tau_{syn} = 1 ms $ we were unable to have the network showing an irregular state; see \cref{fig:state}\labelcref{fig:CV}.
+To verify that indeed this was due to $ \tau_{syn} $ we performed several simulations with lower $ \tau_{syn} $ obtaining $ \mean{CV}_{pop} > 1 $ (data not shown).
+It would likely be easier to achieve irregular network state if synapses were conductance based \citep{Kumar2008}.
+However, we settled with current based synapses as we were mainly interested in achieving an asynchronized state ($ \mean{CC}_{pop} < 0.01 $).
+\begin{figure}
+\makeatletter
+\renewcommand\p@subfigure{}
+\makeatother
+\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{CC}
+\caption{} \label{fig:CC}
+\end{subfigure}\hfill
+\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{CV}
+\caption{} \label{fig:CV}
+\end{subfigure}\medskip
+
+\begin{subfigure}{\textwidth}\centering \includegraphics[scale=1]{syn_dist}
+\caption{} \label{fig:syn_dist}
+\end{subfigure}
+\caption{Network state \label{fig:state}}
+\end{figure}
+
+\begin{table}
+\begin{tabular}{lllll}
+\toprule
+{} &    model 1 &    model 2 &    model 3 &       units \\
+\midrule
+$N_{neurons}$         &       1250 &        &        &             \\
+$\Delta t$            &        0.1 &         &         &             \\
+$N_{ex}$              &       1000 &        &        &             \\
+$N_{in}$              &        250 &         &         &             \\
+$eta$                 &        0.9 &         &         &             \\
+$rate_{p}$            &    3694.26 &          &     &          Hz \\
+$V_{reset}$           &          0 &           &           &          mV \\
+$V_{m}$               &          0 &           &           &          mV \\
+$E_{L}$               &          0 &           &           &          mV \\
+$t_{ref}$             &          2 &           &           &          ms \\
+$\tau_{m}$            &         20 &          &          &          ms \\
+$V_{th}$              &         20 &          &          &          mV \\
+$C_{m}$               &          1 &           &           &          pF \\
+$V_{syn}$             &        0.2 &         &        &          mV \\
+$g$                   &        9.9 &        4.4 &          3 &             \\
+$V_{syn}^{high}$      &       2.05 &        &        &          mV \\
+$V_{syn}^{low}$       &       0.05 &        &       &          mV \\
+$var_{syn}$           &        0.5 &         &         &      mV$^2$ \\
+$\tau_{syn}^{in}$     &          1 &           &           &          ms \\
+$\tau_{syn}^{ex}$     &          1 &           &           &          ms \\
+$delay$               &        1.5 &         &        &          ms \\
+$eps$                 &        0.1 &        &         &             \\
+$C_{ex}$              &        100 &         &         &             \\
+$C_{in}$              &         25 &          &          &             \\
+$J_{in}$              &    0.88727 &   0.394342 &    0.26887 &          pA \\
+$J_{ex}$              &  0.0896232 &   &   &          pA \\
+$J_{high}^{ex}$       &   0.918638 &    &    &          pA \\
+$J_{low}^{ex}$        &  0.0224058 &   &   &          pA \\
+$J_{high}^{in}$       &   0.918638 &    &   &          pA \\
+$J_{low}^{in}$        &  0.0224058 &   &   &          pA \\
+$stim_{N}^{in}$       &          0 &           &           &             \\
+$stim_{N}^{ex}$       &        800 &         &         &             \\
+$stim_{amp}^{in}$     &          0 &           &           &          pA \\
+$stim_{amp}^{ex}$     &         10 &          &          &          pA \\
+$stim_{duration}$     &          2 &           &           &          ms \\
+$stim_{period}$       &        100 &         &         &          ms \\
+$stim_{max}^{period}$ &        150 &         &         &          ms \\
+$rate_{in}$           &    7.17024 &    9.66335 &     11.754 &          Hz \\
+$rate_{ex}$           &    9.05793 &    10.9176 &    13.1032 &          Hz \\
+$density$             &     100000 &      &      &  Nmm$^{-3}$ \\
+$S$                   &       10.3 &       &      &   mm$^{-1}$ \\
+$NA$                  &       0.37 &       &       &             \\
+$r$                   &        0.1 &         &         &      $\mu$m \\
+$n$                   &       1.36 &        &        &             \\
+\bottomrule
+\end{tabular}
+\caption{\label{tab:params} Simulation parameters of three different models.}
+\end{table}
+
+\subsection{Perturbation intensity}\label{sec:method:opto}
+In order to replicate an optogenetic experiment we modeled transmission of light through brain tissue with the Kubelka-Munk model for diffuse scattering in planar, homogeneous, ideal diffusing media given by
+\begin{align}
+T = \frac{1}{Sz + 1}.
+\end{align}
+Here $ T $ denotes a transmisison fraction, $ S $ is the scattering coefficient for mice \citep{Aravanis2007} and $ z $ is the distance from a light source \citep{Ho2017}.
+Further we combined diffusion with geometric loss assuming that absorption is negligible as in \cite{Aravanis2007} and computed the intensity as presented in \cref{fig:concept} by
+\begin{align}
+\label{eq:intensity}
+\frac{I(r)}{I(r=0)} = \frac{\rho^2}{(Sr + 1)(r + \rho)^2}
+\end{align}
+where $ r $ is the distance from the optical fiber and
+\begin{align}
+\rho = \frac{d}{2}\sqrt{\left(\frac{n}{NA}\right)^2 - 1}.
+\end{align}
+Here $ d $ is the diameter of the optical fiber, $ NA $ is the numerical aperture of the optical fiber and $ n $ is the refraction index for gray matter \citep{Ho2017}; see numerical values for parameters in \cref{tab:params}.
+
+To estimate the distribution of light intensity on affected neurons we assumed a neuron density of $ 10^4 Nmm^{-3} $ and found the volume of a cut cone that could contain the number of stimulated excitatory neurons which were found to yield the depth $ r_{max} = 0.175 mm $.
+We then selected $ N_stim $ neurons that were given a random position in the range $ [0, r_{max}] $ and were assigned a stimulation strength as the maximum stimulation strength multiplied by \cref{eq:intensity}.
+Then we selected $ 50 $ of the excitatory neurons that were not stimulated as the ``target" population which together with the inhibitory neurons were not perturbed directly by the light stimulus.
+
+In order to keep the stimulus model as simple as possible we let set the maximum stimulation strength to $ 10 pA $ which was found suitable by investigating the percentage of successful stimulations to be around $ 50\% $.
 
@@
 %-------------------------------------------------------------------------------
-\newpage
-\listoftodos
+%\newpage
+%\listoftodos
 \end{document}

### 2018-06-25 analyse.py

@@ -9,10 +9,14 @@ from exana.statistics.tools import ccg_significance
 from tqdm import tqdm
 
 data_path = 'results'
-# N_trials = 30000
-trials = [5000, 10000, 15000, 20000, 25000, 30000]
+trials = [30000]
+# trials = [5000, 10000, 15000, 20000, 25000, 30000]
 N = 100
-winsize = 4
+winsize_iv = 4
+# latency_iv = 4
+
+winsize_cch = 3
+latency_cch = 3
 binsize_corr = 5.
 min_stim = 1.
 
@@ -23,7 +27,8 @@ jobname = param_module.replace('.py', '')
 
 dataa = np.load(op.join(data_path, jobname + '.npz'))['data'][()]
 
-latency = dataa['params']['delay'] + dataa['params']['tau_syn_ex']
+latency_iv = dataa['params']['delay'] + dataa['params']['tau_syn_ex']
+
 
 conn = dataa['connections']
 spiketrains = {s['sender']: {'pop': pop, 'times': s['times']}
@@ -33,7 +38,7 @@ rec = spiketrains.keys()
 sources = [s for s in dataa['stim_nodes']['ex'] if s in rec][:int(N/2)]
 targets = [n for n in dataa['nodes']['ex'] if n not in sources and n in rec][:int(N/2)]
 assert len(sources) + len(targets) == N
-pbar = tqdm(total=int(N/2)**2 * len(trials))
+pbar = tqdm(total=int(N / 2)**2 * len(trials))
 for N_trials in trials:
     for source in sources:
         for target in targets: #NOTE different latency in inhibitory neurons
@@ -48,7 +53,7 @@ for N_trials in trials:
             spike_trains = [source_t[source_t <= t_stop],
                             target_t[target_t <= t_stop]]
             iv = IV(*spike_trains, stim_times,
-                     winsize=winsize, latency=latency)
+                     winsize=winsize_iv, latency=latency_iv)
             stim_amp = dataa['stim_amps'][spiketrains[source]['pop']]
             try:
                 lr = iv.logreg
@@ -78,7 +83,7 @@ for N_trials in trials:
                     *spike_trains, limit=15, binsize=1,
                     hollow_fraction=.6, width=10)
                 assert len(pcausal) == len(cch)
-                mask = (bins >= latency) & (bins <= latency + winsize)
+                mask = (bins >= latency_cch) & (bins <= latency_cch + winsize_cch)
[…]

### 2018-06-25 method.py

@@ -5,7 +5,7 @@ from sklearn.linear_model import LogisticRegression
 import seaborn as sns
 
 def hist_stim(stim_times, source, target, period, winsize, latency):
-    hist = np.zeros([len(stim_times), 4])
+    hist = np.zeros([len(stim_times), 3])
     src = np.searchsorted
     for idx, t in enumerate(stim_times):
         hist[idx,:] = (
@@ -16,11 +16,9 @@ def hist_stim(stim_times, source, target, period, winsize, latency):
           src(target, t + latency, 'left') <
           src(target, t + latency + winsize, 'right'),
           # no stim response
-          src(source, t + period - winsize - latency, 'left') <
-          src(source, t + period - latency, 'right'),
-          # no stim synaptic response
-          src(target, t + period - winsize, 'left') <
-          src(target, t + period, 'right'))
+          src(source, t - 2*winsize, 'left') <
+          src(source, t - winsize, 'right')
+          )
     return hist
 
 class IV:
@@ -58,8 +56,8 @@ class IV:
 
     @property
     def wald_ns(self):
-        ys = self.lams[self.NoStim, 3]
-        ysr = self.lams[self.NoStimRef, 3]
+        ys = self.lams[self.NoStim, 1]
+        ysr = self.lams[self.NoStimRef, 1]
         return(ys.mean() - ysr.mean())
 
     @property
@@ -71,7 +69,7 @@ class IV:
 
     @property
     def logreg_ns(self):
-        X, Y = self.lams[:, 2], self.lams[:, 3]
+        X, Y = self.lams[:-1, 2], self.lams[1:, 1]
         lr = LogisticRegression()
         lr.fit(X.reshape(-1, 1).astype(int), Y.astype(int))
         return lr

### 2018-06-25 params_1_nostim.py

@@ -1,0 +1,51 @@
+from quantities import s, ms, mV, pF, Hz, nS, pA
+import numpy as np
+
+parameters = {
+    'msd'             : 1234, # Master seed
+    'num_threads'     : 4,
+    'N_neurons'       : 1250,
+    'N_ex'            : 1000,
+    'N_in'            : 250,
+    'N_rec_spike_ex'  : None,
+    'N_rec_spike_in'  : None,
+    'N_rec_state_ex'  : None,
+    'N_rec_state_in'  : None,
+    'verbose'         : False,
+    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
+    'delay'           : 1.5, # Synaptic delay in ms
+    # Neuron parameters
+    't_ref'           : 2.0, # Duration of refractory period in ms
+    'V_m'             : 0.0, # Membrane potential, initial condition in mV
+    'E_L'             : 0.0, # Leak reversal potential in mV
+    'V_reset'         : 0.0, # Reset potential of the membrane in mV
+    'tau_m'           : 20.0, # Membrane timeconstant in ms
+    'C_m'             : 1.0, # Capacity of the membrane in pF
+    'V_th'            : 20.0, # Spike threshold in mV
+    'tau_syn_ex'      : 1., # Time constants of the excitatory synaptic exponential function in ms
+    'tau_syn_in'      : 1., # Time constants of the inhibitory synaptic exponential function in ms
+    # Connection parameters
+    'eta'             : .9, # external poisson rate in Hz
+    'J'               : .2, # mV
+    'g'               : 9.9,
+    'eps'            : 0.1, # connection prob
+    'J_high'         : 2.05, # max connection strength
+    'J_low'          : 0.05,
+    'p_var'          : .5, # percentage variation of mean in lognormal dist
+    # Stimulation parameters
+    'stim_N_ex'       : 800,
+    'stim_N_in'       : 0,
+    'stim_dist'       : 'poisson',
+    'stim_amp_ex'     : 0.0, # pA
+    'stim_amp_in'     : 0.0, # pA
+    'stim_period'     : 100.0, # ms
+    'stim_max_period' : 150, # only applies to poisson
+    'stim_duration'   : 2.0, # ms
+    'stim_N'          : 30000,
+    # Optogenetics
+    'r': 100e-3, # 100 um radius of fiber
+    'n': 1.36, # refraction index of gray matter
[…]

### 2018-06-25 params_3.py

@@ -33,7 +33,7 @@ parameters = {
     'J_low'          : 0.05,
     'p_var'          : .5, # percentage variation of mean in lognormal dist
     # Stimulation parameters
-    'stim_N_ex'       : 950,
+    'stim_N_ex'       : 800,
     'stim_N_in'       : 0,
     'stim_dist'       : 'poisson',
     'stim_amp_ex'     : 10.0, # pA

### 2018-06-25 simulator.py

@@ -22,6 +22,7 @@ except ImportError:
 
 nest.set_verbosity(100)
 
+
 def hasattrs(o, *args):
     if any(hasattr(o, a) for a in args):
         return True
@@ -244,9 +245,11 @@ class Simulator:
         nest.Connect(self.nodes_ex[:N_rec_ex], self.spikes_ex)
         nest.Connect(self.nodes_in[:N_rec_in], self.spikes_in)
 
-    def set_state_rec(self, sampling_period=None):
-        sampling_period = sampling_period or self.p['res']
+    def set_state_rec(self):
+        sampling_period = self.p.get('sampling_period') or self.p['res']
         def connect(nodes, label, N_rec):
+            if not N_rec:
+                return
             vm = nest.Create("multimeter",
                 params = {"interval": sampling_period,
                           "record_from": ['V_m'],
@@ -254,13 +257,14 @@ class Simulator:
                           "label": label, "withtime": True})
             nest.Connect(vm, nodes[:N_rec])
             return vm
-        N_rec_ex = self.p.get('N_rec_state_ex') or self.p['N_ex']
-        N_rec_in = self.p.get('N_rec_state_in') or self.p['N_in']
-        self.vm_ex = connect(self.nodes_ex, "exc_v", N_rec_ex)
-        self.vm_in = connect(self.nodes_in, "inh_v", N_rec_in)
+        self.vm_ex = connect(self.nodes_ex, "exc_v",
+                             self.p.get('N_rec_state_ex'))
+        self.vm_in = connect(self.nodes_in, "inh_v",
+                             self.p.get('N_rec_state_in'))
 
     def simulate_trials(self, progress_bar=False):
         progress_bar = progress_bar and PBAR
+        amps = [self.p['stim_amp_ex'], self.p['stim_amp_in']]
 
         def intensity(z):
             rho = self.p['r'] * np.sqrt((self.p['n'] / self.p['NA'])**2 - 1)
@@ -275,17 +279,17 @@ class Simulator:
             dV = A * dz
             N = dV * self.p['density']
             return np.array([np.sum(N[:i+1]) for i in range(len(N))])
-
-        z = np.arange(0,1.01,.01)
-        neurons = affected_neurons(z)
-        mask = neurons <= self.p['stim_N_ex'] + self.p['stim_N_in']
-        z_N = z[mask][-1]
+        if any(a > 0 for a in amps):
+            z = np.arange(0,1.01,.01)
[…]

### 2018-09-16 manuscript/main.tex
removed all deprecated code and results and renamed all notebooks to respective figure names

@@
 \usetikzlibrary{arrows,shapes,calc}
-\tikzset{line/.style = {draw, -latex', ultra thick}}
 \newcommand{\abs}[1]{\left|{#1}\right|}
@@
 %\title{The need to combine optogenetics with neural recordings to estimate causal connectivity}
-\author{Mikkel Elle Lepper\o{}d, Konrad Paul Kording}
+\author{Mikkel Elle Lepper\o{}d, Tristan St$\ddot{\text{o}}$ber, Konrad Paul Kording}
 
@@
 The interaction between optogenetic stimulation and the absolute refractory period produces a weak almost random signal (IV) that we exploit to solve the causal inference problem.
-In a simple simulation of integrate and fire neurons, we find that instrumental variable techniques estimate connectivity better than naive techniques (99.8\% vs 86\% correct).
+In a simple simulation of integrate and fire neurons, we find that instrumental variable techniques estimate connectivity better than naive techniques (99.8\% vs 0.6\% correct).
 Our instrumental variable approach is robust and easy to implement in the context of current experimental paradigms.
@@
 \section{Introduction}
+
+\todo[inline]{Cite Konrad's new paper}
 % neuroscience experiments are observational which implicates a certain set of statistical tools
@@
 Hence, the stimulus will produce a distributed pattern of activity.
-For example, if we focus a stimulation beam on one neuron, there will be a cone of light in front of and behind the neuron, which comes from the rays of light from the lens focused on the cell and again coming out on the backside..
+For example, if we focus a stimulation beam on one neuron, there will be a cone of light in front of and behind the neuron, which comes from the rays of light from the lens focused on the cell and again coming out on the backside.
 This distributed pattern of stimulation produces activity which then percolates through the network of neurons.
@@
 only have a strong effect on spiking activity when the membrane potential of the cell is sufficiently close to the firing threshold.
-The induced effect of optogenetic stimulation as a function of distance should be the product of four parameters:  light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and opsin distribution across neurons.
+The induced effect of optogenetic stimulation as a function of distance should be given by a relation of the four parameters: light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and opsin distribution across neurons.
 
-To estimate the light intensity we calculated the spatial extent of laser light delivered by fiber-optics under plausible experimental conditions according to \cite{Aravanis2007}; see \cref{sec:method:opto}.
+To estimate the light intensity we calculated the spatial extent of laser light delivered by fiber-optics under plausible experimental conditions according to \citet{Aravanis2007}; see \cref{sec:method:opto}.
 This modeling of light intensity yield an approximately $ 1/r^2 $ reduction with distance $ r $ from the stimulation site \cref{fig:concept} (cyan line).
-This is explained by the surface of the 3D shell growing with $4\pi r^2$ and photons will be roughly isotropic beyond the scattering length \cref{fig:concept} (inset).
+This is explained by the surface of the 3D shell growing with $4\pi r^2$ and photons will be roughly isotropic (but see \citep{thunemann2018does}) beyond the scattering length \cref{fig:concept} (inset).
 The same number of photons has to cross each of the spheres around the stimulation location unless they are absorbed.
@@
 The response that is expected due to a pulse of light that induces a charge $Q$ should be proportional to the density of neurons whose membrane potential sit within a $Q/C$ range of the threshold ($C$ is the capacitance).
-The fact that the distribution of membrane potentials is relatively flat (the density close to the threshold is generally within an order of magnitude of the density of its mode) suggests that the response in membrane potential to a perturbation for any neuron is roughly proportional to the light intensity.
+The fact that the distribution of membrane potentials is relatively flat (the density close to the threshold is generally within an order of magnitude of the density of its mode) suggests that the spiking response to a perturbation for any neuron is roughly proportional to the induced photo current.
+The peak amplitude of the photo current relates non-linearly to the light intensity and can be modelled by the Hill equation \citep{wang2007high}.
 Moreover, assuming that opsins are evenly distributed along the neuron, the opsin density in the neuron population will be relatively even.
-Based on this, we calculate the overall stimulation effect to be the product of neuron density and light intensity which is approximately constant in distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line).
-Thus, optogenetic stimulation utilizing single photon activation does not actually produce a localized effect.
+Based on this, we calculate the overall stimulation effect to be the product of the number of neurons in a spherical slice and the peak amplitude photo current which actually increases with distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line).
+Thus, optogenetic stimulation utilizing single photon activation does not produce a localized effect.
 
@@
 \caption{{\bf Spatial extent of optogenetic stimulus}.
-Due to scattering and geometric loss the light intensity (I, cyan line) plotted as percentage of intensity exiting the optogenetic fiber follows approximately an inverse square law $ r^{-2} $ where $ r $ is the distance from the fiber.
-If neurons are uniformly distributed the number of affected neurons increase by $ r^{2} $ (N, black line) rendering the probability of activating a neuron approximately constant (IN, blue line).
+Due to scattering and geometric loss the light intensity (I, cyan line) with an intensity of $10 mW/mm^2$ exiting the optogenetic fiber follows approximately an inverse square law $ r^{-2} $ where $ r $ is the distance from the fiber.
+If neurons are uniformly distributed the number of affected neurons increase by $ r^{2} $ (N, black line).
+The total photo current (sum of single neuron currents) thus increases by distance due to the nonlinear relation between photo intensity and photo current (P, blue line).
 \label{fig:concept}}
 \end{figure}
-
+\FloatBarrier
 \subsection{Confounding as a problem for the estimation of causal effects}
 When we stimulate many neurons at the same time, and observe a post-synaptic neuron to be active after our stimulation, it is hard to know which of the stimulated neurons produced the activity.
-To illustrate such confounding effects we simulated a network comprised of three neurons (A,B, and C).
+To illustrate such confounding effects we simulated a network comprised of three neurons (A,B, and C) \cref{fig:intro}\labelcref{fig:intro:1}.
 The neurons receive Poisson spike trains and have added Gaussian white noise to the membrane potential.
-We also gave them interactions, where spikes of neuron B increase the probability of firing for neuron C but there are no other interactions \cref{fig:intro}\labelcref{fig:intro:1}.
+We also gave them interactions, where spikes of neuron B increase the probability of firing for neuron C but there are no other interactions.
 Finally, we allowed simulated optogenetic stimulation to affect neurons A and B (but not C). We thus have a simple system for exploring questions of causality.
@@
 \end{figure}
-
+\FloatBarrier
 \subsection{Instrumental variables to resolve confounding}
@@
 For such a simple system, it produces meaningful estimates of the causal interactions between neurons.
-
+\tikzset{line/.style = {draw, -latex', ultra thick}}
 \begin{figure}
@@
 \begin{subfigure}{0.485\textwidth} \centering
-% * <tristan.stoeber@posteo.net> 2018-06-24T21:23:23.987Z:
-% What is S?
-% ^.
 \begin{tikzpicture}
@@
 The IV estimator calculated by \cref{eq:wald} converges to $ \hat{\beta}_{BC} \approx 0.2, \hat{\beta}_{AC} \approx 0 $ after approximately $ 5000 $ trials as seen in \labelcref{fig:cchvswald:1,fig:cchvswald:2} respectively.
-% I would have put in a take home message / interpretation of the results in the figure text.
-Insets represent high resolution zoom of cross correlation histograms of BC and AC where horizontal and vertical axes represents time lag and counts of coincident spikes in bins of 0.1 ms respectively.
  \label{fig:cchvswald}}
 \end{figure}
-
+\FloatBarrier
 \subsection{Larger simulated networks}
@@
 The trial onset had a temporal Poisson distribution with period 100 ms and was further clipped between 100-150 ms.
-For weight estimates we randomly selected among the excitatory population 50 stimulated neurons and 50 unstimulated neurons.
+For weight estimates we randomly selected among the excitatory population 100 stimulated neurons and 100 unstimulated neurons.
 
@@
 \begin{subfigure}{\textwidth} 
-\includegraphics[scale=1]{mse_pos}
+\includegraphics[scale=1]{mse}
 \caption{} \label{fig:error:mse}
@@
 \labelcref{fig:error:mse} The IV- and CCH estimator is evaluated for the asynchronous recurrent neural network at three different amounts of relative inhibition $ g $.
-The MSE as a function of number of trials is shown on a logarithmic scale where the slopes was found to be $ -0.77, -0.48, -0.40 $ for $g = 9.9, 4.4, 3.0$ respectively.
+The MSE as a function of number of trials is shown on a logarithmic scale where the slopes was found to be $ -0.67, -0.69, -0.45 $ for $g = 9.9, 4.4, 3.0$ respectively.
 \label{fig:error}}
@@
 We want to compare the IV estimator which exploits the refractory period with the CCH method given by \cref{eq:ptrans} which ignores network confounding.
+We defined hit rate as the relative amount of stimulations in non-refractory states and required a minimum of 90$\%$.
 To indicate the amount of connections which are falsely attributed a non-zero weight we calculated the amount of false positives.
-This was given as the percentage of estimated synapses larger than $ 0.05 $ where the true weight was $ 0 $, finding $ 13.3\% $ for CCH and $ 0.2\% $ for the IV estimator; see \cref{fig:network-class}\labelcref{fig:network-class:1}.
-In addition we compared the size of the estimations at false positive instances and found that the CCH have slightly higher median.
+This was given as the percentage of estimated synapses larger than $ 0.05 $ where the true weight was $ 0 $, finding $ 99.4\% $ for CCH and $ 0.2\% $ for the IV estimator; see \cref{fig:network-class}\labelcref{fig:network-class:1}.
+In addition we compared the size of the estimations at false positive instances and found that the CCH have significantly higher median than IV (p=0, difference ($\Delta$) = 0.17, permutation resampling \citep{wassermann2006all}).
 The IV approach, while not being perfect, thus outperforms the simple CCH approach.
@@
 We thus performed a logistic regression as seen in \cref{fig:network-class}\labelcref{fig:network-class:1} denoted LOGIT.
-As seen here, LOGIT performs even worse than CCH showing that it really helps to use the refractory period as an instrumental variable.
-To further evaluate the methods we calculated false negatives as instances where the true weight is non-zero but estimated to be equal to zero in \cref{fig:network-class}\labelcref{fig:network-class:2} showing that the CCH and IV estimators performs equally well.
+As seen here, LOGIT performs even worse than CCH showing that it really helps to use the refractory period as an instrumental variable (p=0, $\Delta$ = 0.7, permutation resampling).
+To further evaluate the methods we calculated false negatives as instances where the true weight is non-zero but estimated to be equal to zero in \cref{fig:network-class}\labelcref{fig:network-class:2} showing that the CCH and IV estimators performs equally well (p=0.78, $\Delta$ = 0.021, permutation resampling).
 Finally we wanted to evaluate the estimated weights as a function of true weights shown in \cref{fig:network-class}\labelcref{fig:network-class:3,fig:network-class:4} after 30000 trials.
-The IV estimator yields a good prediction $ r^2 = 0.55 $, while the CCH estimator performs surprisingly bad with $ r^2 = 0.002 $.
+The IV estimator yields a good prediction ($ R^2 = 0.845 $), while the CCH mainly estimates the strength of stimulation while true weights are poorly estimated ($ R^2 = 0.048 $).
 Utilizing refractory periods as an (imperfect) instrumental variable considerably improves estimates.
@@
 \end{subfigure}
-% * <tristan.stoeber@posteo.net> 2018-06-24T21:29:41.690Z:
-% 
-% Why is logit not considered for false negatives?
-% 
-% ^.
 
 \medskip
-\begin{subfigure}{0.46\textwidth} \includegraphics[scale=1]{fit_wald_pos}
+\begin{subfigure}{0.46\textwidth} \includegraphics[scale=1]{fit_wald}
 \caption{} \label{fig:network-class:3}
 \end{subfigure}\hfill
-\begin{subfigure}{0.6\textwidth} \includegraphics[scale=1]{fit_cch_pos}
+\begin{subfigure}{0.6\textwidth} \includegraphics[scale=1]{fit_cch}
 \caption{} \label{fig:network-class:4}
@@
 Shaded area show the 95 \% confidence interval calculated by bootstrapping.
+Data were obtained from simulation of model 2 \cref{tab:params}.
 \label{fig:network-class}}
 \end{figure}
+\FloatBarrier
+\subsection{Application to hippocampal recordings with Optotrodes}
+% % aim: test method on state of the art, publicly available data of optostimulation
+%   % compare results for transmission probability and wald estimate
+\begin{figure}
+\makeatletter
+\renewcommand\p@subfigure{}
+\makeatother
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=1.\textwidth]{Optotrode.jpg}
+\caption{} \label{}
+\end{subfigure}\hfill
+\begin{subfigure}{0.6\textwidth}
+\includegraphics[width=1.\textwidth]{Optodata_IVvsCCH.pdf}
+\caption{} \label{}
+\end{subfigure}
 
+\caption{}.
+\label{fig:optotrodedata}
+ \end{figure}
+\FloatBarrier
 \section{Discussion}
@@
 %Still some troubles
-We found some negative values of the IV estimator which were suppressed as we knew that we only stimulated excitatory neurons.
-This happens because some neurons  have correlated refractory times which we hypothesize to be mainly due to synchrony induced by the stimulation \citep{ermentrout2008reliability}.
-This hypothesis was strengthened by observing that this negative values did not occur when the stimulation intensity was set to zero (data not shown).
+We found some negative values of the IV estimator which were largely suppressed by requiring a hit rate $ < 90\%$.
+Hit rates larger than $90\%$ happens mainly at strong stimulation intensity which lower statistical power in the IV and can lead to correlated refractory times.
+We hypothesize strong stimulations can lead to synchrony induced by the stimulation \citep{ermentrout2008reliability}.
+This hypothesis was strengthened by observing that the negative values did not occur when the stimulation intensity was set to zero (data not shown).
 Furthermore, respective simulated neural network introduces much response overlap due to synapses having identical synaptic time constants and transfer delays.
@@
 \section{Methods}
+\tikzset{line/.style = {draw, -latex', thick}}
 \subsection{Instrumental variable estimation}
@@
 Here $ y $ is the dependent variable, $ x $ is the explanatory variable, $ \beta $ is the effect of the $ x$ on $y $ and $ u $ is an unknown error term.
-\Cref{eq:regress} follows from the causal path diagram (invented by \cite{wright1921correlation})
+\Cref{eq:regress} follows from the causal path diagram \citep{wright1921correlation,wright1923theory}
 \begin{center}
@@
     \node [above right = of u] (y) {$ y $};
-    \path [line] (x) -- (y);
+    \path [line] (x) -- (y) node [midway, above] {$ \beta $};
     \path [line] (u) -- (y);
@@
     \node [above right = of u] (y) {$ y $};
-    \path [line] (x) -- (y);
+    \path [line] (x) -- (y) node [midway, above] {$ \beta $};
     \path [line] (u) -- (x);
@@
 Let us assume that the stimulus renders only a subset of $u$ correlated with $x$, namely $u_s$ ($s$ denotes stimulated).
-To disassociate $x$ from $u_s$ we need something that can distinguish between different neurons that are stimulated.
-We thus require some instrument $ x_{s_r} $ which is (1) uncorrelated with the network $ u $ and (2) is correlated with the regressor $ x $ \citep{angrist2008mostly}.
+To disassociate $x_s$ from $u_s$ we need something that can distinguish between different neurons that are stimulated.
+We thus require some instrument $ x_{s_r} $ which is (1) uncorrelated with the network $ u $, (2) is correlated with the regressor $ x_s $ and (3) not correlated with $ y $ \citep{angrist2008mostly}.
 We assume that the neurons are independent at small time scales and that stimulation additionally randomize membrane potential individually in neurons.
-We may thus use the fact that a neuron that has fired just before the stimulation will be in an absolute refractory state and hence have $ x_{s_r}=0 $ independently of $u$, where the subscript $s_r$ denotes stimulation during refractory state.
+We may thus use the fact that a neuron that has fired just before the stimulation will be in an absolute refractory state and hence have $ x_{s_r}=0 $ independently of $ u $, where the subscript $ s_r $ denotes stimulation during refractory state.
 This introduces times where the spike from one of the stimulated neurons are missing.
@@
     \node (u) {$ u_s $};
-    \node [above = of u] (x) {$ x $};
+    \node [above = of u] (x) {$ x_s $};
     \node [above right = of u] (y) {$ y $};
     \node [left = of x] (sr) {$ x_{s_r} $};
-    \path [line] (x) -- (y);
+    \path [line] (x) -- (y) node [midway, above] {$ \beta $};
     \path [line] (u) -- (x);
@@
 \end{center}
-Here $ x_r $ represent times where the pre-synaptic neuron is refractory during stimulation.
-This is then an estimator that compares the post-synaptic activity when a given neuron is non-refractory with the post-synaptic activity when it is refractory, thus removing the confounding.
-The true $ \beta $ is given by
-% * <tristan.stoeber@posteo.net> 2018-06-24T22:11:38.685Z:
-% Shouldnt it be  $ x_s_r $
-% ^.
+Here $ x_{s_r} $ represent times where the pre-synaptic neuron is refractory during stimulation. The true $ \beta $ is given by \citep{wright1928tariff}
+
 \begin{equation}
@@
 \end{equation}
-Since our instrument $ x_{s_r} $ is binary we may use the IV (or more precisely Wald) estimator \citep{wald1940fitting} to estimate $ \beta_{IV} $ by
+This is then an estimator that compares the post-synaptic activity when a given neuron is non-refractory with the post-synaptic activity when it is refractory, thus removing the confounding.
+
+Since our instrument $ x_{s_r} $ is binary we may calculate the IV (or more precisely Wald) estimator \citep{wald1940fitting} $ \beta_{IV} $ by
 \begin{equation}
@@
 The probability $ p_{fast} $ represents the difference between CCH and it's convolution with a hollow Gaussian kernel \citep{Stark2009}.
+These two measures of significance were required to be $<0.01$
 \begin{equation}
@@
 
+\subsection{Calculating errors}
+\subsubsection{Mean square error}
+The conditional probability of neuron $y$ firing given a spike from neuron $x$ denoted $P(y|x)$ is related to the connection strength $w_{xy}$ and the background activity.
+Since LIF neurons integrate linearly and have fixed thresholds we expect $P(y|x)$ to be proportional to the connection strength $w_{xy}$, i.e.
+$P(y|x) ~ w_{xy}$.
+To calculate the mean squared errors we thus normalized $w_{xy}$ to the range $[0, max(IV_{xy})]$ or $[0, max(CCH_{xy})]$ for the IV and CCH estimators respectively by
+\begin{equation*}
+    \frac{1}{N}\sum(\hat{w} - \bar{w})^2).
+\end{equation*}
+Here $\bar{w}$ is the normalized weight and $\hat{w}$ is the estimated connection coefficient.
+
+\subsubsection{Goodness of fit}
+The goodness of fit was indicated by the $R^2$ value of a linear regression calculated by ordinary least squares \citep{seabold2010statsmodels} of the relation between estimated values (IV or CCH) with the true weight.
+
+\subsubsection{False positives}
+
+\subsubsection{False negatives}
+
 \subsection{Perturbation intensity}\label{sec:method:opto}
@@
 \begin{align}
-T = \frac{1}{Sz + 1}.
+T = \frac{1}{Sr + 1}.
 \end{align}
-Here $ T $ denotes a transmisison fraction, $ S $ is the scattering coefficient for mice \citep{Aravanis2007} and $ z $ is the distance from a light source \citep{Ho2017}.
+Here $ T $ denotes a transmisison fraction, $ S $ is the scattering coefficient for mice \citep{Aravanis2007} and $ r $ is the distance from a light source \citep{Ho2017}.
 Further we combined diffusion with geometric loss assuming that absorption is negligible as in \cite{Aravanis2007} and computed the intensity as presented in \cref{fig:concept} by
@@
 
-In order to keep the stimulus model as simple as possible we let set the maximum stimulation strength to $ 10 pA $ which was found suitable by investigating the percentage of successful stimulations to be around $ 50\% $.
+In order to keep the stimulus model as simple as possible we let set the maximum stimulation strength to $ 10 pA $ which was found suitable by investigating the percentage of successful stimulations to be almost $ 100\% $.
 

### 2018-09-16 generate_triple_data.py

@@ -1,0 +1,171 @@
+from tools import fix_figure, savefig, despine, set_style
+import nest
+import matplotlib.pyplot as plt
+from matplotlib.gridspec import GridSpec, GridSpecFromSubplotSpec
+import pandas as pd
+import numpy as np
+import quantities as pq
+import seaborn as sns
+
+from exana.statistics.tools import ccg_significance
+from exana.statistics.plot import plot_xcorr
+from exana.stimulus.plot import plot_psth
+# from exana.statistics import plot_xcorr
+import copy
+
+par = {
+    'msd'            : 1234, # Master seed
+    'num_threads'    : 4,
+    'N_neurons'      : 3,
+    'res'            : 0.1, # Temporal resolution for simulation Delta t in ms
+    'delay'          : 1.5, # Synaptic delay in ms
+    'rate_p'         : 1000., # external poisson rate in Hz
+    'J_p'            : .3,
+    'gauss_mean'     : .1,
+    'gauss_std'      : 1.1,
+    # Neuron parameters
+    't_ref'          : 2.0, # Duration of refractory period in ms
+    'V_m'            : 0.0, # Membrane potential, initial condition in mV
+    'E_L'            : 0.0, # Leak reversal potential in mV
+    'V_reset'        : 0.0, # Reset potential of the membrane in mV
+    'tau_m'          : 20.0, # Membrane timeconstant in ms
+    'C_m'            : 1.0, # Capacity of the membrane in pF
+    'V_th'           : 20.0, # Spike threshold in mV
+    'tau_syn_ex'     : 1., # Time constants of the excitatory synaptic exponential function in ms
+    'tau_syn_in'     : 1., # Time constants of the inhibitory synaptic exponential function in ms
+    # Connection parameters
+    'J_AB'           : 0., # mV
+    'J_AC'           : 0,
+    'J_BC'           : 2.,
+    'J_BA'           : 0,
+    'J_CA'           : 0,
+    'J_CB'           : 0,
+    'C'              : 1, # indegree
+    # Stimulation parameters
+    'stim_amp_A'      : 10.0, # pA
+    'stim_amp_B'      : 10.0, # pA
+    'stim_period'     : 30.0, # ms
+    'stim_max_period' : 60.0, # only applies to poisson
+    'stim_duration'   : 2.0, # ms
+    'stim_N'          : 30000,
+    'stim_dist'       : 'poisson'
+}
+
+
+def poisson_clipped(N, period, low, high):
+    poisson = []
[…]

### 2018-09-16 method.py

@@ -33,6 +33,10 @@ class IV:
             putative receiver neuron
         stim_times : array
             stimulation times
+        winsize : float
+            size of window around PSTH
+        latency : float
+            time untill response of post-synaptic neuron (begining of PSTH peak)
         '''
         p = {
             'width': 10,

### 2018-09-16 params_1_nostim.py

@@ -1,51 +1,0 @@
-from quantities import s, ms, mV, pF, Hz, nS, pA
-import numpy as np
-
-parameters = {
-    'msd'             : 1234, # Master seed
-    'num_threads'     : 4,
-    'N_neurons'       : 1250,
-    'N_ex'            : 1000,
-    'N_in'            : 250,
-    'N_rec_spike_ex'  : None,
-    'N_rec_spike_in'  : None,
-    'N_rec_state_ex'  : None,
-    'N_rec_state_in'  : None,
-    'verbose'         : False,
-    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
-    'delay'           : 1.5, # Synaptic delay in ms
-    # Neuron parameters
-    't_ref'           : 2.0, # Duration of refractory period in ms
-    'V_m'             : 0.0, # Membrane potential, initial condition in mV
-    'E_L'             : 0.0, # Leak reversal potential in mV
-    'V_reset'         : 0.0, # Reset potential of the membrane in mV
-    'tau_m'           : 20.0, # Membrane timeconstant in ms
-    'C_m'             : 1.0, # Capacity of the membrane in pF
-    'V_th'            : 20.0, # Spike threshold in mV
-    'tau_syn_ex'      : 1., # Time constants of the excitatory synaptic exponential function in ms
-    'tau_syn_in'      : 1., # Time constants of the inhibitory synaptic exponential function in ms
-    # Connection parameters
-    'eta'             : .9, # external poisson rate in Hz
-    'J'               : .2, # mV
-    'g'               : 9.9,
-    'eps'            : 0.1, # connection prob
-    'J_high'         : 2.05, # max connection strength
-    'J_low'          : 0.05,
-    'p_var'          : .5, # percentage variation of mean in lognormal dist
-    # Stimulation parameters
-    'stim_N_ex'       : 800,
-    'stim_N_in'       : 0,
-    'stim_dist'       : 'poisson',
-    'stim_amp_ex'     : 0.0, # pA
-    'stim_amp_in'     : 0.0, # pA
-    'stim_period'     : 100.0, # ms
-    'stim_max_period' : 150, # only applies to poisson
-    'stim_duration'   : 2.0, # ms
-    'stim_N'          : 30000,
-    # Optogenetics
-    'r': 100e-3, # 100 um radius of fiber
-    'n': 1.36, # refraction index of gray matter
[…]

### 2018-09-16 params_AI.py

@@ -1,48 +1,0 @@
-from quantities import s, ms, mV, pF, Hz, nS, pA
-import numpy as np
-
-parameters = {
-    'msd'             : 1234, # Master seed
-    'num_threads'     : 4,
-    'N_neurons'       : 1250,
-    'N_ex'            : 1000,
-    'N_in'            : 250,
-    'N_rec_spike_ex'  : None,
-    'N_rec_spike_in'  : None,
-    'N_rec_state_ex'  : None,
-    'N_rec_state_in'  : None,
-    'verbose'         : False,
-    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
-    'delay'           : 1.5, # Synaptic delay in ms
-    # Neuron parameters
-    't_ref'           : 2.0, # Duration of refractory period in ms
-    'V_m'             : 0.0, # Membrane potential, initial condition in mV
-    'E_L'             : 0.0, # Leak reversal potential in mV
-    'V_reset'         : 0.0, # Reset potential of the membrane in mV
-    'tau_m'           : 20.0, # Membrane timeconstant in ms
-    'C_m'             : 1.0, # Capacity of the membrane in pF
-    'V_th'            : 20.0, # Spike threshold in mV
-    'tau_syn_ex'      : 1., # Time constants of the excitatory synaptic exponential function in ms
-    'tau_syn_in'      : 1., # Time constants of the inhibitory synaptic exponential function in ms
-    # Connection parameters
-    'eta'             : .9, # external poisson rate in Hz
-    'J'               : .2, # mV
-    'g'               : 9.9,
-    'eps'            : 0.1, # connection prob
-    'J_high'         : 2.05, # max connection strength
-    'J_low'          : 0.05,
-    'p_var'          : .5, # percentage variation of mean in lognormal dist
-    # Stimulation parameters
-    'stim_nodes_ex'   : tuple(np.arange(1,602,1)),
-    'stim_nodes_in'   : (),
-    'stim_dist'       : 'poisson',
-    'stim_amp_ex'     : 10.0, # pA
-    'stim_amp_in'     : 0.0, # pA
-    'stim_period'     : 100.0, # ms
-    'stim_max_period' : 150, # only applies to poisson
-    'stim_duration'   : 2.0, # ms
-    'stim_N'          : 30000,
-    # analysis params
-    'latency': 6,
-    'winsize': 4
-}

### 2018-09-16 params_AI_longstim.py

@@ -1,46 +1,0 @@
-from quantities import s, ms, mV, pF, Hz, nS, pA
-import numpy as np
-
-parameters = {
-    'msd'             : 1234, # Master seed
-    'num_threads'     : 4,
-    'N_neurons'       : 1250,
-    'N_ex'            : 1000,
-    'N_in'            : 250,
-    'N_rec_spike_ex'  : None,
-    'N_rec_spike_in'  : None,
-    'N_rec_state_ex'  : None,
-    'N_rec_state_in'  : None,
-    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
-    'delay'           : 1.5, # Synaptic delay in ms
-    'eta'             : 1.0, # external poisson rate in Hz
-    # Neuron parameters
-    't_ref'           : 2.0, # Duration of refractory period in ms
-    'V_m'             : 0.0, # Membrane potential, initial condition in mV
-    'E_L'             : 0.0, # Leak reversal potential in mV
-    'V_reset'         : 0.0, # Reset potential of the membrane in mV
-    'tau_m'           : 20.0, # Membrane timeconstant in ms
-    'C_m'             : 1.0, # Capacity of the membrane in pF
-    'V_th'            : 20.0, # Spike threshold in mV
-    'tau_syn_ex'      : 4., # Time constants of the excitatory synaptic exponential function in ms
-    'tau_syn_in'      : 4., # Time constants of the inhibitory synaptic exponential function in ms
-    # Connection parameters
-    'J'               : .5, # mV
-    'g'               : 9.4,
-    'eps'            : 0.1, # connection prob
-    'J_high'         : 15., # max connection strength (low is 0)
-    'p_var'          : .5, # percentage variation of mean in lognormal dist
-    # Stimulation parameters
-    'stim_nodes_ex'   : tuple(np.arange(1,102,1)),
-    'stim_nodes_in'   : (),
-    'stim_dist'       : 'poisson',
-    'stim_amp_ex'     : 5.0, # pA
-    'stim_amp_in'     : 0.0, # pA
-    'stim_period'     : 30.0, # ms
-    'stim_max_period' : 100, # only applies to poisson
-    'stim_duration'   : 10.0, # ms
-    'stim_N'          : 30000,
-    # channelnoise
-    'gauss_mean'      : 0.,
-    'gauss_std'       : 0.
-}

### 2018-09-16 params_AI_reglongstim.py

@@ -1,46 +1,0 @@
-from quantities import s, ms, mV, pF, Hz, nS, pA
-import numpy as np
-
-parameters = {
-    'msd'             : 1234, # Master seed
-    'num_threads'     : 4,
-    'N_neurons'       : 1250,
-    'N_ex'            : 1000,
-    'N_in'            : 250,
-    'N_rec_spike_ex'  : None,
-    'N_rec_spike_in'  : None,
-    'N_rec_state_ex'  : None,
-    'N_rec_state_in'  : None,
-    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
-    'delay'           : 1.5, # Synaptic delay in ms
-    'eta'             : 1.0, # external poisson rate in Hz
-    # Neuron parameters
-    't_ref'           : 2.0, # Duration of refractory period in ms
-    'V_m'             : 0.0, # Membrane potential, initial condition in mV
-    'E_L'             : 0.0, # Leak reversal potential in mV
-    'V_reset'         : 0.0, # Reset potential of the membrane in mV
-    'tau_m'           : 20.0, # Membrane timeconstant in ms
-    'C_m'             : 1.0, # Capacity of the membrane in pF
-    'V_th'            : 20.0, # Spike threshold in mV
-    'tau_syn_ex'      : 4., # Time constants of the excitatory synaptic exponential function in ms
-    'tau_syn_in'      : 4., # Time constants of the inhibitory synaptic exponential function in ms
-    # Connection parameters
-    'J'               : .5, # mV
-    'g'               : 9.4,
-    'eps'            : 0.1, # connection prob
-    'J_high'         : 15., # max connection strength (low is 0)
-    'p_var'          : .5, # percentage variation of mean in lognormal dist
-    # Stimulation parameters
-    'stim_nodes_ex'   : tuple(np.arange(1,102,1)),
-    'stim_nodes_in'   : (),
-    'stim_dist'       : None,
-    'stim_amp_ex'     : 5.0, # pA
-    'stim_amp_in'     : 0.0, # pA
-    'stim_period'     : 30.0, # ms
-    'stim_max_period' : 100, # only applies to poisson
-    'stim_duration'   : 10.0, # ms
-    'stim_N'          : 30000,
-    # channelnoise
-    'gauss_mean'      : 0.,
-    'gauss_std'       : 0.
-}

### 2018-09-16 params_AI_regstim.py

@@ -1,46 +1,0 @@
-from quantities import s, ms, mV, pF, Hz, nS, pA
-import numpy as np
-
-parameters = {
-    'msd'             : 1234, # Master seed
-    'num_threads'     : 4,
-    'N_neurons'       : 1250,
-    'N_ex'            : 1000,
-    'N_in'            : 250,
-    'N_rec_spike_ex'  : None,
-    'N_rec_spike_in'  : None,
-    'N_rec_state_ex'  : None,
-    'N_rec_state_in'  : None,
-    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
-    'delay'           : 1.5, # Synaptic delay in ms
-    'eta'             : 1.0, # external poisson rate in Hz
-    # Neuron parameters
-    't_ref'           : 2.0, # Duration of refractory period in ms
-    'V_m'             : 0.0, # Membrane potential, initial condition in mV
-    'E_L'             : 0.0, # Leak reversal potential in mV
-    'V_reset'         : 0.0, # Reset potential of the membrane in mV
-    'tau_m'           : 20.0, # Membrane timeconstant in ms
-    'C_m'             : 1.0, # Capacity of the membrane in pF
-    'V_th'            : 20.0, # Spike threshold in mV
-    'tau_syn_ex'      : 4., # Time constants of the excitatory synaptic exponential function in ms
-    'tau_syn_in'      : 4., # Time constants of the inhibitory synaptic exponential function in ms
-    # Connection parameters
-    'J'               : .5, # mV
-    'g'               : 9.4,
-    'eps'            : 0.1, # connection prob
-    'J_high'         : 15., # max connection strength (low is 0)
-    'p_var'          : .5, # percentage variation of mean in lognormal dist
-    # Stimulation parameters
-    'stim_nodes_ex'   : tuple(np.arange(1,602,1)),
-    'stim_nodes_in'   : (),
-    'stim_dist'       : None,
-    'stim_amp_ex'     : 5.0, # pA
-    'stim_amp_in'     : 0.0, # pA
-    'stim_period'     : 30.0, # ms
-    'stim_max_period' : 100, # only applies to poisson
-    'stim_duration'   : 2.0, # ms
-    'stim_N'          : 30000,
-    # channelnoise
-    'gauss_mean'      : 0.,
-    'gauss_std'       : 0.
-}

### 2018-09-16 params_A~I.py

@@ -1,48 +1,0 @@
-from quantities import s, ms, mV, pF, Hz, nS, pA
-import numpy as np
-
-parameters = {
-    'msd'             : 1234, # Master seed
-    'num_threads'     : 4,
-    'N_neurons'       : 1250,
-    'N_ex'            : 1000,
-    'N_in'            : 250,
-    'N_rec_spike_ex'  : None,
-    'N_rec_spike_in'  : None,
-    'N_rec_state_ex'  : None,
-    'N_rec_state_in'  : None,
-    'verbose'         : False,
-    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
-    'delay'           : 1.5, # Synaptic delay in ms
-    # Neuron parameters
-    't_ref'           : 2.0, # Duration of refractory period in ms
-    'V_m'             : 0.0, # Membrane potential, initial condition in mV
-    'E_L'             : 0.0, # Leak reversal potential in mV
-    'V_reset'         : 0.0, # Reset potential of the membrane in mV
-    'tau_m'           : 20.0, # Membrane timeconstant in ms
-    'C_m'             : 1.0, # Capacity of the membrane in pF
-    'V_th'            : 20.0, # Spike threshold in mV
-    'tau_syn_ex'      : 4., # Time constants of the excitatory synaptic exponential function in ms
-    'tau_syn_in'      : .1, # Time constants of the inhibitory synaptic exponential function in ms
-    # Connection parameters
-    'eta'             : .7, # external poisson rate in Hz
-    'J'               : .5, # mV
-    'g'               : 3.9,
-    'eps'            : 0.1, # connection prob
-    'J_high'         : 2.05, # max connection strength
-    'J_low'          : 0.05,
-    'p_var'          : .5, # std of lognormal dist
-    # Stimulation parameters
-    'stim_nodes_ex'   : tuple(np.arange(1,602,1)),
-    'stim_nodes_in'   : (),
-    'stim_dist'       : 'poisson',
-    'stim_amp_ex'     : 10.0, # pA
-    'stim_amp_in'     : 0.0, # pA
-    'stim_period'     : 30.0, # ms
-    'stim_max_period' : 100, # only applies to poisson
-    'stim_duration'   : 2.0, # ms
-    'stim_N'          : 30000,
-    # analysis params
-    'latency': 6,
-    'winsize': 4
-}

### 2018-09-16 params_SI.py

@@ -1,48 +1,0 @@
-from quantities import s, ms, mV, pF, Hz, nS, pA
-import numpy as np
-
-parameters = {
-    'msd'             : 1234, # Master seed
-    'num_threads'     : 4,
-    'N_neurons'       : 1250,
-    'N_ex'            : 1000,
-    'N_in'            : 250,
-    'N_rec_spike_ex'  : None,
-    'N_rec_spike_in'  : None,
-    'N_rec_state_ex'  : None,
-    'N_rec_state_in'  : None,
-    'verbose'         : False,
-    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
-    'delay'           : 1.5, # Synaptic delay in ms
-    # Neuron parameters
-    't_ref'           : 2.0, # Duration of refractory period in ms
-    'V_m'             : 0.0, # Membrane potential, initial condition in mV
-    'E_L'             : 0.0, # Leak reversal potential in mV
-    'V_reset'         : 0.0, # Reset potential of the membrane in mV
-    'tau_m'           : 20.0, # Membrane timeconstant in ms
-    'C_m'             : 1.0, # Capacity of the membrane in pF
-    'V_th'            : 20.0, # Spike threshold in mV
-    'tau_syn_ex'      : .1, # Time constants of the excitatory synaptic exponential function in ms
-    'tau_syn_in'      : .1, # Time constants of the inhibitory synaptic exponential function in ms
-    # Connection parameters
-    'eta'             : .7, # external poisson rate in Hz
-    'J'               : .5, # mV
-    'g'               : 2.5,
-    'eps'            : 0.1, # connection prob
-    'J_high'         : 2.05, # max connection strength
-    'J_low'          : 0.05,
-    'p_var'          : .5, # percentage variation of mean in lognormal dist
-    # Stimulation parameters
-    'stim_nodes_ex'   : tuple(np.arange(1,602,1)),
-    'stim_nodes_in'   : (),
-    'stim_dist'       : 'poisson',
-    'stim_amp_ex'     : 2.0, # pA
-    'stim_amp_in'     : 0.0, # pA
-    'stim_period'     : 30.0, # ms
-    'stim_max_period' : 100, # only applies to poisson
-    'stim_duration'   : 2.0, # ms
-    'stim_N'          : 30000,
-    # analysis params
-    'latency': 3,
-    'winsize': 3
-}

### 2018-09-16 params_addnoise.py

@@ -1,53 +1,0 @@
-from quantities import s, ms, mV, pF, Hz, nS, pA
-import numpy as np
-
-parameters = {
-    'msd'             : 1234, # Master seed
-    'num_threads'     : 4,
-    'N_neurons'       : 1250,
-    'N_ex'            : 1000,
-    'N_in'            : 250,
-    'N_rec_spike_ex'  : None,
-    'N_rec_spike_in'  : None,
-    'N_rec_state_ex'  : None,
-    'N_rec_state_in'  : None,
-    'verbose'         : False,
-    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
-    'delay'           : 1.5, # Synaptic delay in ms
-    # Neuron parameters
-    't_ref'           : 2.0, # Duration of refractory period in ms
-    'V_m'             : 0.0, # Membrane potential, initial condition in mV
-    'E_L'             : 0.0, # Leak reversal potential in mV
-    'V_reset'         : 0.0, # Reset potential of the membrane in mV
-    'tau_m'           : 20.0, # Membrane timeconstant in ms
-    'C_m'             : 1.0, # Capacity of the membrane in pF
-    'V_th'            : 20.0, # Spike threshold in mV
-    'tau_syn_ex'      : .1, # Time constants of the excitatory synaptic exponential function in ms
-    'tau_syn_in'      : .1, # Time constants of the inhibitory synaptic exponential function in ms
-    # Connection parameters
-    'eta'             : .7, # external poisson rate in Hz
-    'J'               : .5, # mV
-    'g'               : 2.5,
-    'eps'            : 0.1, # connection prob
-    'J_high'         : 15., # max connection strength (low is 0)
-    'p_var'          : .5, # percentage variation of mean in lognormal dist
-    # Stimulation parameters
-    'stim_nodes_ex'   : tuple(np.arange(1,402,1)),
-    'stim_nodes_in'   : (),
-    'stim_dist'       : 'poisson',
-    'stim_amp_ex'     : 10.0, # pA
-    'stim_amp_in'     : 0.0, # pA
-    'stim_period'     : 30.0, # ms
-    'stim_max_period' : 100, # only applies to poisson
-    'stim_duration'   : 2.0, # ms
-    'stim_N'          : 30000,
-    # channelnoise
-    'gauss_mean'      : 0.,
-    'gauss_std'       : 0.,
-    # pulse packets
[…]

### 2018-09-16 run.py

@@ -1,17 +1,0 @@
-import sys
-from simulator import Simulator
-import imp
-import os.path as op
-
-if len(sys.argv) != 2:
-    raise IOError('Usage: "python run.py parameters"')
-
-param_module = sys.argv[1]
-jobname = param_module.replace('.py', '')
-currdir = op.dirname(op.abspath(__file__))
-f, p, d = imp.find_module(jobname, [currdir])
-parameters = imp.load_module(jobname, f, p, d).parameters
-
-sim = Simulator(parameters, mpi=False, data_path='results',
-                fname=jobname)
-sim.simulate(save=True, progress_bar=True)

### 2018-09-16 sweep.py

@@ -1,88 +1,0 @@
-import pandas as pd
-import numpy as np
-import quantities as pq
-import os.path as op
-from tools import corrcoef, csv_append_dict
-from params_AI import parameters
-from tqdm import tqdm
-from simulator import Simulator
-
-
-parameters.update({
-    'stim_N': 9,
-    'stim_period': 500,
-    'stim_amp_ex': 0.0,
-    'stim_amp_in': 0.0,
-    'stim_dist': None,
-})
-
-simtime_s = (parameters['stim_N'] + 1) * parameters['stim_period'] / 1000
-rate_th = 1
-spike_n_th = rate_th * simtime_s
-binsize_corr = 5.
-N_corr = 100
-
-
-def sweep(a, b, fname, amin, amax, astep, bmin, bmax, bstep, **kwargs):
-    print('Sweeping ' + fname)
-    arange = np.arange(amin, amax, astep)
-    brange = np.arange(bmin, bmax, bstep)
-    pbar = tqdm(total=len(arange) * len(brange))
-    for _a in arange:
-        for _b in brange:
-            pbar.update(1)
-            par = {a: _a, b: _b}
-            if kwargs:
-                par.update(kwargs)
-            sim = Simulator(parameters, **par)
-            sim.simulate(save=False, raster=False)
-            p = sim.data['params']
-            t_stop = sim.data['params']['status']['time']
-            spiketrains = [s['times'] for s in sim.data['spiketrains']['ex']
-                           if len(s['times']) > spike_n_th]
-            if len(spiketrains) > 1:
-                cv = np.array([np.diff(t).std()/np.diff(t).mean()
-                               for t in spiketrains])
-                ccg = corrcoef(spiketrains[:N_corr], t_stop, binsize_corr)
-                m = np.triu(ccg, k=1)
-                p.update({
-                    'CC_mean': np.nanmean(m),
-                    'CC_min': np.nanmin(m),
-                    'CC_max': np.nanmax(m),
-                    'CV_mean': cv.mean(),
-                    'CV_min': cv.min(),
-                    'CV_max': cv.max()})
[…]

### 2018-09-16 trun.py

@@ -1,125 +1,0 @@
-import matplotlib
-matplotlib.use('TkAgg')
-import matplotlib.pyplot as plt
-from simulator import Simulator
-import pandas as pd
-import numpy as np
-import quantities as pq
-import neo
-from tools import corr_coef
-from matplotlib.gridspec import GridSpec, GridSpecFromSubplotSpec
-
-par = {
-    'msd'             : 1234, # Master seed
-    'num_threads'     : 4,
-    'N_neurons'       : 1250,
-    'N_ex'            : 1000,
-    'N_in'            : 250,
-    'N_rec'           : None,
-    'res'             : 0.1, # Temporal resolution for simulation Delta t in ms
-    'delay'           : 1.5, # Synaptic delay in ms
-    'eta'             : 1., # external poisson rate in Hz
-    # Neuron parameters
-    't_ref'           : 2.0, # Duration of refractory period in ms
-    'V_m'             : 0.0, # Membrane potential, initial condition in mV
-    'E_L'             : 0.0, # Leak reversal potential in mV
-    'V_reset'         : 0.0, # Reset potential of the membrane in mV
-    'tau_m'           : 20.0, # Membrane timeconstant in ms
-    'C_m'             : 1.0, # Capacity of the membrane in pF
-    'V_th'            : 20.0, # Spike threshold in mV
-    'tau_syn_ex'      : 4., # Time constants of the excitatory synaptic exponential function in ms
-    'tau_syn_in'      : 4., # Time constants of the inhibitory synaptic exponential function in ms
-    # Connection parameters
-    'J'               : .1,
-    'g'               : 4,
-    'eps'            : 0.1, # connection prob
-    'J_high'         : 6., # max connection strength (low is 0)
-    'p_var'          : .5, # percentage variation of mean in lognormal dist
-    # Stimulation parameters
-    'stim_nodes_ex'   : tuple(np.arange(1,102,1)),
-    'stim_nodes_in'   : (),
-    'stim_dist'       : 'poisson',
-    'stim_amp_ex'     : 0.0, # pA
-    'stim_amp_in'     : 0.0, # pA
-    'stim_period'     : 30.0, # ms
-    'stim_max_period' : 100,
-    'stim_duration'   : 2.0, # ms
[…]

### 2018-09-17 tools.py
combine figures with matplotlib

@@ -126,7 +126,7 @@ def despine(ax=None, left=False, right=True, top=True, bottom=False,
         left, right, top, bottom = [True] * 4
     if ax is None:
         ax = plt.gca()
-    if not isinstance(ax, list):
+    if not isinstance(ax, (list, tuple)):
         ax = [ax]
     for a in ax:
         try:
@@ -161,14 +161,15 @@ def set_style(style='article', sns_style='white', w=1, h=1):
         },
         'notebook': {
             'figure.figsize' : (16, 9),
-            'axes.labelsize' : 25,
-            'lines.linewidth': 2,
-            'xtick.labelsize': 25,
-            'ytick.labelsize': 25,
+            'axes.labelsize' : 50,
+            'lines.linewidth': 4,
+            'lines.markersize': 20,
+            'xtick.labelsize': 30,
+            'ytick.labelsize': 30,
             'axes.titlesize' : 20,
             'font.size'      : 20,
             'legend.frameon' : False,
-            'legend.fontsize': 20,
+            'legend.fontsize': 35,
             'font.family'    : 'serif',
             'text.usetex'    : True
         }
@@ -177,3 +178,14 @@ def set_style(style='article', sns_style='white', w=1, h=1):
     plt.rcParams.update(rc)
     sns.set(rc=rc, style=sns_style,
             color_codes=True)
+
+
+def add_caption(axs, start='a', hspace=1, **kw):
+    import string
+    alph = string.ascii_lowercase
+    alph = alph[alph.index(start.lower()):]
+    for ax, capt in zip(axs, alph):
+        label = ax.get_xlabel()
+        ax.set_xlabel(
+            r'\begin{center}{}\\[{}ex]({})\end{center}'.format(label, hspace, capt),
+            **kw)

### 2018-09-18 tools.py
position funcitons in appropriate modules

@@ -1,110 +1,6 @@
 import os.path as op
-import matplotlib.pyplot as plt
-import numpy as np
 import pandas as pd
-import seaborn as sns
 
-
-def fix_figure(padLegend=None, padHeight=0):
-    plt.draw() #to know size of legend
-    fig = plt.gcf()
-    ax = plt.gca()
-    dpi = fig.get_dpi()
-    if padLegend is None:
-        padLegend = ax.get_legend().get_frame().get_width() / dpi
-    else:
-        assert isinstance(padLegend, float)
-
-    widthAx, heightAx = fig.get_size_inches()
-    pos = ax.get_position()
-    padLeft   = pos.x0 * widthAx
-    padBottom = pos.y0 * heightAx
-    padTop    = (1 - pos.y0 - pos.height) * widthAx
-    padRight  = (1 - pos.x0 - pos.width) * heightAx
-
-    widthTot = widthAx + padLeft + padRight + padLegend
-    heightTot = heightAx + padTop + padBottom
-
-    # set figure size and ax position
-    fig.set_size_inches(widthTot, heightTot)
-    ax.set_position([padLeft / widthTot, padBottom / heightTot,
-                     widthAx / widthTot, heightAx / heightTot + padHeight])
-    plt.draw()
-
-def savefig(fig, fname, ext='.svg', **kwargs):
-    if ext[0] != '.':
-        ext = '.' + ext
-    fname = op.splitext(fname)[0]
-    fig.savefig(op.join('manuscript',fname + ext), **kwargs)
-
-def coef_var(spike_trains):
-    """
-    Calculate the coefficient of variation in inter spike interval (ISI)
-    distribution over several spike_trains
-
-    Parameters
-    ----------
-    spike_trains : list
-        contains spike trains where each spike train is an array of spike times
-
-    Returns
-    -------
-    out : list
-        Coefficient of variations for each spike_train, nan if len(spike_train) == 0
-
-    Examples
-    --------
-    >>> np.random.seed(12345)
-    >>> spike_trains = [np.arange(10), np.random.random((10))]
-    >>> print('{d[0]:.2f}, {d[1]:.2f}'.format(d=coeff_var(spike_trains)))
-    0.00, -9.53
-
-    """
-    cvs = []
-    for spike_train in spike_trains:
-        isi = np.diff(spike_train)
-        if len(isi) > 0:
-            cvs.append(np.std(isi) / np.mean(isi))
-        else:
-            cvs.append(np.nan)
-    return cvs
-
-
[…]

### 2018-09-18 tools_analysis.py

@@ -1,5 +1,80 @@
 import quantities as pq
 import numpy as np
+import neo
+
+
+def coef_var(spike_trains):
+    """
+    Calculate the coefficient of variation in inter spike interval (ISI)
+    distribution over several spike_trains
+
+    Parameters
+    ----------
+    spike_trains : list
+        contains spike trains where each spike train is an array of spike times
+
+    Returns
+    -------
+    out : list
+        Coefficient of variations for each spike_train, nan if len(spike_train) == 0
+
+    Examples
+    --------
+    >>> np.random.seed(12345)
+    >>> spike_trains = [np.arange(10), np.random.random((10))]
+    >>> print('{d[0]:.2f}, {d[1]:.2f}'.format(d=coeff_var(spike_trains)))
+    0.00, -9.53
+
+    """
+    cvs = []
+    for spike_train in spike_trains:
+        isi = np.diff(spike_train)
+        if len(isi) > 0:
+            cvs.append(np.std(isi) / np.mean(isi))
+        else:
+            cvs.append(np.nan)
+    return cvs
+
+
+def corrcoef(spike_trains, t_stop, binsize=10.):
+    """
+    Calculate the pairwise Pearson correlation coefficients
+
+    Parameters
+    ----------
+    spike_trains : list
+        contains spike trains where each spike train is an array of spike times
+    t_stop : float
+        stop time of spike trains
+    binsize : float
+        binsize for histograms to be correlated
+
+
+    Returns
+    -------
+    out : array
+        Pearson correlation coefficients
+    """
+    N = len(spike_trains)
+    cmat = [corr_spikes(m1, m2, t_stop, binsize)
+            for m1 in spike_trains for m2 in spike_trains]
+    return np.array(cmat).reshape((N, N))
+
+
+def corr_spikes(s1, s2, t_stop, binsize):
+    bins = np.arange(0, t_stop + binsize, binsize)
+    mat = [np.histogram(t, bins=bins)[0] for t in [s1, s2]]
+    return corr(*mat)
+
+
+def corr(a, b):
+    mat = [(m - m.mean()) / m.std() for m in [a, b]]
+    return cov(*mat)
+
+
+def cov(a, b):
+    return np.mean((a - a.mean())*(b - b.mean()))
+
 
 def poisson_continuity_correction(n, observed):
     """
@@ -258,3 +333,70 @@ def correlogram(t1, t2=None, binsize=.001, limit=.02, auto=False,
         count = count[::-1]
 
[…]

### 2018-09-18 tools_plot.py

@@ -1,0 +1,380 @@
+import matplotlib.pyplot as plt
+import numpy as np
+import pandas as pd
+import seaborn as sns
+import quantities as pq
+from tools_analysis import make_spiketrain_trials
+import neo
+import os.path as op
+
+
+def fix_figure(padLegend=None, padHeight=0):
+    plt.draw() #to know size of legend
+    fig = plt.gcf()
+    ax = plt.gca()
+    dpi = fig.get_dpi()
+    if padLegend is None:
+        padLegend = ax.get_legend().get_frame().get_width() / dpi
+    else:
+        assert isinstance(padLegend, float)
+
+    widthAx, heightAx = fig.get_size_inches()
+    pos = ax.get_position()
+    padLeft   = pos.x0 * widthAx
+    padBottom = pos.y0 * heightAx
+    padTop    = (1 - pos.y0 - pos.height) * widthAx
+    padRight  = (1 - pos.x0 - pos.width) * heightAx
+
+    widthTot = widthAx + padLeft + padRight + padLegend
+    heightTot = heightAx + padTop + padBottom
+
+    # set figure size and ax position
+    fig.set_size_inches(widthTot, heightTot)
+    ax.set_position([padLeft / widthTot, padBottom / heightTot,
+                     widthAx / widthTot, heightAx / heightTot + padHeight])
+    plt.draw()
+
+def savefig(fig, fname, ext='.svg', **kwargs):
+    if ext[0] != '.':
+        ext = '.' + ext
+    fname = op.splitext(fname)[0]
+    fig.savefig(op.join('manuscript',fname + ext), **kwargs)
+
+
+def despine(ax=None, left=False, right=True, top=True, bottom=False,
+            xticks=True, yticks=True, all_sides=False):
+    """
+    Removes axis lines
+    """
+    if all_sides:
+        left, right, top, bottom = [True] * 4
+    if ax is None:
+        ax = plt.gca()
+    if not isinstance(ax, (list, tuple)):
+        ax = [ax]
+    for a in ax:
+        try:
+            a.spines['top'].set_visible(not top)
+            a.spines['right'].set_visible(not right)
+            a.spines['left'].set_visible(not left)
+            a.spines['bottom'].set_visible(not bottom)
+        except AttributeError:
+            raise
+        except:
+            raise
+        if not xticks:
+            a.get_xaxis().tick_bottom()
+            plt.setp(a.get_xticklabels(), visible=False)
+        if not yticks:
+            a.get_yaxis().tick_left()
[…]

### 2018-09-19 method.py
updates

@@ -2,8 +2,8 @@ import numpy as np
 import matplotlib.pyplot as plt
 from matplotlib.gridspec import GridSpec, GridSpecFromSubplotSpec
 from sklearn.linear_model import LogisticRegression
-from cross_correlation import (correlogram, poisson_continuity_correction,
-                               cch_convolve)
+from tools_analysis import (correlogram, poisson_continuity_correction,
+                            cch_convolve)
 import seaborn as sns
 
 

### 2018-09-19 tools_analysis.py

@@ -335,7 +335,8 @@ def correlogram(t1, t2=None, binsize=.001, limit=.02, auto=False,
     return count, bins[1:]
 
 
-def make_spiketrain_trials(spike_train, epoch, t_start=None, t_stop=None):
+def make_spiketrain_trials(spike_train, epoch, t_start=None, t_stop=None,
+                           dim=None):
     '''
     Makes trials based on an Epoch and given temporal bound
 
@@ -347,14 +348,42 @@ def make_spiketrain_trials(spike_train, epoch, t_start=None, t_stop=None):
         time before epochs, default is 0 s
     t_stop : quantities.Quantity
         time after epochs default is duration of epoch
+    dim : str
+        if spike_train is numpy.array, the unit must be provided, e.g. "s"
 
     Returns
     -------
     out : list of neo.SpikeTrains
     '''
+
+    if isinstance(spike_train, neo.Unit):
+        sptr = []
+        dim = unit.spiketrains[0].times.dimensionality
+        unit = unit.spiketrains[0].times.units
+        for st in unit.spiketrains:
+            sptr.append(spike_train.rescale(dim).magnitude)
+        sptr = np.sort(sptr) * unit
+    elif isinstance(spike_train, neo.SpikeTrain):
+        sptr = spike_train.times
+        dim = sptr.dimensionality
+        unit = sptr.units
+    elif isinstance(spike_train, pq.Quantity):
+        assert is_quantities(spike_train, 'vector')
+        sptr = spike_train
+        dim = sptr.dimensionality
+        unit = sptr.units
+    elif isinstance(spike_train, np.array):
+        sptr = spike_train * pq.Quantity(1, unit)
+        dim = sptr.dimensionality
+        unit = sptr.units
+    else:
+        raise TypeError('Expected (neo.Unit, neo.SpikeTrain, ' +
+                        'quantities.Quantity, numpy.array), got "' +
+                        str(type(spike_train)) + '"')
+
     from neo.core import SpikeTrain
     if t_start is None:
-        t_start = 0 * pq.s
+        t_start = 0 * unit
     if t_start.ndim == 0:
         t_starts = t_start * np.ones(len(epoch.times))
     else:
@@ -368,24 +397,7 @@ def make_spiketrain_trials(spike_train, epoch, t_start=None, t_stop=None):
         t_stops = t_stop
         assert len(epoch.times) == len(t_stops), 'epoch.times and t_stops have different size'
 
[…]

### 2018-09-19 tools_plot.py

@@ -104,7 +104,7 @@ def set_style(style='article', sns_style='white', w=1, h=1):
             color_codes=True)
 
 
-def add_caption(axs, start='a', vspace=1, **kw):
+def add_caption(ax, start='a', vspace=1, **kw):
     '''
     Adds a caption e.g. (a) underneath the xlabel.
 
@@ -121,21 +121,46 @@ def add_caption(axs, start='a', vspace=1, **kw):
     import string
     alph = string.ascii_lowercase
     alph = alph[alph.index(start.lower()):]
-    for ax, capt in zip(axs, alph):
-        label = ax.get_xlabel()
-        ax.set_xlabel(
-            r'\begin{center}{}\\[{}ex]({})\end{center}'.format(label, vspace, capt),
-            **kw)
+    ax = ax if isinstance(ax, (list, tuple)) else [ax]
+    for a, capt in zip(ax, alph):
+        label = a.get_xlabel()
+        if label == '':
+            a.set_xlabel(r'({})'.format(capt), **kw)
+        else:
+            a.set_xlabel(
+                r'\begin{{center}}{}\\[{}ex]({})\end{{center}}'.format(label, vspace, capt),
+                **kw)
+
+def label_diff(x1, x2, y, text, ax, color='k', pad_txt=.03, pad_line=0.01):
+    '''
+    Make line connected to x1 and x2 with text on top
+
+    Parameters
+    ----------
+    x1 : float
+        Postion one
+    x2 : float
+        Postion two
+    y : float
+        Hight
+    text : str
+        Text on top of line
+    '''
+    h = y * pad_line
+    ax.plot([x1, x1, x2, x2], [y, y + h, y + h, y], lw=1.5, color=color)
+    ax.text((x1 + x2) * .5, y + y * pad_txt, text, ha='center', va='bottom',
+            color=color)
 
 
 def plot_psth(spike_train=None, epoch=None, trials=None, xlim=[None, None],
               fig=None, axs=None, legend_loc=1, color='b',
               title='', stim_alpha=.2, stim_color=None,
               stim_label='Stim on', stim_style='patch', stim_offset=0*pq.s,
-              rast_ylabel='Trials', rast_size=10.,
-              hist_color=None, hist_edgecolor=None, hist_ylim=None,
-              hist_ylabel=None, hist_output='counts', hist_binsize=None,
-              hist_nbins=100, hist_alpha=1.):
+              rast_ylabel='Trials', rast_size=10,
+              hist_color=None, hist_edgecolor=None,
[…]

### 2018-10-23 manuscript/main.tex
version 1.0

@@
 \usepackage{natbib}
+\usepackage{authblk}
+\usepackage{url}
+
 \usetikzlibrary{arrows,shapes,calc}
+
 \newcommand{\abs}[1]{\left|{#1}\right|}
@@
 \newcommand{\mean}[1]{\left<{#1}\right>}
+\newcommand{\cor}[2]{\mathrm{cor}\left(#1, #2\right)}
+
 \crefname{appendix}{Appendix}{appendices}
@@
 \Crefname{subfigure}{Figure}{figures}
+
 \captionsetup[subfigure]{labelformat=simple}
@@
 %\title{The need to combine optogenetics with neural recordings to estimate causal connectivity}
-\author{Mikkel Elle Lepper\o{}d, Tristan St$\ddot{\text{o}}$ber, Konrad Paul Kording}
-
+\author[1,2]{Mikkel Elle Lepper\o{}d}
+\author[2,3,4]{Tristan St$\ddot{\text{o}}$ber}
+\author[1,2]{Torkel Hafting}
+\author[2,5]{Marianne Fyhn}
+\author[6]{Konrad Paul Kording}
+\affil[1]{Institute of Basic Medical Sciences, University of Oslo, Oslo, Norway}
+\affil[2]{Centre for Integrative Neuroplasticity, University of Oslo, Oslo, Norway}
+\affil[3]{Department of Informatics, University of Oslo, Oslo, Norway}
+\affil[4]{Department of Computational Physiology, Simula Research Laboratory, Lysaker, Norway}
+\affil[5]{Department of Bioscience, University of Oslo, Oslo, Norway}
+\affil[6]{Department of Neuroscience, University of Pennsylvania, Pennsylvania, USA}
 \begin{document}
@@
  \maketitle
+% suggestions for referee?
+\begin{abstract} % max words 150
+\noindent To study how the brain works, it is essential to identify causal interactions between neurons, usually requiring perturbations.
+However, when using optogenetics we typically perturb multiple cells, producing a confound - any of the stimulated neurons can have affected the target neuron.
+Here we show how this produces large biases, and how they can be reduced using the instrumental variable (IV) framework from econometrics.
+The interaction between stimulation and the absolute refractory period produces a weak, approximately random signal which can be exploited to estimate causality.
+When simulating integrate-and-fire neurons, we find that IV estimates connectivity better than na\"ive techniques ($R^2=0.845$ vs $R^2=0.048$).
+The difference is important as the estimates differ when applied to experimental data from stimulated neurons with recorded spiking activity.
+Our IV approach is easy to implement in the context of current experimental paradigms.
 
-\begin{abstract}
-\noindent Neuroscientists often want to identify how neurons causally interact.
-Perturbations with optogenetics are a popular way of quantifying such interactions.
-However, even focused light will typically stimulate multiple cells, which produces a confound - we cannot know which of the stimulated neurons affected the activity of a given target neuron.
-Here we show how this confound produces large biases in estimates of connectivity, and how we can reduce these biases by a data analysis strategy that exploits the absolute refractory period of neurons.
-We use the instrumental variable (IV) framework from economics which estimates causal strengths by exploiting a variable that affects only one of the variables, here the presynaptic neuron.
-The interaction between optogenetic stimulation and the absolute refractory period produces a weak almost random signal (IV) that we exploit to solve the causal inference problem.
-In a simple simulation of integrate and fire neurons, we find that instrumental variable techniques estimate connectivity better than naive techniques (99.8\% vs 0.6\% correct).
-Our instrumental variable approach is robust and easy to implement in the context of current experimental paradigms.
 \end{abstract}
@@
 \section{Introduction}
-
-\todo[inline]{Cite Konrad's new paper}
-% neuroscience experiments are observational which implicates a certain set of statistical tools
-% spike timing cannot give a causal estimate of connectivity, thus we need to perturbe the system
-%a quest for causality and optogenetics is a main tool
-% We want mechanism. Seriously. Neuroscience is to a large part experimentally driven.
-%set up the drive to causality
-We want to understand the mechanisms or causal chains that give rise to activity in the brain, to perception, action, and cognition.
-For such an understanding it is not sufficient to know the correlations between variables or even be able to predict them.
-After all, there can be many ways how the same activities come about by distinct causal chains \citep{drton2011global, peters2017elements}.
-Identifiability of causal networks is complicated when it comes to brain data.
+% The brain is infinitely confounded, thus correlations are not sufficient to draw conclusions about mechanisms and this is a hard problem in brain data.
+A goal of neuroscience is to understand the mechanisms or causal chains that give rise to activity in the brain, to perception, cognition, and action. 
 Complex systems such as the brain are hard to understand because of the numerous ways the contributing elements may interact internally \citep{Jonas2017}. 
-Thus, brain data which is almost always observational, will almost never satisfy the criteria needed for identifiability \citep{pearl2009causality}.
-While observing correlations within the system is usually relatively easy, transitioning from observed correlations to a causal or mechanistic understanding is hard or, more commonly, impossible.
-Getting at such an understanding in the human brain is nearly impossible as it contains approximately 86 billion neurons \citep{azevedo2009equal}, each of which influences many other neurons.
-Only under certain assumptions about nonlinearity or noise sources does a fully observed system become identifiable \citep{daniusis2012inferring,shimizu2006linear}.
+Because of this, it is not sufficient to know the correlations between variables or even be able to predict them.
+While observing correlations within the system is relatively easy, transitioning from observed correlations to a causal or mechanistic understanding is hard.
+After all, there can be many ways that the same activities emerge from distinct causal chains \citep{drton2011global, peters2017elements}. 
+Reaching a mechanistic level of understanding in the human brain is nearly impossible as it contains approximately 86 billion neurons \citep{azevedo2009equal}, each of which influences many other neurons.
 Even if we could record all neurons at the same time, estimating causality and producing a mechanistic understanding would be extremely challenging.
 
-% we use observational studies. But causality is typically wrong. 
-Moreover, we generally only record from a small subset of all neurons.
-The data we obtain from typical recordings, e.g.
-from electrophysiology  or calcium imaging, is very low dimensional relative to the dimensionality of the brain.
-Moreover, it is observational, which means that it does not result from randomized perturbations.
-In such cases, we can never know to which level the observed activity was caused by other observed activity, or by the activity of the unobserved neurons.
-Such unobserved activity is then called confounders.
-If mechanisms are estimated from observational data in the presence of confounders, we will generally make large errors and draw incorrect conclusions \citep{angrist2008mostly}.
+% We use observational studies, where identifiability is questionable and causality is thus typically wrong. 
+Typically, we only record from a small subset of all neurons. 
+The data we obtain from such recordings, e.g. from electrophysiology or calcium imaging, is observational, which means that it does not result from randomized perturbations. 
+In such cases, we can never know to which level the observed activity was caused by other observed activity, or by unobserved activity. 
+The unobserved neurons are called confounders.
+If mechanisms are estimated from observational data in the presence of confounders, the consequence may be large errors and incorrect conclusions \citep{angrist2008mostly}. 
 Unobserved neural activity confounds estimates of causal interactions and makes it difficult to estimate underlying mechanisms.
 
-%set up confounding
+% Set up confounding
 Confounding is the big threat to causal validity \citep{UpcomingMehlerPaper} irrespective of the use of simple regression techniques or advanced functional connectivity techniques \citep{stevenson2008inferring, honey2009predicting, aitchison2017or, pfau2013robust}.
-A much used method for estimating the output of single neurons is to perform multiple regression analyses \citep{pillow2008spatio}, modeling each neuron with a generalized linear model (GLM).
-Multiple regression may be perceived as a solution to confounding problems as they support the concept of ``explaining away" \citep{stevenson2008inferring}.
-However, in current work we will exclude approaches that look at multiple neurons at the same time.
-This is because, ``explaining away" can only be a good strategy if most neurons are included in the recordings but this is rarely the case for most experimental settings, especially in the mammalian brain.
-We will therefore focus on neuron pairs where stimulus and the pre- and post-synaptic neurons are known.
-
-To estimate connectivity it is first and foremost important that it stems from cause and effect e.g.
-B's action potential is causing an increase in C's membrane potential, therefore we use the term causal connectivity.
-For a given example we want to estimate causal connectivity between two neurons, $ A $ and $ C $ (\cref{fig:intro}\labelcref{fig:intro:1}).
-Two neurons, $ A $ and $ B $, are driven by a common input $ D $ and because $ B $ and $ C $ are connected they are strongly correlated.
-Consequently, $ A $ and $ C $ are also correlated and a regression  $ C = \beta A $ will misleadingly conclude that there is a direct interaction if it was causally interpreted.
-In this case we say the regressor $ A $ is endogenous and the regression coefficient $ \beta $ estimates the magnitude of association rather than the magnitude of causation.
+To estimate connectivity it is first and foremost important that it stems from cause and effect, therefore we use the term causal connectivity. 
+Consider an example where we want to estimate causal connectivity between two observed, but unconnected neurons, $ A $ and $ C $ (\cref{fig:intro}(a)). 
+A third unobserved neuron $ B $, is driven together with $ A $ by a common input and they are thus strongly correlated.
+Furthermore, $ B $ and $ C $ are connected, thus the input drives $ C $ through $ B $.
+Consequently, $ A $ and $ C $ are also correlated and the regression $ C = \beta A $ will misleadingly conclude a direct interaction when causally interpreted. 
+In this case, we say the regressor $ A $ is endogenous and the regression coefficient $ \beta $ estimates the magnitude of association rather than the magnitude of causation. 
 Na\"ive regressions in partially observed systems will generally not reveal causality.
 
+% Why do we not use GLM
+A much used method for estimating the output of single neurons is to perform multiple regression analyses \citep{pillow2008spatio}, modeling each neuron with a generalized linear model (GLM). 
+Multiple regression may be a solution to confounding problems as they support ``explaining away" background activity \citep{stevenson2008inferring}.
+However, this is only a good strategy if most neurons are included in the recordings.
+Furthermore, only under certain assumptions about nonlinearity or noise sources does a fully observed system become identifiable \citep{daniusis2012inferring,shimizu2006linear}.
+This is rarely the case in experimental settings, especially in the mammalian brain.
+Thus, brain data, will almost never satisfy the criteria needed for identifiability \citep{pearl2009causality}. 
+We thus aim to develop methods that can identify causal connectivity between neuron pairs.
+
+% Stimulation is an intervention which is needed to estimate causal interactions
 To estimate causal relationships between neurons, stimulating the pre-synaptic neuron is the gold standard. 
-In fact, a common definition of causality is in terms of what would happen if one would change the value of one variable in the system, independently of changing other variables -- an intervention \citep{pearl2009causality}.
+In fact, a common definition of causality is in terms of the effect of changing one variable in the system, independently of changing other variables -- an intervention \citep{pearl2009causality}. 
 If we stimulate single neurons, the ability to estimate causal relationships by regression is within reach. 
-However, this is experimentally challenging and yield low cell count because it either requires intracellular, juxtacellular or two-photon stimulation \citep{pinault1996novel, lerman2017two, nikolenko2007two, emiliani2015all}.
-Because gold-standard perturbations are challenging, it would be highly desirable if causality could be obtained from optogenetic stimulation  \citep{boyden2005millisecond, zemelman2002selective}.
+However, this is experimentally challenging and yield low cell count because it requires intracellular, juxtacellular or two-photon stimulation \citep{pinault1996novel, lerman2017two, nikolenko2007two, emiliani2015all}.
+Because gold-standard perturbations are challenging, it is necessary and highly desirable if causality could be obtained from optogenetic stimulation in combination with neural recordings of large populations of neurons \citep{boyden2005millisecond, zemelman2002selective}.
 
-%optogenetics is totally nonlocal
-Interpreting the results of optogenetic stimulation in terms of causal interactions is difficult.
-The expression of opsins are often restricted to a population of neurons, but regular optogenetic stimulation will still affect many neurons simultaneously.
-In \textit{in vivo} settings it is generally impossible to direct photons to only one neuron.
+% Optogenetic stimulation is usually totally non-local, producing confounders
+Interpreting the results from optogenetic stimulation in terms of causal interactions is difficult. 
+In most experimental settings, optogenetic stimulation will affect many neurons simultaneously. 
 Hence, the stimulus will produce a distributed pattern of activity.
-For example, if we focus a stimulation beam on one neuron, there will be a cone of light in front of and behind the neuron, which comes from the rays of light from the lens focused on the cell and again coming out on the backside.
 This distributed pattern of stimulation produces activity which then percolates through the network of neurons. 
@@
 
-%causal inference is a thing, if not done well you say stupid things
+% Causal inference is a thing, if not done well you say stupid things
+For insights into how we may resolve the confounding problem induced by optogenetic stimulation, we may look to other fields that have addressed the problem of endogenous regressors. 
 The inference of causality from observational data is addressed in the fields of statistics \citep{pearl2009causality}, machine learning \citep{peters2017elements} and econometrics \citep{angrist2008mostly}.
-Within these fields, the problem of endogenous regressors is commonly addressed.
-We may thus look towards these fields for insights into how we may resolve the confounding problem induced by optogenetic stimulation.
 
-%example
+% Example
 A commonly used approach towards causal inference in economics are instrumental variables. 
@@
 Here $ u $ are the factors other than education that contribute to yearly wages. 
-One of the factors in $ u $ is a person's cognitive ability.
-However, a person's cognitive ability may also affect education and thus the regressor $ x $  is correlated with the error term $ u $.
+One of the factors in $ u $ is a person's cognitive ability which may also affect education and thus the regressor $ x $  is correlated with the error term $ u $. 
 This will imply that the regression estimate $ \beta $ will not estimate the magnitude of causation from education on wages, but rather, its association. 
 In this case one may use the proximity to a college or university as an instrumental variable (IV) \citep{card1993using}. 
-Proximity is a good instrument since it can only affect wages through schooling, since we expect living in proximity of colleges and universities give higher probability to attend but not to affect other contributing factors to wages such as cognitive ability.
+Proximity is a good instrument since it can only affect wages through schooling. 
+We expect living in proximity of colleges and universities to give higher probability to attend without affecting other contributing factors to wages such as cognitive ability. 
 Then, in order to attribute the causal effect of education on wages one may calculate the ratio of covariances $ \beta = cov(\mathrm{proximity},\mathrm{wages})/cov(\mathrm{proximity},\mathrm{education})$.
@@
 
-%worry is this really correct? Yes, its provable. IVs are (under the right assumptions) actually correct.
+% Worry is this really correct? Yes, its provable. IVs are (under the right assumptions) actually correct.
 The IV technique has been used extensively in econometrics and is provable causal given three assumptions \citep{angrist2008mostly}. 
-First, the instrument must be uncorrelated with the error term.
+First, the instrument must be decorrelated with the error term. 
 Second, the instrument must be correlated with the regressor. 
-Third, there must be no direct influence of the instrument on the outcome variable but only an influence through the regressor variable.
-The validity of these assumptions is central when using the IV.
+Third, there must be no direct influence of the instrument on the outcome variable, but only an influence through the regressor variable. 
+The validity of these assumptions is central when using the IV approach.
 
-% Refractory period as IV
+% Refractory period as an IV
 For an instrument to be good, it needs to be unaffected by other variables. 
@@
 However, certain variables can be more or less affected. 
-For example, the overall activity of the network is through slow and strongly nonrandom dynamics.
+For example, the overall activity of the network is due to slow and strongly nonrandom dynamics. 
 In contrast, the temporal pattern of when a neuron is in a refractory state may be random. 
-First, if neurons are spiking according to conditional Poisson distributions, their exact timing will, conditioned on the network state, be random.
-While refractoriness may not be perfectly random, the exact times of spiking are notoriously difficult to predict \citep{stevenson2008inferring} suggesting that refractoriness is likely quite random.
+First, if neurons are spiking according to conditional Poisson distributions, their exact timing conditioned on the network state, will be random. 
+While refractoriness may not be perfectly random, the exact times of spiking are notoriously difficult to predict \citep{stevenson2008inferring} suggesting that refractoriness is sufficiently random.
 
-%here we
-Here we show that the instrumental variable (IV) technique can be employed if one seeks to estimate the causal connectivity between neuron pairs.
-We begin by showing how confounding factors are introduced by regular optogenetic stimulations.
+% Here we
+Here we show that the IV technique can be employed if one seeks to estimate the causal connectivity between neuron pairs. 
+We begin by showing how confounding factors are introduced by conventional optogenetic stimulations. 
 We then simulate this confounding effect in a simple network of three leaky integrate and fire (LIF) neurons. 
 With this simple model we show that by using the refractory period as an IV we are able to distinguish between connected and unconnected neuron pairs. 
-We compare these estimates with a naive although widely used cross-correlation histogram (CCH) method that fails to distinguish respective pairs.
+We compare these estimates with a na\"ive, although widely used, cross-correlation histogram (CCH) method that fails to distinguish respective pairs. 
 We then turn to a simulated network of randomly recurrent connections of excitatory and inhibitory LIF neurons with distributed synaptic weights. 
-With this data at hand we first calculate the mean squared errors of the IV method and show that it is robust to different simulated network states.
-Finally, we compare the amount and size of false positive and false negative estimates and goodness of fit on synaptic weights with pairwise assessments using CCH and logistic regression.
+With this data at hand, we first calculate the mean squared errors of the IV method and show that it is robust to different simulated network states. 
+We also compare the amount and size of false positive and false negative estimates and goodness of fit on synaptic weights with pairwise assessments using CCH and logistic regression.
+Finally, we tested the methods on experimental data from extracellular recordings of single unit activity from optogenetically perturbed ensembles of neurons in the hippocampus.
+The observed differences between the IV and the CCH estimate, underline the importance of considering potential confounding, when estimating connections based on neural activity.
 
@@
 \subsection{Optogenetics is not local}
-Optogenetics is generally seen as a perturbation method that by and large affects neurons in close proximity of the light source.
-However, it can be questioned if this is the correct way of conceptualizing the spatial effect of stimulation.
-The stimulation effects depend on multiple factors.
-Firstly, light intensity and opsin density are important as more light and ion channels will cause a stronger effect on each cell.
-Secondly, the number of potentially stimulated neurons is critical as more neurons will have a larger impact on the overall population activity.
-Lastly, physiological properties of the cells are important as light may e.g.
-only have a strong effect on spiking activity when the membrane potential of the cell is sufficiently close to the firing threshold.
-The induced effect of optogenetic stimulation as a function of distance should be given by a relation of the four parameters: light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and opsin distribution across neurons.
+Optogenetic stimulation is generally seen as a perturbation that by and large affects neurons in proximity (< 1mm) of the light source with the effect decreasing with distance. 
+However, this may be an incorrect way of conceptualizing the spatial effect of stimulation as it depends on multiple factors. 
+Light intensity and opsin density are important as more light and ion channels will cause a stronger effect on each cell. 
+Moreover, the number of potentially stimulated neurons is critical as more neurons will have a larger impact on the overall population activity. 
+Finally, physiological properties of the cells are important as light may have a stronger effect on spiking activity when the membrane potential of the cell is sufficiently close to the firing threshold.
+The induced effect of optogenetic stimulation as a function of distance should be given by a relation between the four parameters: light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and the distribution of induced photo current.
 
-To estimate the light intensity we calculated the spatial extent of laser light delivered by fiber-optics under plausible experimental conditions according to \citet{Aravanis2007}; see \cref{sec:method:opto}.
+To estimate the light intensity, we calculated the spatial extent of laser light delivered by fiber-optics under plausible experimental conditions according to \citet{Aravanis2007}; see \cref{sec:method:opto}.
 This modeling of light intensity yield an approximately $ 1/r^2 $ reduction with distance $ r $ from the stimulation site \cref{fig:concept} (cyan line). 
-This is explained by the surface of the 3D shell growing with $4\pi r^2$ and photons will be roughly isotropic (but see \citep{thunemann2018does}) beyond the scattering length \cref{fig:concept} (inset).
+This is explained by the surface of a 3D shell growing with $4\pi r^2$ and photons will be roughly isotropic (but see \citep{thunemann2018does}) beyond the scattering length \cref{fig:concept} (inset).
 The same number of photons has to cross each of the spheres around the stimulation location unless they are absorbed. 
-The density of photons thus decreases rapidly with distance.
+As a result, the density of photons decreases rapidly with distance.
 
-The number of illuminated neurons at a given distance will be larger the further away they are from the stimulation site given that neurons are uniformly distributed in brain tissue \cref{fig:concept} (black line).
+The number of illuminated neurons at a given distance will increase with distance to the stimulation site given that neurons are uniformly distributed in brain tissue \cref{fig:concept} (black line). 
 In fact, it will increase by approximately $ r^2 $ with distance. 
@@
 Surprisingly, this distribution has been observed to be symmetrically distributed and relatively flat \citep{pare1998impact,destexhe1999impact,rudolph2006use}. 
-The response that is expected due to a pulse of light that induces a charge $Q$ should be proportional to the density of neurons whose membrane potential sit within a $Q/C$ range of the threshold ($C$ is the capacitance).
-The fact that the distribution of membrane potentials is relatively flat (the density close to the threshold is generally within an order of magnitude of the density of its mode) suggests that the spiking response to a perturbation for any neuron is roughly proportional to the induced photo current.
-The peak amplitude of the photo current relates non-linearly to the light intensity and can be modelled by the Hill equation \citep{wang2007high}.
-Moreover, assuming that opsins are evenly distributed along the neuron, the opsin density in the neuron population will be relatively even.
-Based on this, we calculate the overall stimulation effect to be the product of the number of neurons in a spherical slice and the peak amplitude photo current which actually increases with distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line).
+The expected response from a pulse of light that induces a charge $Q$ should be proportional to the number of affected neurons whose membrane potential sit within a $Q/C$ range of the threshold ($C$ is the capacitance).
+Given that the distribution of membrane potentials is relatively flat (the density close to the threshold is generally within an order of magnitude of the density of its mode) suggests that the spiking response to a perturbation for any neuron is roughly proportional to the induced photo current.
+The peak amplitude of the photo current relates approximately logarithmically to the light intensity \citep{wang2007high}; see \cref{sec:method:opto}. 
+Assuming that opsins are evenly distributed across neurons, the induced photo current will not be proportional with light intensity - it will fall slower. 
+Based on this, we calculate the overall stimulation effect to be the product of the number of neurons in a spherical slice and the peak amplitude photo current. 
+This product actually increases with distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line). 
 Thus, optogenetic stimulation utilizing single photon activation does not produce a localized effect.
@@
 Due to scattering and geometric loss the light intensity (I, cyan line) with an intensity of $10 mW/mm^2$ exiting the optogenetic fiber follows approximately an inverse square law $ r^{-2} $ where $ r $ is the distance from the fiber.
-If neurons are uniformly distributed the number of affected neurons increase by $ r^{2} $ (N, black line).
-The total photo current (sum of single neuron currents) thus increases by distance due to the nonlinear relation between photo intensity and photo current (P, blue line).
-\label{fig:concept}}
+If neurons are uniformly distributed, the number of affected neurons in a spherical slice increases by $ r^{2} $ (N, black line). 
+The total photo current (sum over neurons of peak amplitude photo current in a spherical slice) thus increases with distance due to the nonlinear relation between light intensity and photo current (P, blue line) depicted as percentage of maximum.}
+\label{fig:concept}
 \end{figure}
+
 \FloatBarrier
@@
 When we stimulate many neurons at the same time, and observe a post-synaptic neuron to be active after our stimulation, it is hard to know which of the stimulated neurons produced the activity. 
-To illustrate such confounding effects we simulated a network comprised of three neurons (A,B, and C) \cref{fig:intro}\labelcref{fig:intro:1}.
-The neurons receive Poisson spike trains and have added Gaussian white noise to the membrane potential.
-We also gave them interactions, where spikes of neuron B increase the probability of firing for neuron C but there are no other interactions.
-Finally, we allowed simulated optogenetic stimulation to affect neurons A and B (but not C). We thus have a simple system for exploring questions of causality.
+To illustrate such confounding effects we simulated a network comprised of three neurons (A,B, and C) \cref{fig:intro}(a). 
+The neurons receive Poisson spike trains and have Gaussian white noise added to the membrane potential. 
+Neurons were also interacting, where spikes of neuron B increase the probability of firing for neuron C, but there were no other interactions.
+Finally, we allowed simulated optogenetic stimulation (current pulse) to affect neurons A and B (but not C). 
+We thus have a simple system for exploring questions of causality.
 
-After running the simulation, the peri-stimulus time histogram of the stimulated neurons show the result of both the stimulation itself (suppressed for visibility) and the neuron's refractory period \cref{fig:intro}\labelcref{fig:intro:2} (AA, BB).
-Since the stimulation affects A and B simultaneously, it induces a strong correlation between A and B \cref{fig:intro}\labelcref{fig:intro:3} (AB).
-This further generates a strong correlation between A and C, confounding the system by rendering the cross correlation histograms (CCHs) between BC and AC both statistically significant ($ p_{\mathrm{fast}} < 0.001, p_{\mathrm{diff}} < 0.001 $; see \cref{sec:method:cch}) shown in \cref{fig:intro}\labelcref{fig:intro:3}.
-Even though the correlation peak between B and C is larger than between A and C due to correlated spikes outside the periods with stimulation one may imagine a situation where only A and C is measured, giving rise to a false prediction that they are connected.
+After running the simulation, the peri-stimulus time histogram of the stimulated neurons (\cref{fig:intro}(b)) shows the result of both the stimulation itself (suppressed for visibility) and the neuron's refractory period \cref{fig:intro}(c) (AA, BB).
+Since the stimulation affects A and B simultaneously, it induces a strong correlation between A and B \cref{fig:intro}(c) (AB). 
+This further generates a strong correlation between A and C, confounding the system by rendering the cross-correlation histograms (CCHs) between BC and AC both statistically significant ($ p_{\mathrm{fast}} < 0.001, p_{\mathrm{diff}} < 0.001 $; see \cref{sec:method:cch}).
+Even though the correlation peak between B and C is larger than between A and C, due to correlated spikes outside the periods with stimulation, one may imagine a situation where only A and C is measured, giving rise to a false prediction that they are connected.
 If stimulation affects multiple neurons simultaneously, there is a real confounding problem.
@@
 \end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{raster}
+\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{psth_triple}
 \caption{} \label{fig:intro:2}
-\end{subfigure}
-
-\medskip
+\end{subfigure}\medskip\\
 \begin{subfigure}{\textwidth}\centering \includegraphics[scale=1]{xcorr}
@@
 \end{subfigure}
-\caption{{\bf A,B,C network}. A sketch of the simple network containing three neurons %depicted in \labelcref{fig:intro:3} 
-shows stimulation configuration with blue laser light and the connections with arrows (a).
-The neurons A and B are stimulated 1000 trials and the corresponding peristimulus time histogram are shown in \labelcref{fig:intro:2} upper panel with a raster plot in the lower panel.
-Cross correlation histograms (CCHs) are shown in \labelcref{fig:intro:3} where horizontal and vertical axes represents time lag in ms and counts of coincident spikes in bins of 1 ms.
-\label{fig:intro}}
+\caption{{\bf Optogenetic stimulation induces spurious correlations}. 
+A sketch of the simple network containing three neurons shows stimulation configuration with blue laser light and the connections with arrows (a). 
+The neurons A and B are stimulated in 1000 trials and the corresponding peristimulus time-histogram are shown in (b) upper panel with a raster plot in the lower panel. 
+Cross-correlation histograms (CCHs) are shown in (c) where horizontal and vertical axes represent time lag in ms and counts of coincident spikes in bins of 1 ms.}
+\label{fig:intro}
 \end{figure}
+
 \FloatBarrier
 \subsection{Instrumental variables to resolve confounding}
-In order to estimate the actual influence of stimulation of a neuron on post-synaptic neurons we need something that can distinguish the influence of one stimulated neuron from the influence of another stimulated neuron.
+In order to estimate the actual influence of stimulation of a neuron on post-synaptic neurons, we need to distinguish the influence of one stimulated neuron from the influence of another stimulated neuron.
 We would thus need something that affects the stimulation effect separately between neurons. 
@@
 If a neuron is in its absolute refractory period, then no amount of stimulation will make it spike. 
-This gives us an interesting way of getting at causality, by comparing the network state between a time when a neuron is able to spike and a time where the neuron is unable to spike.
+This gives us an interesting way of inferring causality, by comparing the network state between a time when a neuron is able to spike and when the neuron is unable to spike.
 
-Instrumental variables requires the existence of a random (or sufficiently random) variable that affects a variable of interest.
-This independent influence among neurons then allows quantifying the influence of the variable of interest on the rest of the network.
+Instrumental variables require the existence of a random (or sufficiently random) variable that affects the variable of interest. 
+This independent influence then allows quantifying the influence of the variable of interest on the rest of the network. 
 In our case, the refractory states of a neuron is in good approximation independent on small time scales (see Discussion for caveats). 
-It affects the influence of stimulation on the potentially pre-synaptic neuron (\cref{fig:cchvswald}\labelcref{fig:cchvswald:0}).
-The trials where a stimuli is unable to elicit a spike due to the refractory state can then be used to identify causal effect on the potentially post-synaptic neuron.
+It affects the influence of stimulation on the putative pre-synaptic neuron (\cref{fig:cch_iv_triple}(a)). 
+The trials where a stimuli is unable to elicit a spike due to the refractory state can then be used to identify causal effect on the putative post-synaptic neuron.
 
-We can now investigate if the use of an instrumental variable gives a better estimate of connectivity strength than simply analyzing the lagged correlations by means of the CCH; see \cref{eq:ptrans}.
-We thus use the IV estimator \cref{eq:wald} on the three neuron system (\cref{fig:cchvswald}\labelcref{fig:cchvswald:1,fig:cchvswald:2}).
-It converges to the correct causal conclusions that the weights $ w_{BC} = 0.2 $ and $ w_{AC} = 0 $.
+We can now investigate if the use of an instrumental variable gives a better estimate of connectivity strength than simply analyzing the lagged correlations by means of the CCH calculated with \cref{eq:ptrans} (\cref{fig:cch_iv_triple}(b, c) dashed lines).
+We use the IV estimator given by \cref{eq:wald} on the three neuron system (\cref{fig:cch_iv_triple}(b, c) solid lines). 
+It converges to the correct causal conclusions that the weights $ w_{BC} = 0.2 $ and $ w_{AC} = 0 $ as opposed to the CCH method which falsely concludes that $ w_{AC} \approx 0.1 $. 
 For such a simple system, it produces meaningful estimates of the causal interactions between neurons.
@@
 \makeatother
-\begin{subfigure}{0.485\textwidth} \centering
+\begin{subfigure}{\textwidth} \centering
 \begin{tikzpicture}
@@
 \end{tikzpicture}
-\caption{} \label{fig:cchvswald:0}
-\end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \centering
-\includegraphics[scale=1]{wald_legend}
-\end{subfigure}\medskip
-
-\begin{subfigure}{0.485\textwidth} 
-\includegraphics[scale=1]{wald_BC}
-\caption{} \label{fig:cchvswald:1}
-\end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} 
-\includegraphics[scale=1]{wald_AC}
-\caption{} \label{fig:cchvswald:2}
+\caption{} \label{fig:cch_iv_triple:0}
+\end{subfigure}\medskip\\
+\begin{subfigure}{\textwidth} 
+\includegraphics[scale=1]{triple}
 \end{subfigure}
 \caption{{\bf Instrumental variable estimation (IV) of connectivity}. 
-\labelcref{fig:cchvswald:0} In an instrumental variable estimation procedure we use a variable that is assumed to be random (here refractoriness) which influences a variable of interest (here spiking) and to use this influence to infer the causal interaction of that variable on other variables (here spiking of A or B onto C).
-\labelcref{fig:cchvswald:1} A popular estimation approach for IVs,  the Wald technique correctly estimates causal connectivity in the A,B,C neural network using the refractory period.
-Subfigure \labelcref{fig:cchvswald:0} shows a path diagram \citep{wright1921correlation} between the upstream neuron $ A $ or $ B $, the stimulation $ S $, the post-synaptic neuron $ C $ and the IV as $ A_{r} $ or $ B_{r} $.
-Arrows represent associations where $ S $ is associated with $ A,B $ and potentially also with $ C $ both directly and through $ A,B $.
-The IV estimator calculated by \cref{eq:wald} converges to $ \hat{\beta}_{BC} \approx 0.2, \hat{\beta}_{AC} \approx 0 $ after approximately $ 5000 $ trials as seen in \labelcref{fig:cchvswald:1,fig:cchvswald:2} respectively.
- \label{fig:cchvswald}}
+(a) during instrumental variable estimation we use a variable that is assumed to be random (here refractoriness) which influences a variable of interest (here spiking) and to use this influence to infer the causal interaction of that variable on other variables (here spiking of A or B onto C).
+A popular estimation approach for IVs, the Wald technique, correctly estimates causal connectivity in the A, B, C system using the refractory period. 
+The path diagram in (a) shows the associations between neurons $ A $, $ B $ or $ C $, the stimulation $ S $, and the IV as $ A_{r} $ or $ B_{r} $ (r for refractory).
+The IV estimator calculated by \cref{eq:wald} converges to $ \hat{\beta}_{BC} \approx 0.2, \hat{\beta}_{AC} \approx 0 $ after approximately $ 5000 $ trials as seen in (b, c). 
+Black line indicates ground truth and the results of the cross-correlations are showed as dotted lines. 
+Note the difference between the IV estimate and CC method in (c).}
+\label{fig:cch_iv_triple}
 \end{figure}
 \FloatBarrier
+
 \subsection{Larger simulated networks}
-The interacting neurons in a biological network exhibit inhibition and interact in many kinds of ways.
-To evaluate the IV method in a more meaningful setting we thus simulated a recurrent neural network consisting of 1250 randomly connected leaky integrate and fire neurons where 250 had inhibitory synapses.
-The network was tuned to be in an asynchronous regime; see \cref{fig:state}\labelcref{fig:CC} with log-normally distributed synaptic weights according to patch clamp experiments \citep{Sayer1990,Mason1991}; see \cref{fig:state}\labelcref{fig:syn_dist} and \cref{tab:params} for parameters.
-Further, we selected 800 excitatory neurons for stimulation and gave each neuron a random spatial distance from the simulated optogenetic stimulus.
-The stimulus intensity was then set according to \cref{eq:intensity} with a maximum of 10 pA, and was constant throughout trials.
+Interacting neurons in a biological network exhibit inhibition and interact in many ways. 
+To evaluate the IV method in a more meaningful setting we simulated a recurrent neural network consisting of 1250 randomly connected LIF neurons where 250 had inhibitory synapses. 
+The network was tuned to be in an asynchronous regime; see \cref{fig:state}\labelcref{fig:CC} with log-normally distributed synaptic weights according to patch-clamp experiments \citep{Sayer1990,Mason1991}; see \cref{fig:state}\labelcref{fig:syn_dist} and \cref{tab:params} for parameters.
+Furthermore, we selected 800 excitatory neurons for stimulation and gave each neuron a random spatial distance from the simulated optogenetic stimulus. 
+The stimulus intensity was then set according to \cref{eq:hill} with a maximum of 8 pA, and was constant throughout trials. 
 The trial onset had a temporal Poisson distribution with period 100 ms and was further clipped between 100-150 ms. 
-For weight estimates we randomly selected among the excitatory population 100 stimulated neurons and 100 unstimulated neurons.
+For weight estimates, we randomly selected among the excitatory population 100 stimulated neurons and 100 nonstimulated neurons.
 
-To evaluate how well the IV method estimates the weights as a function of number of trials we calculated the mean squared error of the weight connecting the 100 neuron pairs \cref{fig:error}.
-As seen here, the IV estimator's precision decreases similarly in three different settings with varying amounts of relative inhibition $ g $.
+To evaluate how well the IV method estimates the weights as a function of number of trials we calculated the mean squared error of the connection weight \cref{fig:error}. 
+The IV estimator's precision decreases similarly in three different settings with varying amounts of relative inhibition $ g $.
 \begin{figure}
@@
 \includegraphics[scale=1]{network-raster}
-\caption{} \label{fig:error:raster}
 \end{subfigure}\medskip
@@
 \includegraphics[scale=1]{mse}
-\caption{} \label{fig:error:mse}
 \end{subfigure}
-\caption{{\bf Mean square error (MSE) of IV estimator in a large network}.
-\labelcref{fig:error:raster} Excitatory neurons (in blue raster) are stimulated with varying intensity indicated by the y-axis in lower panel, upper panel shows inhibitory neurons (red raster) where y-axis indicate random neuron-id.
-\labelcref{fig:error:mse} The IV- and CCH estimator is evaluated for the asynchronous recurrent neural network at three different amounts of relative inhibition $ g $.
-The MSE as a function of number of trials is shown on a logarithmic scale where the slopes was found to be $ -0.67, -0.69, -0.45 $ for $g = 9.9, 4.4, 3.0$ respectively.
-\label{fig:error}}
+
+\caption{{\bf Mean square error (MSE) of IV and CCH based estimators in a network of two populations.} 
+(a) Raster plots showing inhibitory neurons (upper panel, red) and excitatory neurons (lower panel, blue) are stimulated with varying intensity with the strongest at lower neuron number.
+(b) Histogram over two stimulations, all neurons where time zero indicates stimulation onset, stimulated excitatory neurons drive inhibition which then silents the entire network.
+(c, d) The IV- and CCH estimators are evaluated for the asynchronous recurrent neural network at three different amounts of relative inhibition $ g $. 
+The MSE as a function of number of trials is shown in (c) on a logarithmic scale where the slopes was found to be $ -0.53, -0.48, -0.31 $ for $g = 9.9, 4.4, 3.0$ respectively. 
+Finally, the MSE as a function of stimulation effect (hit rate) is shown in (d) where $hit rate = 1$ indicates that the stimulation induces a spike for each trial. \label{fig:error}}
 \end{figure}
 
-We want to compare the IV estimator which exploits the refractory period with the CCH method given by \cref{eq:ptrans} which ignores network confounding.
+We then compared the IV estimator which exploits the refractory period, with the CCH method given by \cref{eq:ptrans} which ignores network confounding. 
 We defined hit rate as the relative amount of stimulations in non-refractory states and required a minimum of 90$\%$. 
-To indicate the amount of connections which are falsely attributed a non-zero weight we calculated the amount of false positives.
-This was given as the percentage of estimated synapses larger than $ 0.05 $ where the true weight was $ 0 $, finding $ 99.4\% $ for CCH and $ 0.2\% $ for the IV estimator; see \cref{fig:network-class}\labelcref{fig:network-class:1}.
-In addition we compared the size of the estimations at false positive instances and found that the CCH have significantly higher median than IV (p=0, difference ($\Delta$) = 0.17, permutation resampling \citep{wassermann2006all}).
-The IV approach, while not being perfect, thus outperforms the simple CCH approach.
+To indicate the amount of connections that are falsely attributed to a non-zero weight, we calculated the amount of false positives. 
+This was given as the percentage of estimated synapses larger than $ 0.05 $ where the true weight was $ 0 $, finding $ 99.4\% $ for CCH and $ 0.2\% $ for the IV estimator; see \cref{fig:fitness}(a). 
+In addition, we compared the size of the estimations at false positive instances and found that the CCH method have significantly higher median than IV (p=0, difference ($\Delta$) = 0.164, permutation resampling \citep{wassermann2006all}).
+The IV approach, while not being perfect, thus outperforms the CCH approach.
 
-It might be that modeling refractory periods in the context of a naive regression estimates connectivity equally well as the IV method.
-We thus performed a logistic regression as seen in \cref{fig:network-class}\labelcref{fig:network-class:1} denoted LOGIT.
-As seen here, LOGIT performs even worse than CCH showing that it really helps to use the refractory period as an instrumental variable (p=0, $\Delta$ = 0.7, permutation resampling).
-To further evaluate the methods we calculated false negatives as instances where the true weight is non-zero but estimated to be equal to zero in \cref{fig:network-class}\labelcref{fig:network-class:2} showing that the CCH and IV estimators performs equally well (p=0.78, $\Delta$ = 0.021, permutation resampling).
-Finally we wanted to evaluate the estimated weights as a function of true weights shown in \cref{fig:network-class}\labelcref{fig:network-class:3,fig:network-class:4} after 30000 trials.
-The IV estimator yields a good prediction ($ R^2 = 0.845 $), while the CCH mainly estimates the strength of stimulation while true weights are poorly estimated ($ R^2 = 0.048 $).
-Utilizing refractory periods as an (imperfect) instrumental variable considerably improves estimates.
+It might be that modeling refractory periods in the context of a na\"ive regression estimates connectivity equally well as the IV method. 
+We thus performed a logistic regression \cref{fig:fitness}a denoted LOGIT. 
+Here, we show that LOGIT performs worse than CCH illustrating the advantage of using the refractory period as an instrumental variable (p=0, $\Delta$ = 0.577, permutation resampling). 
+To further evaluate the methods, we calculated false negatives as instances where the true weight is non-zero but estimated to be zero in \cref{fig:fitness}(b) shows that the CCH and IV estimators perform equally well (p=0.59, $\Delta$ = 0.017, permutation resampling), while LOGIT have no false negatives.
+Finally, we wanted to evaluate the estimated weights as a function of true weights shown in \cref{fig:fitness}(b,c) after 30000 trials. 
+The IV estimator yields a good prediction ($ R^2 = 0.770 $), while the CCH method mainly estimates the strength of stimulation while true weights are poorly estimated ($ R^2 = 0.009 $).
+Utilizing refractory periods as an instrumental variable considerably improves the estimations.
 
@@
 \makeatother
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{false_positive}
-\caption{} \label{fig:network-class:1}
-\end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{false_negative}
-\caption{} \label{fig:network-class:2}
+\begin{subfigure}{\textwidth} 
+\includegraphics[scale=1]{false_estimate}
 \end{subfigure}
@@
 \medskip
-\begin{subfigure}{0.46\textwidth} \includegraphics[scale=1]{fit_wald}
-\caption{} \label{fig:network-class:3}
-\end{subfigure}\hfill
-\begin{subfigure}{0.6\textwidth} \includegraphics[scale=1]{fit_cch}
-\caption{} \label{fig:network-class:4}
+\begin{subfigure}{\textwidth} 
+\includegraphics[scale=1]{goodness_fit}
 \end{subfigure}
 \caption{{\bf False estimates and goodness of fit}. 
-False positives are shown in \labelcref{fig:network-class:1} for the cross correlation histogram (CCH) method, logistic regression (LOGIT) and the IV estimator.
-False negatives for CCH and IV is shown in \labelcref{fig:network-class:2}.
-Positive estimates of weight as a function of true weight is scattered for the IV estimator in \labelcref{fig:network-class:3} and CCH in \labelcref{fig:network-class:4} color coded by the size of perturbation intensity.
-Shaded area show the 95 \% confidence interval calculated by bootstrapping.
-Data were obtained from simulation of model 2 \cref{tab:params}.
-\label{fig:network-class}}
+False positives are shown in (a) for the CCH method, logistic regression (LOGIT) and the IV estimator. 
+False negatives for CCH and IV are shown in (b). 
+Positive estimates of weight as a function of true weight are scattered for the IV estimator in (b) and CCH in (c), color coded by the size of perturbation intensity. 
+Shaded area shows the 95 \% confidence interval calculated by boot-strapping. 
+Data were obtained from simulation of model 1 \cref{tab:params}.
+\label{fig:fitness}}
 \end{figure}
 \FloatBarrier
-\subsection{Application to hippocampal recordings with Optotrodes}
-% % aim: test method on state of the art, publicly available data of optostimulation
-%   % compare results for transmission probability and wald estimate
+
+\subsection{IV and CCH estimates from hippocampal spike recording with concurrent optogenetic stimulation}
+To test the feasibility of the IV approach on \textit{in vivo} data, we compared the CCH and IV estimators on openly available extracellular single unit recordings from two mice with light-pulse stimulations of CA1 pyramidal neurons \citep{English2017}.
+We investigated connections from potentially pre-synaptic units, that exhibited a significant increase in their firing rate upon stimulation, with potentially post-synaptic units that did not respond to the stimulation \citep{English2017}.
+These experiments were not optimally designed for the use of the IV method to infer causal connectivity.
+Because of a low stimulation hit rate, we used an IV window of $ 7.5 $ ms, larger than the absolute refractory period of pyramidal cells, $ 4 $ ms.
+This allows a proof-of-principle evaluation of the IV method on experimental data.
+
+We want to know to what degree the IV method can be used on experimental data and if it gives different results. 
+Indeed, when we apply the IV method, \cref{fig:optodata}, it seems to work robustly. 
+Although we find that the results from IV are somewhat correlated with those from CCH (correlation coefficient of $0.24$), there are considerable differences between the methods. 
+In many cases the confidence bounds of the two methods are truly non-overlapping, suggesting that the differences cannot only be explained by noise. 
+These preliminary results illustrate that the IV approach may be applicable to address a broad range of causality estimation problems in neuroscience.
+
 \begin{figure}
@@
 \includegraphics[width=1.\textwidth]{Optotrode.jpg}
+%\vspace{3em}
 \caption{} \label{}
@@
 \begin{subfigure}{0.6\textwidth}
-\includegraphics[width=1.\textwidth]{Optodata_IVvsCCH.pdf}
+\includegraphics[width=1.\textwidth]{Optodata_comparisonIV_CCH.pdf}
 \caption{} \label{}
 \end{subfigure}
-
-\caption{}.
-\label{fig:optotrodedata}
+\caption{\textbf{IV and CCH estimates on hippocampal recording}
+We compared the IV and CCH estimates of synaptic strengths on CA1 recordings with optogenetic activation of pyramidal cells \citep{English2017}.
+(a) Left: schematic of the $\mu$-LED silicon probe, right: photograph of one shank tip with three illuminated $\mu$-LEDs with a scale bar at $15 \mu m$.
+(b) IV and CCH estimates for connections from optogenetically activated to non-activated units that showed a significant peak in the CCH \citep{English2017}.
+Dashed line represents identity. 
+Error bars are $ \pm $ one standard deviation.
+}
+\label{fig:optodata}
  \end{figure}
@@
 \section{Discussion}
-%% maybe add somewhere that using other recorded neurons may be useful as well.
+% We have..
+Here we have asked if the refractory period of neurons can be used as an instrumental variable to reverse engineer the causal flow of activity in a network of simulated neurons. 
+We have found that this approach performs considerably better than the na\"ive method. 
+We have found that neither na\"ive linear nor na\"ive logit models produce reliable estimates of connectivity between neuron pairs. 
+The IV approach effectively reverse engineers causality by looking at the response that is missing because of refractoriness which effectively allows better estimates of causal effects.
+When applied to real data, we obtain robust estimates that differ from those of na\"ive estimators.
 
-Here we have asked if the refractory period of neurons can be used as an instrumental variable to reverse engineer the causal flow of activity in a network of simulated neurons.
-We have found that this approach performs considerably better than the naive method.
-We have found that neither naive linear nor naive logit models produce good estimates of weights.
-Our system effectively reverse engineers causality by looking at the response that is missing because of refractoriness which effectively allows better estimates of causal effects.
+% No ground truth available, if they existed they should be random, short and intense
+At the moment, we have no ground-truth data set at hand to test our technique and compare with other approaches. 
+Ideally, we would have known causal effects from single-cell stimulation (e.g. from two-photon optogenetics) to establish causal effects. 
+Such data should contain many randomly distributed, short and intensive stimulation trials combined with traditional optogenetics, designed in a way where refractoriness matter. 
+Such a dataset, to the best of our knowledge, is currently not available and prevents us from testing how good our estimator would work on experimental data. 
+Future experiments are needed to obtain reliable insights.
+
+% Correlated refractory periods is *the* failure mode. How to randomize them
+For the refractory period to be a good instrument, it is necessary that it is not overly affected by the network activity. 
+This will clearly be problematic in many cases. 
+After all, network activity affects neuron activity and hence refractoriness. 
+However, there are multiple scenarios where refractoriness will be a good instrument. 
+For example, if we have balanced excitation and inhibition, we may expect largely independent refractory states of individual neurons. 
+If a neuron biophysically implements something like conditional Poisson spiking, its refractory states will be random. 
+Importantly, we may expect the phase of a neuron to be far more random than the activity of the network as a whole.
+
+% Importance of properly designed experiments 
+The randomness of refractory times is the one factor which makes or breaks the IV approach. 
+Even if neurons' refractory states are strongly correlated during normal network operation, there may be ways of randomizing refractoriness. 
+First, it would help to use a task and situation where neurons are as uncorrelated as possible. 
+Second, we may use a set of conditioning pulses of stimulation to increase independence of refractory states. 
+Giving one burst of stimulation which is strong enough to elicit multiple spikes from each neuron may effectively randomize their phases \citep{ermentrout2008reliability}.
+Third, we may utilize chemical, behavioral, or molecular perturbations to randomize refractoriness. 
+For example, we may be able to construct intracellular oscillators that are unaffected by neural activities. 
+To the best of our knowledge, there has not been any attempts in neuroscience to produce good instrumental variables, so there may be many possibilities for improvements.
 
 %Wald is pairwise, GLMs require multiple neurons
-One popular way of estimating causal effects is fitting generalized linear models (GLMs) to simultaneously recorded neurons \citep{pillow2008spatio}.
-The GLMs are basically multiple regressions and require multiple neurons in order to perform well.
-In fact, if one recorded all neurons, GLMs could be sufficient to estimate causal connections.
-However, this is not the case for research in the mammalian brain, especially not for primates, where we only record a very small subset of the neurons that do the actual computation.
-The GLM field is very strong at modeling latency distributions and sequences of spikes in individual neurons.
+One popular way of estimating causal effects is fitting generalized linear models (GLMs) to simultaneously recorded neuron activities \citep{pillow2008spatio, roudi2009ising}. 
+GLMs are multiple nonlinear regressions and require multiple neurons to perform well. 
+In fact, if activity from all neurons were recorded, GLMs might be sufficient to estimate causal connections. 
+However, complete recordings are not possible in the mammalian brain, especially not in primates, where recordings include only a very small subset of the neurons involved in the actual computation. 
+When using GLMs one may accuratly estimate latency distributions and sequences of spikes from individual neurons. 
 These ideas should, arguably, be merged with IV approaches. 
-The main strength of the  IV estimator presented here compared with GLM methods is that we only require one pair because we can utilize the randomness in refractoriness.
+The main strength of the IV estimator presented here is that it only requires one pair to be recorded because we can utilize the randomness of refractory periods.
 
-%We recommend the WALD, easily usable in the context of many experiment. Potentials for future improvement.
-%The IV estimator seems to be useful when using the refractory period.
-This method yields a direct estimation by computing the difference in mean activity of a putative post-synaptic neuron when the upstream neuron is responsive to the stimulus or not.
-Moreover, as this method only requires one pair it is easily usable in many types of experiments, with large or small recording counts.
-However, there would be many ways of improving the estimates, we could better model the temporal sequence of spikes, we could utilize latency in a more meaningful way.
-While there are many ways of improving the IV estimators, we have here represented the most basic implementation using the Wald estimator.
-
-% Maybe nonlinearities in stimulation rescue old school approaches
-% Discuss distribution of V. Discuss 2p stimulation.  Conclude that there is absolutely no reason to hope things will work with 1p.
-The main problem with optogenetics when using it to infer connectivity is its non-local property.
-This is due to the inverse relation between changes in light intensity and affected number of neurons combined with the observation that distribution of membrane potentials across neurons is flat \citep{destexhe1999impact,rudolph2006use,pare1998impact}.
-One could however imagine situations where optogenetic activation were more local.
-If membrane potential distributions were in general skewed with the mode far from threshold the optogenetic perturbation would be more local.
-This is because each neuron would require a very strong stimulus in order to  elicit spikes at all.
-But there could be other ways of making optogenetics more local.
-For example, if one could engineer opsins with far more absorbent wavelengths one could stimulate locally.
-How to engineer more localized stimulation is an important problem if one wants to causally interrogate systems.
+The main problem with optogenetic stimulation, when used to infer connectivity, is its non-local property. 
+This is due to the inverse relation between changes in light intensity and affected number of neurons combined with a logarithmic relation between light intensity and photocurrent \cite{wang2007high}.
+In addition, the distribution of membrane potentials across neurons is flat \citep{destexhe1999impact,rudolph2006use,pare1998impact} making neurons highly sensitive to perturbations. 
+One could however, imagine situations where optogenetic activation was more local. 
+If for example, the membrane potential distributions were skewed with the mode far from threshold, a very strong stimulus would be required for a neuron to elicit spikes. 
+There could also be other ways of making optogenetic stimulation more local. 
+For example, if one engineered opsins or brain tissue that are more light absorbent one could stimulate more locally. 
+How to engineer more localized stimulation is an important problem when causally interrogating a system.
 
 %Maybe weak stimulation will rescue us. No way, but cite Buszaki.
-Very weak laser pulses in noisy networks might only affect very few neurons each trial \citep{English2017}.
-However, the stimulus will still affect the many far away neurons by a tiny bit.
-Therefore, weak stimulation does not remove the principal problem of the CCH estimator.
-Moreover, the network still acts as a confounder and, if anything, the weak stimulation will dramatically reduce the statistical power of the approach.
+Very weak laser pulses in noisy networks might mainly elicit spikes in very few close-by neurons in each trial \citep{English2017}. 
+However, the stimulus will still affect the membrane potential of many neurons further away, some of which will spike. 
+Therefore, weak stimulation does not remove the principal problem of correlation based techniques. 
+After all, the network still acts as a confounder and, if anything, the weak stimulation will reduce the statistical power of the approach. 
 Lowering stimulation amplitudes does not appear to be a way of obtaining meaningful causal estimates.
 
-%Correlated refractory periods is *the* failure mode. How to randomize them
-For the refractory period to be a good instrument, it is necessary that it is not overly affected by the network activity.
-This clearly is going to be problematic in many cases, after all the network activity affects refractoriness.
-However, there are multiple scenarios where refractoriness will be a good instrument.
-For example, if we have balanced excitation and inhibition we may expect largely independent refractory states of individual neurons.
-If a neuron biophysically implements something like conditional Poisson spiking its refractory states will be random.
-Even if neurons refractory states are strongly correlated during normal network operation there may be ways of randomizing them.
-Giving one burst of stimulation which is strong enough to elicit multiple spikes from each neuron may effectively randomize the phase of each neuron \citep{ermentrout2008reliability}.
-Importantly, we may expect the phase of a neuron to be far more random than the activity of the network as a whole.
-
-%Still some troubles
-We found some negative values of the IV estimator which were largely suppressed by requiring a hit rate $ < 90\%$.
-Hit rates larger than $90\%$ happens mainly at strong stimulation intensity which lower statistical power in the IV and can lead to correlated refractory times.
-We hypothesize strong stimulations can lead to synchrony induced by the stimulation \citep{ermentrout2008reliability}.
-This hypothesis was strengthened by observing that the negative values did not occur when the stimulation intensity was set to zero (data not shown).
-Furthermore, respective simulated neural network introduces much response overlap due to synapses having identical synaptic time constants and transfer delays.
-This makes inference quite hard since multiple neurons are affecting the same cell at the same time for each stimulation.
-However, this is less likely to occur in biological networks, which have high variability of synaptic properties and where firing patterns are sparser.
-These biological aspects would most likely work to the advantage of the IV method.
-Independence of refractoriness would be further improved, if in addition a clever stimulation routine was implemented such that the distribution of stimulation strength varies spatially from trial to trial.
-
-% Importance of randomized stimulus times and short pulses 
-The independence of refractory times is the one factor which makes or breaks the IV approach.
-Therefore we may think of ways of making the refractoriness distribution more random.
-First, it would help to use a task and situation where neurons are as uncorrelated as possible.
-Second, we may use a set of conditioning pulses of stimulation to increase independence of refractory states.
-Third, we may utilize chemical, behavioral, or molecular perturbations to randomize refractoriness.
-There has not, to the authors knowledge, been any attempts in neuroscience to randomize times of refractory states, so there may be a lot of possibilities to improve.
-
-Generalizing this idea, we may ask if there are ways of constructing good instrumental variables.
-One may assume a way of building molecular oscillators or otherwise pattern generators into neurons which affect their firing rate.
-Such modulatory activity would be observable in the extracellular activity of the neuron.
-Any neuron that then correlates to this modulation must be post-synaptic of the recorded neuron.
-Any kind of a signal that is local to a cell and not affected by network activity could produce a meaningful instrumental variable and it might be reasonably doable to link membrane channels to intracellular signal generation.
-Even if perfect IVs do not exist in brains, we may be able to make them.
-
 %Causal inference techniques to fix experimental issues
-There are many techniques for causal inference and most of them are largely unknown in neuroscience and are based on approximating randomness in a world that does not have it.
+There are many techniques for causal inference, most of which are largely unknown to the field of neuroscience, and are based on approximating randomness in a seemingly regular world. 
 In many cases, one could use regression discontinuity designs in a spiking system \citep{BenPaper, imbens2008regression}. 
 Moreover, one could use a difference in difference approach \citep{abadie2005semiparametric}. 
-Matching approaches \citep{stuart2010matching}, but see \citep{king2016propensity}, can be used where one compares similar network states and their evolution over time.
+Matching approaches \citep{stuart2010matching, king2016propensity}, can be used when comparing similar network states and their evolution over time. 
 In general, neuroscience is in a quest for causal interpretations, we should therefore be able to benefit considerably by utilizing techniques that are popular in the field of causal inference.
@@
 % Many trials needed here, good IV could possibly decrease this number
-%We note that we obtained low errors from the IV estimator after relatively few trials when estimating weights in the neural network of 1250 neurons.\todo{how would this be in the brain with more variable synaptic properties?} This amount of trials is not experimentally possible and we therefore foresee the need for improvements of the IV method.
-It can be improved in two ways, either with algorithms that uses more of the information available, such as the exact stimulus latencies and/or the amount of spikes in each trial.
-The other improvement could be to design experimental procedures that open for causal interpretation.
+%We note that we obtained low errors from the IV estimator after relatively few trials when estimating weights in the neural network of 1250 neurons.\todo{how would this be in the brain with more variable synaptic properties?}
+%This amount of trials is not experimentally possible and we therefore foresee the need for improvements of the IV method. 
+%It can be improved in two ways, either with algorithms that uses more of the information available, such as the exact stimulus latencies and/or the amount of spikes in each trial. 
+%The other improvement could be to design experimental procedures that open for causal interpretation.
 
 %If we assume that the assumption of random refractory periods are violated somehow by the physiology, e.g.a very synchronized network. 
-To ameliorate one could introduce multiple light sources that is activated in a randomized fashion such that the intensity distribution is randomized.
-This could be realized with e.g.
-laser holograms/filter or multiple micro LEDs This would in addition increase the amount of neurons targeted and thus give larger experimental yield.
-Another possibility is to design opsins that have a fluctuating baseline levels randomizing membrane potential of affected neurons, however this could also lead to a chronic deficit in respective neural network and would thus shorten the time span of the experimental procedure.
-But importantly, designing experiments so that causal inference is possible should be a major consideration for experimental design.
+%To ameliorate one could introduce multiple light sources that is activated in a randomized fashion such that the intensity distribution is randomized. 
+%This could be realized with e.g. laser holograms/filter or multiple micro LEDs. 
+%This would in addition increase the amount of neurons targeted and thus give larger experimental yield. 
+%Another possibility is to design opsins that have a fluctuating baseline levels randomizing membrane potential of affected neurons, however this could also lead to a chronic deficit in respective neural network and would thus shorten the time span of the experimental procedure.
+%But importantly, designing experiments so that causal inference is possible should be a major consideration for experimental design.
+
+%We applied the IV method to openly-availabe optotrode recordings of hippocampal neurons with short, pulselike stimulations. 
+%To its applicability, we apllied the IV method to an openly-availabe optotrode recording of hippocampal neurons with short, pulselike stimulations with high intensities and many repetitions. 
+%In accordance with our simulation results, the analysis revealed that the IV and CCH estimator differ in their interpretation of synaptic strengths. 
+%The IV and CCH estimates were weakly correlated. Several relative strong CCH estimates corresponded to small or even negative IV estimates. 
+%But without ground truth data, we are not able to determine the causes of the diverging interpretations.
+%Because of the low stimulation efficacy, we used an IV time window bigger than the refractory period, making the estimation susceptible to indirect paths, for example via feed-forward inhibition. 
+%Other aspects artifacts For future applications of the IV approach, we recommend the use of strong, short and unchanging pulse stimulations with a high number of repetitions, randomized times and little onset artifacts.
+%To verify the IV approach \textit{in vivo}, one could use for example recently developed two-photon set-ups, which allow for simultaneous recording and stimulation of many neurons with single-cell precision \citep{packer2014simultaneous, emiliani2015all}.
+%Single-cell randomized manipulations and pairwise recordings allow for causal interrogation of synaptic connections. 
+%Comparing transmission probabilities with IV estimates from recordings of potentially confounded wide-field stimulations may reveal the validity of the IV approach.
 
@@
 \end{equation}
-Here $ y $ is the dependent variable, $ x $ is the explanatory variable, $ \beta $ is the effect of the $ x$ on $y $ and $ u $ is an unknown error term.
+Here $ y $ is the dependent variable, $ x $ is the explanatory variable, $ \beta $ is the effect of $ x$ on $y $ and $ u $ is an unknown error term. 
 \Cref{eq:regress} follows from the causal path diagram \citep{wright1921correlation,wright1923theory}
@@
 \end{center}
-Assuming that changes in spike times $ y $ are described by $ \beta x $ i.e.
-$ \de{y}{x} = \beta $ for spike times $ x $.
+Assuming that changes in spike times $ y $ are described by $ \beta x $ i.e. $ \de{y}{x} = \beta $ for spike times $ x \forall x,y \in C^1$. 
 One problem with this idea is that in a confounded system, perfectly correlated neurons will give statistically indistinguishable $ \beta $. 
@@
 Here we have the relation $ \de{y}{x} = \beta + \de{u}{x} $. 
-To get at causality we thus require some stimulation that only highlight the activity in $ y $ caused by $ x $, disassociating $x$ from $u$.
+To get at causality we thus require some stimulation that only highlights the activity in $ y $ caused by $ x $, disassociating $x$ from $u$. 
 However, the optogenetic stimulation is not specific to $x$ and will activate parts of the network activity $ u $. 
@@
 \end{center}
-Here $ x_{s_r} $ represent times where the pre-synaptic neuron is refractory during stimulation. The true $ \beta $ is given by \citep{wright1928tariff}
+Here $ x_{s_r} $ represent times where the pre-synaptic neuron is refractory during stimulation. 
+The true $ \beta $ is given \citep{wright1928tariff} by
 
@@
 The successful stimulations of $x$ are denoted $x_s$ and thus $\bar{x}_s \equiv 1$. 
-Conversely $x_{s_r}$ denotes unsuccessful stimulations of $x$ i.e.
-stimulations of $x$ during its refractory state and $\bar{x}_{s_r} \equiv 0$.
+Conversely $x_{s_r}$ denotes unsuccessful stimulations of $x$ i.e. stimulations of $x$ during its refractory state and $\bar{x}_{s_r} \equiv 0$.
 
-To utilize the refractory period as an IV we first picked out one window of $ 4 $ ms for each of the pre-synaptic and post-synaptic neuron with a latency relative to stimulation time of $ 0 $ and $ \tau_{syn} + D $ ms (see \cref{eq:syn}) respectively.
-By classifying each window for each trial whether $x$ contained a spike we obtained the two binary arrays $ y_s $ and $ y_{sr} $.
+To utilize the refractory period as an IV on the simulated data we first picked out one window of $ 4 $ ms for each of the pre-synaptic and post-synaptic neuron with a latency relative to stimulation time of $ 0 $ and $ \tau_{syn} + D $ ms (see \cref{eq:syn}) respectively.
+By classifying each window for each trial whether $x$ contained a spike we obtained the two arrays $ y_s $ and $ y_{sr} $.
+
+We found some negative values of the IV estimator which were largely suppressed by requiring a hit rate $ < 90\%$. 
+Hit rates larger than $90\%$ happens mainly at strong stimulation intensity which lower statistical power in the IV or can lead to correlated refractory times. 
+We hypothesize that strong stimulations can lead to synchrony induced by the stimulation \citep{ermentrout2008reliability}. 
+This hypothesis was strengthened by observing that the negative values did not occur when the stimulation intensity was set to zero (data not shown). 
+Furthermore, the simulated neural network introduces much response overlap due to synapses having identical synaptic time constants and transfer delays. 
+This can interfere with inference since multiple neurons are affecting the same cell at the same time for each stimulation. 
+However, this is less likely to occur in biological networks, which have high variability of synaptic properties and where firing patterns are sparser. 
+These biological aspects would most likely work to the advantage of the IV method. 
+Independence of refractoriness would be further improved, if in addition a clever stimulation routine was implemented such that the distribution of stimulation strength varies spatially from trial to trial.
 
@@
 Briefly, to test if the cross correlation histogram (CCH) peak was significant we employed two tests. 
-By using the Poisson distribution with a continuity correction \citep{Stark2009} given by \cref{eq:poissoncontcor} we calculated $ p_{diff} $ by comparing the peak in positive time lag with the maximum peak in negative time lag \citep{English2017}.
+By using the Poisson distribution with a continuity correction \citep{Stark2009} given by \cref{eq:poissoncontcor} we calculated $ p_{diff} $ by comparing the peak in positive time lag with the maximum peak in negative time lag, called $p_{causal}$ in \citet{English2017}.
 The probability $ p_{fast} $ represents the difference between CCH and it's convolution with a hollow Gaussian kernel \citep{Stark2009}. 
-These two measures of significance were required to be $<0.01$
+These two measures of significance were required to be $<0.01$ and given by
 \begin{equation}
-p(N|\lambda(m)) = 1 - \sum_{k=0}^{N-1}\frac{e^{-\lambda(m)}\lambda(m)^k}{k!} - \frac{e^{-\lambda(m)}\lambda(m)^N}{2N!}
+p(N|\lambda(m)) = 1 - \sum_{k=0}^{N-1}\frac{e^{-\lambda(m)}\lambda(m)^k}{k!} - \frac{e^{-\lambda(m)}\lambda(m)^N}{2N!}.
 \label{eq:poissoncontcor}
@@
 Here $ \lambda $ represents the counts at bin $ m $ and $ N $ is the number of bins considered. 
-To estimate the connection weight between pairs we used the spike transmission probability defined in \cite{English2017} as
+To estimate the connection weight between pairs we used the spike transmission probability first defined in \citet{English2017} as 
 \begin{align}
 \label{eq:ptrans}
-P_{trans} = \frac{1}{n}\sum_{m=3ms}^{6ms} CCH(m) - \lambda_{Gauss}(m),
+p_{trans} = \frac{1}{n}\sum_{m=3ms}^{6ms} CCH(m) - \lambda_{Gauss}(m),
 \end{align}
@@
 \end{align}
-where $ u $ is an error term. Further, we used the logit link function such that the the probability giving the proxy for synaptic weight is given by
+where $ u $ is an error term. 
+Further, we used the logit link function such that the the probability giving the proxy for synaptic weight is given by
 \begin{align}
@@
 \end{align}
-Here $ \tau_{syn} $ denotes the synaptic integration time constant and $ H $ is the Heaviside step function. All neurons were driven by an external Poisson process with rate $ rate_{p} $.
+Here $ \tau_{syn} $ denotes the synaptic integration time constant and $ H $ is the Heaviside step function. 
+All neurons were driven by an external Poisson process with rate $ rate_{p} $.
 
-Synaptic weights were log-normally distributed such that the increase in membrane potential $ V_m^i $ due to one spike were restricted to lie between $ V_{syn} = 0.05 mV $ and $ V_{syn} = 2.05 mV $ based on experimental findings \citep{Sayer1990,Mason1991}.
+Synaptic weights were log-normally distributed such that the increase in membrane potential $ V_m^i $ due to one spike were restricted to lie between $ V_{syn} = 0.0 mV $ and $ V_{syn} = 2.0 mV $ based on experimental findings \citep{Sayer1990,Mason1991}.
 The synaptic distribution is shown in \cref{fig:state}\labelcref{fig:syn_dist} where the inhibitory PSC amplitude is given by $ J_{in} = g J_{ex} $ where $ J_{ex} $ denotes the excitatory synaptic weight.
@@
 To find suitable parameters yielding asynchronous activity we measured the population correlation coefficient given by
-
 \begin{align}
@@
 where $ h $ is the spike time histogram with binsize at $ 5ms $ for neuron $ i,j $ and $ \mean{\cdot} $ is the mean operator. 
-The distribution of $ CC $ is shown in \cref{fig:cchvswald} which were found by performing several parameter sweeps picking three parameter sets which mainly differed in firing rate (data not shown).
+The distribution of $ CC $ is shown in \cref{fig:cch_iv_triple} which were found by performing several parameter sweeps picking three parameter sets which mainly differed in firing rate (data not shown).
 
@@
 where $ ISI $ denotes the inter-spike interval of neuron $ i $. 
-Due to the finite time synaptic integration time constant $ \tau_{syn} = 1 ms $ we were unable to have the network showing an irregular state; see \cref{fig:state}\labelcref{fig:CV}.
+We were unable to have the network showing an irregular state; see \cref{fig:state}\labelcref{fig:CV} partly to the finite synaptic integration time constant $ \tau_{syn} = 1 ms $.
 To verify that indeed this was due to $ \tau_{syn} $ we performed several simulations with lower $ \tau_{syn} $ obtaining $ \mean{CV}_{pop} > 1 $ (data not shown).
@@
 \begin{table}
+\centering
 \begin{tabular}{lllll}
 \toprule
-{} &    model 1 &    model 2 &    model 3 &       units \\
+name &    model 1 &    model 2 &    model 3 &       units \\
 \midrule
@@
 $J_{low}^{in}$        &  0.0224058 &   &   &          pA \\
+$time_{simulation}$   &    3685312 &        &           &         ms \\
+$rate_{in}$           &    8.69 &    10.56 &     12.7 &          Hz \\
+$rate_{ex}$           &    6.5 &    9.22 &    11.5 &          Hz \\
+\bottomrule
+\end{tabular}
+\caption{\label{tab:params} Simulation parameters of three different models.}
+\end{table}
+
+\begin{table}
+\centering
+\begin{tabular}{lllll}
+\toprule
+name &    model 1 &    model 2 &    model 3 &       units \\
+\midrule
 $stim_{N}^{in}$       &          0 &           &           &             \\
@@
 $stim_{max}^{period}$ &        150 &         &         &          ms \\
-$rate_{in}$           &    7.17024 &    9.66335 &     11.754 &          Hz \\
-$rate_{ex}$           &    9.05793 &    10.9176 &    13.1032 &          Hz \\
-$density$             &     100000 &      &      &  Nmm$^{-3}$ \\
+
+$density$             &     7514 &      &      &  Nmm$^{-3}$ \\
 $S$                   &       10.3 &       &      &   mm$^{-1}$ \\
@@
 $n$                   &       1.36 &        &        &             \\
+$n_{Hill}$            &        0.76 &        &        &             \\
+$K$                   &       0.84 &        &        &             \\
+$depth$               &       0.7 &        &        &             mm\\
 \bottomrule
 \end{tabular}
-\caption{\label{tab:params} Simulation parameters of three different models.}
+\caption{\label{tab:params_stim} Stimulation parameters of three different models.}
 \end{table}
 
-\subsection{Calculating errors}
-\subsubsection{Mean square error}
+\subsection{Calculating the mean square error}
 The conditional probability of neuron $y$ firing given a spike from neuron $x$ denoted $P(y|x)$ is related to the connection strength $w_{xy}$ and the background activity. 
-Since LIF neurons integrate linearly and have fixed thresholds we expect $P(y|x)$ to be proportional to the connection strength $w_{xy}$, i.e.
-$P(y|x) ~ w_{xy}$.
-To calculate the mean squared errors we thus normalized $w_{xy}$ to the range $[0, max(IV_{xy})]$ or $[0, max(CCH_{xy})]$ for the IV and CCH estimators respectively by
+Since LIF neurons integrate linearly and have fixed thresholds we expect $P(y|x)$ to be proportional to the connection strength $w_{xy}$, i.e.  $P(y|x) ~ w_{xy}$. 
+To calculate the mean squared errors we thus normalized $w_{xy}$ to the range $[0, max(IV_{xy})]$ or $[0, max(CCH_{xy})]$ to calculate the MSE of the IV and CCH estimators respectively by
 \begin{equation*}
-    \frac{1}{N}\sum(\hat{w} - \bar{w})^2).
-\end{equation*}
-Here $\bar{w}$ is the normalized weight and $\hat{w}$ is the estimated connection coefficient.
+    MSE(w) = \frac{1}{N}\sum(\hat{w} - \bar{w})^2).
+\end{equation*}We found some negative values of the IV estimator which were largely suppressed by requiring a hit rate $ < 90\%$. 
+Hit rates larger than $90\%$ happens mainly at strong stimulation intensity which lower statistical power in the IV or can lead to correlated refractory times. 
+We hypothesize that strong stimulations can lead to synchrony induced by the stimulation \citep{ermentrout2008reliability}. 
+This hypothesis was strengthened by observing that the negative values did not occur when the stimulation intensity was set to zero (data not shown). 
+Furthermore, the simulated neural network introduces much response overlap due to synapses having identical synaptic time constants and transfer delays. 
+This can interfere with inference since multiple neurons are affecting the same cell at the same time for each stimulation. 
+However, this is less likely to occur in biological networks, which have high variability of synaptic properties and where firing patterns are sparser. 
+These biological aspects would most likely work to the advantage of the IV method. 
+Independence of refractoriness would be further improved, if in addition a clever stimulation routine was implemented such that the distribution of stimulation strength varies spatially from trial to trial.
+Here $\bar{w}$ is the normalized weight and $\hat{w}$ is the estimated weight.
+
+\subsection{Calculating goodness of fit and false estimates}
+The strength of the perturbations was at the maximum very large and led to many instances where the hit-rate was above 90\%. 
+This represents extreme experimental conditions with very high light intensity and would yield large errors \cref{fig:error}. 
+To see how the IV and CCH estimators compared we thus selected source neurons that had strictly less hit-rate than 90\%. 
+This led to the maximum perturbation strength of $5pA$.
 
@@
 \subsubsection{False positives}
+We calculated false positives from a subset $N_{sub}$ of the pairs $N$ where the true weight was $w = 0$. 
+Then a false positive was defined as an estimate that were larger than $0.05$. 
+The percentage of false positives was then calculated by $N_{sub}(x > 0.05) / N_{sub}$.
 
 \subsubsection{False negatives}
+We calculated false negatives from a subset $N_{sub}$ of the pairs $N$ where the true weight was $w = 0$.  
+Then a false negative was defined as an estimate that had a difference from the true weight which were larger than $0.05$. 
+The percentage of false negatives was then calculated by $N_{sub}(|x - w| > 0.05) / N_{sub}$.
 
 \subsection{Perturbation intensity}\label{sec:method:opto}
-In order to replicate an optogenetic experiment we modeled transmission of light through brain tissue with the Kubelka-Munk model for diffuse scattering in planar, homogeneous, ideal diffusing media given by
+In order to replicate an optogenetic experiment we modeled transmission of light through brain tissue with the Kubelka-Munk model for diffuse scattering in planar, homogeneous, ideal diffusing media \citep{Ho2017} given by
 \begin{align}
@@
 \end{align}
-Here $ T $ denotes a transmisison fraction, $ S $ is the scattering coefficient for mice \citep{Aravanis2007} and $ r $ is the distance from a light source \citep{Ho2017}.
+Here $ T $ denotes a transmisison fraction, $ S $ is the scattering coefficient for mice \citep{Aravanis2007} and $ r $ is the distance from a light source. 
 Further we combined diffusion with geometric loss assuming that absorption is negligible as in \cite{Aravanis2007} and computed the intensity as presented in \cref{fig:concept} by
@@
 \end{align}
-Here $ d $ is the diameter of the optical fiber, $ NA $ is the numerical aperture of the optical fiber and $ n $ is the refraction index for gray matter \citep{Ho2017}; see numerical values for parameters in \cref{tab:params}.
+Here $ d $ is the diameter of the optical fiber, $ NA $ is the numerical aperture of the optical fiber and $ n $ is the refraction index for gray matter \citep{Ho2017}; see numerical values for parameters in \cref{tab:params_stim}.
 
-To estimate the distribution of light intensity on affected neurons we assumed a neuron density of $ 10^4 Nmm^{-3} $ and found the volume of a cut cone that could contain the number of stimulated excitatory neurons which were found to yield the depth $ r_{max} = 0.175 mm $.
-We then selected $ N_stim $ neurons that were given a random position in the range $ [0, r_{max}] $ and were assigned a stimulation strength as the maximum stimulation strength multiplied by \cref{eq:intensity}.
-Then we selected $ 50 $ of the excitatory neurons that were not stimulated as the ``target" population which together with the inhibitory neurons were not perturbed directly by the light stimulus.
+To estimate the distribution of light intensity on stimulated neurons we distributed $ 795 $ neurons uniformly in 10 spharical slices in the range $[0, 1 mm]$ which had a radius given by the cone shaped light; see \ref{fig:concept} inset.
 
-In order to keep the stimulus model as simple as possible we let set the maximum stimulation strength to $ 10 pA $ which was found suitable by investigating the percentage of successful stimulations to be almost $ 100\% $.
+To further estimate the peak amplitude photo current we used the Hill equation fitted by parameters found in \citet{wang2007high} given by 
+\begin{equation}
+    P = I_{max} \frac{I^n}{K^n + I^n}
+\label{eq:hill}
+\end{equation}
+Here, $I_{max} = 642 pA$ is the maximum curren, $n=0.76$ is the Hill coefficient and $K = 0.84 mW/mm^2$ represents the half-maximal light sensitivity of the ChR2. 
+We further used the light intensity $I$ given by \cref{eq:intensity} multiplied by an initial intensity of $10mW/mm^2$.
 
+Since the model neurons are not scaled to mimic ``realistic'' values in their membrane potential we set the maximum stimulation strength to $ 8 pA $ which was found suitable by investigating the percentage of successful stimulations to be $ < 100\% $.
+We then selected $ 100 $ of the excitatory neurons that were not stimulated as the ``target" population which together with the inhibitory neurons were not perturbed directly by the light stimulus.
+
+\subsection{Application to hippocampal recordings}\label{sec:method:realdata}
+We applied the IV method to CA1 recordings of two mice, downloaded from \url{https://buzsakilab.nyumc.org/datasets/McKenzieS/}.
+These recordings came from silicon probes of four shanks with thirty two channels and had integrated $\mu$LEDs implanted in hippocampal region CA1.
+Mice expressed channelrhodopsin-2 under the control of an excitatory neuron-specific promoter, CaMKII::ChR2 \citep{English2017}.
+Each animal was recorded on several days, while mice freely behaved in their home cage.
+Each recording lasted more than $3$ hours.
+Spikes were then sorted by the experimenter using \todo{klustakwik??}.
+During analysis, we treated each recording day independently.
+During each session, sinusoidal and pulse stimulations lasting $10$ or more milliseconds were applied with different intensities.
+The experimenter determined whether a unit was optogenetically stimulated by applying two criteria \citep{English2017}.
+First, the number of spikes during each stimulation was noted and compared to the number of spikes in the same interval but two seconds before the stimulation.
+A unit was considered optogenetically labeled, if the p-value of a Wilcoxon ranksum non-parametrical test of means was below $10^-10$.
+Second, it was required that the absolute number of spikes during stimulation is on average 50\% larger compared to the number of spikes in same the interval, but two seconds before stimulation.
+
+We considered only sessions containing pulse-stimulations.
+At each shank, we grouped similar stimulation intensities.
+By cross-correlating stimulation onset time with spikes, we quantified the required time for each intensity group to significantly increase the firing rate above baseline.
+We used a bin-width of $3$ ms, calculated the baseline by taking the mean of the cross-correlogram for negative time lags, $-45$ to $-1.5$ ms, and applied a significance threshold of $0.01$ on the probability to obtain the observed or a higher count per bin of the poisson distribution with continuity correction \citep{abeles1982quantification}.
+For each labeled neuron, those intensities were selected that caused a significant increase in firing probability within $7.5$ ms.
+A labeled unit was only considered a putative pre-synaptic partner if the number of significant pulse stimulation events exceeded $2500$.
+In total, connections between 17 putative pre-synaptic and 86 putative post-synaptic units were calculated.
+
+We calculated the CCH based connection strength, $p_{trans}$, according to \citep{English2017} and section \labelcref{sec:method:cch}.
+First, spiketrains of optogenetically labeled, putative pre-synaptic as well as unlabeled, putative post-synaptic units were binned with $0.4 ms$ binwidth.
+The CCH has been convolved with a partially hollow gaussian kernel, with $10 ms$ standard deviation and a hollow fraction of $0.6$.
+We selected a time window for $p_{fast}$ and $p_{trans}$ to be between $0.8$ and $2.8 ms$.
+The reference negative time window for $p_{diff}$ was $-2$ to $0 ms$.
+We further applied the same significance level of 0.01 applied to $p_{fast}$ and $p_{diff}$.
+
+For computing the IV estimate, we used a window size of $7.5 ms$ and a fixed synaptic delay of $1 ms$.
+IV and CCH estimates were separately bootstrapped with a sample size of $1000$.
+For CCH estimates, we employed the additive property of the CCH function.
+We divided spiketrains of a session in chunks of $100 s$, and calculated the CCH for each chunk.
+We then randomly selected chunks, with replacement, to match the full length of the session.
+CCH estimates were calculated on the sum of chunks.
+For the IV estimates, we randomly chose stimulations with replacement to match the original number of stimulations and calculated the estimate on those samples.
+
+\FloatBarrier
+%-------------------------------------------------------------------------------
+%                               Acknowledgments
+%%-------------------------------------------------------------------------------
+\section{Acknowledgments}
+This research was partly funded by the Research Council of Norway Grant 231248 and the University of Oslo.
+T.S (PhD fellow), M.F. (PI) are part of the Simula-UCSD-University of Oslo Research
+and PhD training (SUURPh) program, an international collaboration in computational biology and
+medicine funded by the Norwegian Ministry of Education and Research.
+We thank Sam McKenzie and Daniel Fine English for sharing and explaining the optotrode data.
 %-------------------------------------------------------------------------------

### 2018-10-23 params_1.py

@@ -37,7 +37,7 @@ parameters = {
     'stim_period'     : 100.0, # ms
     'stim_max_period' : 150, # only applies to poisson
     'stim_duration'   : 2.0, # ms
-    'stim_N'          : 300,
+    'stim_N'          : 30000,
     # Optogenetics
     'I0': 10, # light intensity leaving fibre mW/mm2
     'r': 100e-3, # 100 um radius of fiber

### 2018-10-23 run_analysis.py

@@ -70,7 +70,7 @@ for N_trials in trials:
                 logreg = np.nan
                 logreg_intercept = np.nan
 
-            stim_amp = float(stim_amps[stim_amps.node==source].amp.values)
+            stim_amp = float(stim_amps[source])
             # cc, cv and stuff
             w = conn[(conn.source==source) & (conn.target==target)].weight
             n_syn = len(w)

### 2018-10-23 run_batch.py

@@ -6,20 +6,20 @@ import tempfile
 params = sys.argv[1:]
 ################################################################################
 ################################# SIMULATION ###################################
-# processes = []
-# for param in params:
-#     name = op.split(param)[-1].replace('.py', '')
-#     f = tempfile.TemporaryFile()
-#     p = subprocess.Popen(['python', 'simulator.py', param], stdout=f)
-#     processes.append((p, f, name))
-#
-# for p, f, name in processes:
-#     p.wait()
-#     f.seek(0)
-#     logname = op.join("log_sim_{}.txt".format(name))
-#     with open(logname, "wb") as logfile:
-#         logfile.write(f.read())
-#     f.close()
+processes = []
+for param in params:
+    name = op.split(param)[-1].replace('.py', '')
+    f = tempfile.TemporaryFile()
+    p = subprocess.Popen(['python', 'simulator.py', param], stdout=f)
+    processes.append((p, f, name))
+
+for p, f, name in processes:
+    p.wait()
+    f.seek(0)
+    logname = op.join("log_sim_{}.txt".format(name))
+    with open(logname, "wb") as logfile:
+        logfile.write(f.read())
+    f.close()
 
 ################################################################################
 ################################# ANALYSIS #####################################

### 2018-10-23 simulator.py

@@ -285,14 +285,7 @@ class Simulator:
         # Simulate one period without stimulation
         nest.Simulate(self.stim_times[0])
 
-
-        self.stim_nodes_ex = self.nodes_ex[:self.p['stim_N_ex']]
-        self.stim_nodes_in = self.nodes_in[:self.p['stim_N_in']]
-
-        assert all(np.in1d(self.stim_nodes_ex, self.nodes_ex))
-        assert all(np.in1d(self.stim_nodes_in, self.nodes_in))
         # Set dc stimulation
-        self.stim_amps = {'ex': [], 'in': []}
         stims = []
         z = np.linspace(0, self.p['depth'], self.p['N_pos'])
         N_slice = affected_neurons(z).astype(int)
@@ -301,12 +294,16 @@ class Simulator:
         A = hill(self.p['I0'] * I)
         A = A / A.max()
         idx = 0
+        self.stim_amps_ex = {}
+        self.stim_amps_in = {}
+        self.stim_nodes_ex = []
+        self.stim_nodes_in = []
         for i, N_stim in enumerate(N_slice):
-            nodes = self.stim_nodes_ex[idx:idx + N_stim]
+            nodes = self.nodes_ex[idx:idx + N_stim]
+            self.stim_nodes_ex.extend(list(nodes))
             idx += N_stim
             amp = A[i] * self.p['stim_amp_ex']
-            stim_amps = [{'node': n, 'amp': amp} for n in nodes]
-            self.stim_amps['ex'].extend(stim_amps)
+            self.stim_amps_ex.update({n: amp for n in nodes})
             stim = nest.Create(
                 "dc_generator",
                 params={'amplitude': amp,
@@ -315,6 +312,8 @@ class Simulator:
             nest.Connect(stim, nodes)
             stims.append(stim)
 
+        assert all(np.in1d(self.stim_nodes_ex, self.nodes_ex))
+        assert all(np.in1d(self.stim_nodes_in, self.nodes_in))
         # Run multiple trials
         simtimes = np.concatenate((np.diff(self.stim_times),
                                   [np.min(np.diff(self.stim_times))]))
@@ -399,8 +398,8 @@ class Simulator:
                 'ex': self.stim_nodes_ex,
                 'in': self.stim_nodes_in},
             'stim_amps': {
-                'ex': pd.DataFrame(self.stim_amps['ex']),
[…]

### 2018-10-23 tools_plot.py

@@ -3,7 +3,7 @@ import numpy as np
 import pandas as pd
 import seaborn as sns
 import quantities as pq
-#from tools_analysis import make_spiketrain_trials
+from tools_analysis import make_spiketrain_trials
 import neo
 import os.path as op
 
@@ -155,13 +155,11 @@ def label_diff(x1, x2, y, text, ax, color='k', pad_txt=.03, pad_line=0.01):
 
 def plot_psth(spike_train=None, epoch=None, trials=None, xlim=[None, None],
               fig=None, axs=None, legend_loc=1, color='b',
-              title='', stim_alpha=.2, stim_color=None,
-              stim_label='Stim on', stim_style='patch', stim_offset=0*pq.s,
               rast_ylabel='Trials', rast_size=10,
               hist_color=None, hist_edgecolor=None,
               hist_ylim=None,  hist_ylabel=None,
               hist_output='counts', hist_binsize=None, hist_nbins=100,
-              hist_alpha=1.):
+              hist_alpha=1., hist_log=False):
     """
     Visualize clustering on amplitude at detection point
 
@@ -176,12 +174,10 @@ def plot_psth(spike_train=None, epoch=None, trials=None, xlim=[None, None],
     axs : matplotlib axes (must be 2)
     legend_loc : 'outside' or matplotlib standard loc
     color : color of spikes
-    title : figure title
     stim_alpha : float
     stim_color : str
     stim_label : str
     stim_style : 'patch' or 'line'
-    stim_offset : pq.Quantity
         The amount of offset for the stimulus relative to epoch.
     rast_ylabel : str
     hist_color : str
@@ -216,14 +212,7 @@ def plot_psth(spike_train=None, epoch=None, trials=None, xlim=[None, None],
     else:
         assert spike_train is None
     dim = trials[0].times.dimensionality
-    if stim_style == 'patch':
-        if epoch is not None:
-            stim_duration = epoch.durations.rescale(dim).magnitude.max()
-        else:
-            warnings.warn('Unable to acquire stimulus duration, setting ' +
-                          'stim_style to "line". Please provede "epoch"' +
-                          ' in order to use stim_style "patch".')
-            stim_style = 'line'
+
     # raster
     plot_raster(trials, color=color, ax=rast_ax, ylabel=rast_ylabel,
                 marker_size=rast_size)
[…]

### 2018-11-01 manuscript/main.tex
update manuscript to latest

@@
 \usepackage{graphicx,subcaption}
-\usepackage{todonotes} %[disable]
 \usepackage{cleveref}
@@
 \usetikzlibrary{arrows,shapes,calc}
+\usetikzlibrary{positioning}
 
@@
 \begin{abstract} % max words 150
-\noindent To study how the brain works, it is essential to identify causal interactions between neurons, usually requiring perturbations.
-However, when using optogenetics we typically perturb multiple cells, producing a confound - any of the stimulated neurons can have affected the target neuron.
+\noindent To study how the brain works, it is crucial to identify causal interactions between neurons, which is thought to require perturbations.
+However, when using optogenetics we typically perturb multiple neurons, producing a confound - any of the stimulated neurons can have affected the target.
 Here we show how this produces large biases, and how they can be reduced using the instrumental variable (IV) framework from econometrics.
 The interaction between stimulation and the absolute refractory period produces a weak, approximately random signal which can be exploited to estimate causality.
-When simulating integrate-and-fire neurons, we find that IV estimates connectivity better than na\"ive techniques ($R^2=0.845$ vs $R^2=0.048$).
+When simulating integrate-and-fire neurons, we find that causal connectivity estimates from IV are better than na\"ive techniques ($R^2=0.77$ vs $R^2=0.01$).
 The difference is important as the estimates differ when applied to experimental data from stimulated neurons with recorded spiking activity.
@@
 % The brain is infinitely confounded, thus correlations are not sufficient to draw conclusions about mechanisms and this is a hard problem in brain data.
-A goal of neuroscience is to understand the mechanisms or causal chains that give rise to activity in the brain, to perception, cognition, and action. 
+The central goal of neuroscience, arguably, is to understand the mechanisms or causal chains that give rise to activity in the brain, to perception, cognition, and action. 
 Complex systems such as the brain are hard to understand because of the numerous ways the contributing elements may interact internally \citep{Jonas2017}. 
@@
 After all, there can be many ways that the same activities emerge from distinct causal chains \citep{drton2011global, peters2017elements}. 
-Reaching a mechanistic level of understanding in the human brain is nearly impossible as it contains approximately 86 billion neurons \citep{azevedo2009equal}, each of which influences many other neurons.
+Reaching a mechanistic level of understanding in the mammalian brain is incredibly hard as they contain countless neurons (e.g.
+86 billion in the human brain \citep{azevedo2009equal}), each of which influences many other neurons.
 Even if we could record all neurons at the same time, estimating causality and producing a mechanistic understanding would be extremely challenging.
@@
 % We use observational studies, where identifiability is questionable and causality is thus typically wrong. 
-Typically, we only record from a small subset of all neurons. 
+In today's typical studies, we only record from a small subset of all neurons. 
 The data we obtain from such recordings, e.g. from electrophysiology or calcium imaging, is observational, which means that it does not result from randomized perturbations. 
 In such cases, we can never know to which level the observed activity was caused by other observed activity, or by unobserved activity. 
-The unobserved neurons are called confounders.
+The activity of the unobserved neurons is thus called confounders.
 If mechanisms are estimated from observational data in the presence of confounders, the consequence may be large errors and incorrect conclusions \citep{angrist2008mostly}. 
@@
 % Set up confounding
-Confounding is the big threat to causal validity \citep{UpcomingMehlerPaper} irrespective of the use of simple regression techniques or advanced functional connectivity techniques \citep{stevenson2008inferring, honey2009predicting, aitchison2017or, pfau2013robust}.
-To estimate connectivity it is first and foremost important that it stems from cause and effect, therefore we use the term causal connectivity. 
-Consider an example where we want to estimate causal connectivity between two observed, but unconnected neurons, $ A $ and $ C $ (\cref{fig:intro}(a)). 
+Confounding is the big threat to causal validity \citep{pearl2009causality} irrespective of the use of simple regression techniques or advanced functional connectivity techniques \citep{stevenson2008inferring, honey2009predicting, aitchison2017or, pfau2013robust}.
+To estimate connectivity it is first and foremost important that the used signals reflect cause and effect, therefore we use the term causal connectivity. 
+Consider an example where we want to estimate causal connectivity between two observed, but unconnected neurons, $ A $ and $ C $ (\cref{fig:intro}(a)) by stimulating optogenetically. 
 A third unobserved neuron $ B $, is driven together with $ A $ by a common input and they are thus strongly correlated.
 Furthermore, $ B $ and $ C $ are connected, thus the input drives $ C $ through $ B $.
-Consequently, $ A $ and $ C $ are also correlated and the regression $ C = \beta A $ will misleadingly conclude a direct interaction when causally interpreted. 
+Consequently, $ A $ and $ C $ are also correlated and the regression $ C = \beta A + noise$ will misleadingly conclude a direct interaction when causally interpreted. 
 In this case, we say the regressor $ A $ is endogenous and the regression coefficient $ \beta $ estimates the magnitude of association rather than the magnitude of causation. 
@@
 Multiple regression may be a solution to confounding problems as they support ``explaining away" background activity \citep{stevenson2008inferring}.
-However, this is only a good strategy if most neurons are included in the recordings.
+However, this is only a meaningful strategy if most neurons are included in the recordings.
 Furthermore, only under certain assumptions about nonlinearity or noise sources does a fully observed system become identifiable \citep{daniusis2012inferring,shimizu2006linear}.
@@
 If we stimulate single neurons, the ability to estimate causal relationships by regression is within reach. 
-However, this is experimentally challenging and yield low cell count because it requires intracellular, juxtacellular or two-photon stimulation \citep{pinault1996novel, lerman2017two, nikolenko2007two, emiliani2015all}.
+However, this is experimentally challenging and yields low cell count because it requires intracellular, juxtacellular or two-photon stimulation \citep{pinault1996novel, lerman2017two, nikolenko2007two, emiliani2015all}.
 Because gold-standard perturbations are challenging, it is necessary and highly desirable if causality could be obtained from optogenetic stimulation in combination with neural recordings of large populations of neurons \citep{boyden2005millisecond, zemelman2002selective}.
@@
 This distributed pattern of stimulation produces activity which then percolates through the network of neurons. 
-Thus any post-synaptic activity induced by stimulation could in principle come from any of the stimulated neurons introducing problematic confounders.
+Thus any post-synaptic activity induced by stimulation could in principle come from any of the stimulated neurons, introducing problematic confounders.
 
@@
 The inference of causality from observational data is addressed in the fields of statistics \citep{pearl2009causality}, machine learning \citep{peters2017elements} and econometrics \citep{angrist2008mostly}.
+These fields have extensively worked on methods to estimate causality in the face of potential confounding and may offer us clues on how to solve our problems.
 
@@
 A commonly used approach towards causal inference in economics are instrumental variables. 
-Let's say that we want to estimate the return $ \beta $ from education $ x $ to yearly wages $ y $ with the regression $ y = \beta x + u $. 
+Let us say that we want to estimate the return $ \beta $ from education $ x $ to yearly wages $ y $ with the regression $ y = \beta x + u $. 
 Here $ u $ are the factors other than education that contribute to yearly wages. 
-One of the factors in $ u $ is a person's cognitive ability which may also affect education and thus the regressor $ x $  is correlated with the error term $ u $. 
-This will imply that the regression estimate $ \beta $ will not estimate the magnitude of causation from education on wages, but rather, its association. 
+One of the factors in $ u $ is a person's cognitive ability which may also affect education. The regressor $ x $  is thus correlated with the error term $ u $. 
+This will imply that the regression estimate of $ \beta $ will not estimate the magnitude of causation from education on wages, but rather, its association. 
 In this case one may use the proximity to a college or university as an instrumental variable (IV) \citep{card1993using}. 
-Proximity is a good instrument since it can only affect wages through schooling. 
+This is following the idea that proximity to a college does not affect your cognitive ability but may affect your proclivity to attend college.
 We expect living in proximity of colleges and universities to give higher probability to attend without affecting other contributing factors to wages such as cognitive ability. 
@@
 % Worry is this really correct? Yes, its provable. IVs are (under the right assumptions) actually correct.
-The IV technique has been used extensively in econometrics and is provable causal given three assumptions \citep{angrist2008mostly}. 
+The IV technique has been used extensively in econometrics and can provide provable unbiased causal estimates given three main assumptions \citep{angrist2008mostly}. 
 First, the instrument must be decorrelated with the error term. 
@@
 For example, the overall activity of the network is due to slow and strongly nonrandom dynamics. 
-In contrast, the temporal pattern of when a neuron is in a refractory state may be random. 
+In contrast, the temporal pattern of when a neuron is in a refractory state may be in good approximation random. 
 First, if neurons are spiking according to conditional Poisson distributions, their exact timing conditioned on the network state, will be random. 
-While refractoriness may not be perfectly random, the exact times of spiking are notoriously difficult to predict \citep{stevenson2008inferring} suggesting that refractoriness is sufficiently random.
+While refractoriness may not be perfectly random, the exact times of spiking are notoriously difficult to predict \citep{stevenson2008inferring} suggesting that refractoriness is quite random.
 
@@
 \subsection{Optogenetics is not local}
-Optogenetic stimulation is generally seen as a perturbation that by and large affects neurons in proximity (< 1mm) of the light source with the effect decreasing with distance. 
-However, this may be an incorrect way of conceptualizing the spatial effect of stimulation as it depends on multiple factors. 
+Optogenetic stimulation is generally seen as a perturbation that by-and-large affects neurons in proximity (< 1mm) of the light source with the effect decreasing with distance. 
+However, this is a misleading way of conceptualizing the spatial effect of stimulation as it depends on multiple factors. 
 Light intensity and opsin density are important as more light and ion channels will cause a stronger effect on each cell. 
@@
 Finally, physiological properties of the cells are important as light may have a stronger effect on spiking activity when the membrane potential of the cell is sufficiently close to the firing threshold.
-The induced effect of optogenetic stimulation as a function of distance should be given by a relation between the four parameters: light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and the distribution of induced photo current.
+The induced effect of optogenetic stimulation as a function of distance should be given by a relation between the four parameters: light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and the distribution of induced photo currents.
 
 To estimate the light intensity, we calculated the spatial extent of laser light delivered by fiber-optics under plausible experimental conditions according to \citet{Aravanis2007}; see \cref{sec:method:opto}.
+While their experiment uses blue light, their fits assume no absorption making the equations somewhat general.
 This modeling of light intensity yield an approximately $ 1/r^2 $ reduction with distance $ r $ from the stimulation site \cref{fig:concept} (cyan line). 
 This is explained by the surface of a 3D shell growing with $4\pi r^2$ and photons will be roughly isotropic (but see \citep{thunemann2018does}) beyond the scattering length \cref{fig:concept} (inset).
-The same number of photons has to cross each of the spheres around the stimulation location unless they are absorbed. 
-As a result, the density of photons decreases rapidly with distance.
+The same number of photons has to cross each of the spheres around the stimulation location unless they are absorbed or scattered inwards. 
+As a result, the density of photons decreases with distance.
 
-The number of illuminated neurons at a given distance will increase with distance to the stimulation site given that neurons are uniformly distributed in brain tissue \cref{fig:concept} (black line). 
+The number of illuminated neurons at a given distance will, however, increase with distance to the stimulation site given that neurons are roughly uniformly distributed in brain tissue \cref{fig:concept} (black line).
 In fact, it will increase by approximately $ r^2 $ with distance. 
-This derives from the same surface scaling as for the 3D shell. 
+This derives from the same surface scaling as for the 3D shell as for the photon flow. 
 Thus the number of neurons that can be activated increases rapidly with distance.
 
-To estimate the effect of stimulation, the functionality underlying spiking activity needs to be considered. 
+To estimate the effect of stimulation, the mechanism with which light affects spiking activity needs to be considered. 
 This can largely be characterized by the  distribution of membrane potentials across neurons. 
@@
 The peak amplitude of the photo current relates approximately logarithmically to the light intensity \citep{wang2007high}; see \cref{sec:method:opto}. 
-Assuming that opsins are evenly distributed across neurons, the induced photo current will not be proportional with light intensity - it will fall slower. 
+Assuming that opsins are evenly distributed across neurons, the induced photo current will not be proportional to light intensity - it will fall slower. 
 Based on this, we calculate the overall stimulation effect to be the product of the number of neurons in a spherical slice and the peak amplitude photo current. 
-This product actually increases with distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line). 
+This product actually increases with distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line). In other words, there is more activation at 500um than at 100um.
 Thus, optogenetic stimulation utilizing single photon activation does not produce a localized effect.
 
-\begin{figure}\includegraphics[scale=1]{opto-powerlaw}
+\begin{figure}\includegraphics[scale=.95]{opto-powerlaw}
 \caption{{\bf Spatial extent of optogenetic stimulus}. 
@@
 If neurons are uniformly distributed, the number of affected neurons in a spherical slice increases by $ r^{2} $ (N, black line). 
-The total photo current (sum over neurons of peak amplitude photo current in a spherical slice) thus increases with distance due to the nonlinear relation between light intensity and photo current (P, blue line) depicted as percentage of maximum.}
+The total photo current (P, blue line) calculated as the sum over neurons of peak amplitude photo current in a spherical slice thus increases with distance due to the nonlinear relation between light intensity and photo current, depicted as percentage of maximum.}
 \label{fig:concept}
@@
 This further generates a strong correlation between A and C, confounding the system by rendering the cross-correlation histograms (CCHs) between BC and AC both statistically significant ($ p_{\mathrm{fast}} < 0.001, p_{\mathrm{diff}} < 0.001 $; see \cref{sec:method:cch}).
+A na\"ive reading of this result may suggest causal influences of both A and B on C.
+
 Even though the correlation peak between B and C is larger than between A and C, due to correlated spikes outside the periods with stimulation, one may imagine a situation where only A and C is measured, giving rise to a false prediction that they are connected.
+Alternatively, A may have a stronger response to the optogenetic stimulation, in which case we may even have a stronger A-C correlation than B-C correlation.  
 If stimulation affects multiple neurons simultaneously, there is a real confounding problem.
@@
 \caption{} \label{fig:intro:2}
-\end{subfigure}\medskip\\
+\end{subfigure}\\
 \begin{subfigure}{\textwidth}\centering \includegraphics[scale=1]{xcorr}
@@
 \caption{{\bf Optogenetic stimulation induces spurious correlations}. 
-A sketch of the simple network containing three neurons shows stimulation configuration with blue laser light and the connections with arrows (a). 
+Simple network containing three neurons shows stimulation configuration with blue laser light and the connections with arrows (a). 
 The neurons A and B are stimulated in 1000 trials and the corresponding peristimulus time-histogram are shown in (b) upper panel with a raster plot in the lower panel. 
-Cross-correlation histograms (CCHs) are shown in (c) where horizontal and vertical axes represent time lag in ms and counts of coincident spikes in bins of 1 ms.}
+Cross-correlation histograms (CCHs) are shown in (c) where axes represent time lag in ms and counts of coincident spikes in bins of 1 ms.}
 \label{fig:intro}
@@
 In order to estimate the actual influence of stimulation of a neuron on post-synaptic neurons, we need to distinguish the influence of one stimulated neuron from the influence of another stimulated neuron.
-We would thus need something that affects the stimulation effect separately between neurons. 
+We would thus need something that affects the stimulation effect separately across neurons. 
 Arguably, refractoriness is such a variable. 
 If a neuron is in its absolute refractory period, then no amount of stimulation will make it spike. 
-This gives us an interesting way of inferring causality, by comparing the network state between a time when a neuron is able to spike and when the neuron is unable to spike.
+This gives us an interesting way of inferring causality, by comparing the network state between a time when a neuron is able to spike and when the neuron is unable to spike when the stimulation hits.
 
-Instrumental variables require the existence of a random (or sufficiently random) variable that affects the variable of interest. 
+Instrumental variables require the existence of a variable (e.g.
+refractoriness) that affects one variable of interest (the presynaptic neuron), but affects the rest of the network (including the postsynaptic neuron) only through that variable.
 This independent influence then allows quantifying the influence of the variable of interest on the rest of the network. 
-In our case, the refractory states of a neuron is in good approximation independent on small time scales (see Discussion for caveats). 
-It affects the influence of stimulation on the putative pre-synaptic neuron (\cref{fig:cch_iv_triple}(a)). 
-The trials where a stimuli is unable to elicit a spike due to the refractory state can then be used to identify causal effect on the putative post-synaptic neuron.
+In our case, the refractory states of a neuron is in good approximation independent on short time scales (see Discussion for caveats). 
+It affects the influence of stimulation on the pre-synaptic neuron (\cref{fig:cch_iv_triple}(a)). 
+The trials where a stimulus is unable to elicit a spike due to the refractory state can then be used to identify causal effect on the putative post-synaptic neuron.
 
@@
 For such a simple system, it produces meaningful estimates of the causal interactions between neurons.
+
+\begin{figure}
 \tikzset{line/.style = {draw, -latex', ultra thick}}
-\begin{figure}
 \makeatletter
@@
 The network was tuned to be in an asynchronous regime; see \cref{fig:state}\labelcref{fig:CC} with log-normally distributed synaptic weights according to patch-clamp experiments \citep{Sayer1990,Mason1991}; see \cref{fig:state}\labelcref{fig:syn_dist} and \cref{tab:params} for parameters.
-Furthermore, we selected 800 excitatory neurons for stimulation and gave each neuron a random spatial distance from the simulated optogenetic stimulus. 
-The stimulus intensity was then set according to \cref{eq:hill} with a maximum of 8 pA, and was constant throughout trials. 
+Furthermore, we selected 800 excitatory neurons for stimulation and gave each neuron a random spatial distance from the simulated optogenetic stimulus (\cref{fig:error}(a, b)). 
+The stimulus intensity was then set according to \cref{eq:hill} with a maximum of 8 pA, and was constant throughout the trials. 
 The trial onset had a temporal Poisson distribution with period 100 ms and was further clipped between 100-150 ms. 
@@
 
-To evaluate how well the IV method estimates the weights as a function of number of trials we calculated the mean squared error of the connection weight \cref{fig:error}. 
-The IV estimator's precision decreases similarly in three different settings with varying amounts of relative inhibition $ g $.
+To compare the IV and CCH methods we calculated the mean squared error of the weight estimates as a function of the number of trials and hit rate  (\cref{fig:error}(c, d)); see methods section for details.
+The IV estimator's precision decreases similarly in three different settings with varying amounts of relative inhibition $ g $ while CCH remain constant shown in \cref{fig:error}(c) on a logarithmic scale.
+The slopes of the MSE for IV was found to be $ -0.53, -0.48, -0.31 $ for $g = 9.9, 4.4, 3.0$ respectively. 
+Furthermore, the MSE increases with hit rate for both IV and CCH, where hit rate $= 1$ indicates that the stimulation induces a spike for each trial.
 \begin{figure}
@@
 
-\caption{{\bf Mean square error (MSE) of IV and CCH based estimators in a network of two populations.} 
-(a) Raster plots showing inhibitory neurons (upper panel, red) and excitatory neurons (lower panel, blue) are stimulated with varying intensity with the strongest at lower neuron number.
-(b) Histogram over two stimulations, all neurons where time zero indicates stimulation onset, stimulated excitatory neurons drive inhibition which then silents the entire network.
-(c, d) The IV- and CCH estimators are evaluated for the asynchronous recurrent neural network at three different amounts of relative inhibition $ g $. 
-The MSE as a function of number of trials is shown in (c) on a logarithmic scale where the slopes was found to be $ -0.53, -0.48, -0.31 $ for $g = 9.9, 4.4, 3.0$ respectively. 
-Finally, the MSE as a function of stimulation effect (hit rate) is shown in (d) where $hit rate = 1$ indicates that the stimulation induces a spike for each trial. \label{fig:error}}
+\caption{{\bf Mean square error (MSE) of IV and CCH estimators in a network of two populations.} 
+(a) Raster plots showing inhibitory neurons (upper panel, red) and excitatory neurons (lower panel, blue) stimulated with varying intensity with the strongest at lower neuron number.
+(b) Histogram of all neurons, where time zero indicates stimulation onset (marked), stimulated excitatory neurons drive inhibition which then silents the entire network.
+(c, d) The IV- and CCH estimators are evaluated for the recurrent neural network at three different amounts of relative inhibition $ g $ as a function of (c) number of trials and (d) hit rate. 
+ \label{fig:error}}
 \end{figure}
@@
 We then compared the IV estimator which exploits the refractory period, with the CCH method given by \cref{eq:ptrans} which ignores network confounding. 
-We defined hit rate as the relative amount of stimulations in non-refractory states and required a minimum of 90$\%$. 
+To get a good comparison between refractory and non-refractory states we required a maximum of 90$\%$ hit rate. 
 To indicate the amount of connections that are falsely attributed to a non-zero weight, we calculated the amount of false positives. 
 This was given as the percentage of estimated synapses larger than $ 0.05 $ where the true weight was $ 0 $, finding $ 99.4\% $ for CCH and $ 0.2\% $ for the IV estimator; see \cref{fig:fitness}(a). 
-In addition, we compared the size of the estimations at false positive instances and found that the CCH method have significantly higher median than IV (p=0, difference ($\Delta$) = 0.164, permutation resampling \citep{wassermann2006all}).
-The IV approach, while not being perfect, thus outperforms the CCH approach.
+In addition, we compared the size of the estimates at false positive instances and found that the CCH method have significantly higher median than IV (p=0, difference ($\Delta$) = 0.164, permutation resampling \citep{wassermann2006all}).
+The IV approach, while not being perfect, thus considerably outperforms the CCH approach.
 
@@
 We thus performed a logistic regression \cref{fig:fitness}a denoted LOGIT. 
-Here, we show that LOGIT performs worse than CCH illustrating the advantage of using the refractory period as an instrumental variable (p=0, $\Delta$ = 0.577, permutation resampling). 
-To further evaluate the methods, we calculated false negatives as instances where the true weight is non-zero but estimated to be zero in \cref{fig:fitness}(b) shows that the CCH and IV estimators perform equally well (p=0.59, $\Delta$ = 0.017, permutation resampling), while LOGIT have no false negatives.
+Here, we show that LOGIT performs worse than IV (and even CCH)
+illustrating the advantage of using the refractory period as an instrumental variable (p=0, $\Delta$ = 0.577, permutation resampling). 
+To further evaluate the methods, we calculated false negatives as instances where the true weight is non-zero but estimated to be zero in \cref{fig:fitness}(b) shows that the CCH and IV estimators perform equally well on that measure (p=0.59, $\Delta$ = 0.017, permutation resampling), while the LOGIT has no false negatives.
 Finally, we wanted to evaluate the estimated weights as a function of true weights shown in \cref{fig:fitness}(b,c) after 30000 trials. 
-The IV estimator yields a good prediction ($ R^2 = 0.770 $), while the CCH method mainly estimates the strength of stimulation while true weights are poorly estimated ($ R^2 = 0.009 $).
+The IV estimator yields a good prediction ($ R^2 = 0.77 $), while the CCH method mainly estimates the strength of stimulation while true weights are poorly estimated ($ R^2 = 0.01 $).
 Utilizing refractory periods as an instrumental variable considerably improves the estimations.
@@
 \begin{subfigure}{\textwidth} 
-\includegraphics[scale=1]{goodness_fit}
+\includegraphics[scale=.95]{goodness_fit}
 \end{subfigure}
@@
 To test the feasibility of the IV approach on \textit{in vivo} data, we compared the CCH and IV estimators on openly available extracellular single unit recordings from two mice with light-pulse stimulations of CA1 pyramidal neurons \citep{English2017}.
-We investigated connections from potentially pre-synaptic units, that exhibited a significant increase in their firing rate upon stimulation, with potentially post-synaptic units that did not respond to the stimulation \citep{English2017}.
+We investigated connections from pre-synaptic units, that exhibited a significant increase in their firing rate upon stimulation, with putative post-synaptic units that did not respond to the stimulation \citep{English2017}.
 These experiments were not optimally designed for the use of the IV method to infer causal connectivity.
-Because of a low stimulation hit rate, we used an IV window of $ 7.5 $ ms, larger than the absolute refractory period of pyramidal cells, $ 4 $ ms.
+Because of a low stimulation hit rate, we used an IV window of $ 7.5 $ ms, larger than the absolute refractory period of pyramidal cells of about $ 4 $ ms (as estimated from autocorrelograms, data not shown).
 This allows a proof-of-principle evaluation of the IV method on experimental data.
@@
 \makeatother
-\begin{subfigure}{0.4\textwidth}
-\includegraphics[width=1.\textwidth]{Optotrode.jpg}
-%\vspace{3em}
-\caption{} \label{}
-\end{subfigure}\hfill
-\begin{subfigure}{0.6\textwidth}
-\includegraphics[width=1.\textwidth]{Optodata_comparisonIV_CCH.pdf}
-\caption{} \label{}
-\end{subfigure}
-\caption{\textbf{IV and CCH estimates on hippocampal recording}
-We compared the IV and CCH estimates of synaptic strengths on CA1 recordings with optogenetic activation of pyramidal cells \citep{English2017}.
-(a) Left: schematic of the $\mu$-LED silicon probe, right: photograph of one shank tip with three illuminated $\mu$-LEDs with a scale bar at $15 \mu m$.
-(b) IV and CCH estimates for connections from optogenetically activated to non-activated units that showed a significant peak in the CCH \citep{English2017}.
+\centering
+\includegraphics[width=0.6\textwidth]{Optodata_comparisonIV_CCH.pdf}
+\caption{\textbf{IV and CCH estimates on hippocampal recordings.}
+IV and CCH estimates for connections from optogenetically activated to non-activated units that showed a significant peak in the CCH \citep{English2017}.
 Dashed line represents identity. 
-Error bars are $ \pm $ one standard deviation.
+Error bars are 95\% confidence intervals calculated by bootstrapping.
 }
@@
 \FloatBarrier
+
 \section{Discussion}
@@
 For example, if we have balanced excitation and inhibition, we may expect largely independent refractory states of individual neurons. 
-If a neuron biophysically implements something like conditional Poisson spiking, its refractory states will be random. 
+If a neuron biophysically implements something like conditional Poisson spiking, its refractory states will be random conditioned on the network state. 
 Importantly, we may expect the phase of a neuron to be far more random than the activity of the network as a whole.
@@
 Giving one burst of stimulation which is strong enough to elicit multiple spikes from each neuron may effectively randomize their phases \citep{ermentrout2008reliability}.
-Third, we may utilize chemical, behavioral, or molecular perturbations to randomize refractoriness. 
-For example, we may be able to construct intracellular oscillators that are unaffected by neural activities. 
-To the best of our knowledge, there has not been any attempts in neuroscience to produce good instrumental variables, so there may be many possibilities for improvements.
+Third, we may utilize chemical, behavioral, or molecular perturbations to produce a good instrumental variable. 
+For example, we may be able to construct intracellular oscillators that are unaffected by neural activities or constructs that force a neuron into quiescence at random times. 
+In neuroscience there has been no effort yet to produce good instrumental variables, so there may be many possibilities for improvements.
 
@@
 These ideas should, arguably, be merged with IV approaches. 
-The main strength of the IV estimator presented here is that it only requires one pair to be recorded because we can utilize the randomness of refractory periods.
+One of the strengths of the IV estimator presented here is that it only requires one pair to be recorded because we can utilize the randomness of the refractory periods along with random stimulations.
+Under those assumptions, the IV estimator can produce actual causal estimates.
 
@@
 This is due to the inverse relation between changes in light intensity and affected number of neurons combined with a logarithmic relation between light intensity and photocurrent \cite{wang2007high}.
-In addition, the distribution of membrane potentials across neurons is flat \citep{destexhe1999impact,rudolph2006use,pare1998impact} making neurons highly sensitive to perturbations. 
+In addition, the distribution of membrane potentials across neurons is relatively flat \citep{destexhe1999impact,rudolph2006use,pare1998impact} making neurons highly sensitive to perturbations. 
 One could however, imagine situations where optogenetic activation was more local. 
@@
 There could also be other ways of making optogenetic stimulation more local. 
-For example, if one engineered opsins or brain tissue that are more light absorbent one could stimulate more locally. 
+For example, if one engineered opsins or brain tissue that are more light absorbent (e.g. by ubiquitously producing melanin) one could stimulate more locally. 
 How to engineer more localized stimulation is an important problem when causally interrogating a system.
@@
 There are many techniques for causal inference, most of which are largely unknown to the field of neuroscience, and are based on approximating randomness in a seemingly regular world. 
-In many cases, one could use regression discontinuity designs in a spiking system \citep{BenPaper, imbens2008regression}. 
+In many cases, one could use regression discontinuity designs in a spiking system \citep{lansdell2018spiking, imbens2008regression}. 
 Moreover, one could use a difference in difference approach \citep{abadie2005semiparametric}. 
@@
 \subsection{Simulated network}
-To simulate a recurrent network of excitatory and inhibitory neurons we used the leaky integrate and fire (LIF) model given by
+To simulate a recurrent network of excitatory and inhibitory neurons we used the LIF model given by
 \begin{equation}
@@
 \makeatother
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{CC}
+\begin{subfigure}{0.5\textwidth} \includegraphics[scale=.9]{CC}
 \caption{} \label{fig:CC}
-\end{subfigure}\hfill
-\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{CV}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth} \includegraphics[scale=.95]{CV}
 \caption{} \label{fig:CV}
@@
 \subsection{Perturbation intensity}\label{sec:method:opto}
-In order to replicate an optogenetic experiment we modeled transmission of light through brain tissue with the Kubelka-Munk model for diffuse scattering in planar, homogeneous, ideal diffusing media \citep{Ho2017} given by
+In order to replicate an optogenetic experiment we modeled transmission of light through brain tissue with the Kubelka-Munk model for diffuse scattering in planar, homogeneous media \citep{Ho2017} given by
 \begin{align}
@@
 \end{align}
-Here $ T $ denotes a transmisison fraction, $ S $ is the scattering coefficient for mice \citep{Aravanis2007} and $ r $ is the distance from a light source. 
+Here $ T $ denotes a transmission fraction, $ S $ is the scattering coefficient for mice \citep{Aravanis2007} and $ r $ is the distance from a light source. 
 Further we combined diffusion with geometric loss assuming that absorption is negligible as in \cite{Aravanis2007} and computed the intensity as presented in \cref{fig:concept} by
@@
 Each recording lasted more than $3$ hours.
-Spikes were then sorted by the experimenter using \todo{klustakwik??}.
+Spikes were then sorted by the experimenter using kilosort \citep{pachitariu2016kilosort}.
 During analysis, we treated each recording day independently.
@@
 By cross-correlating stimulation onset time with spikes, we quantified the required time for each intensity group to significantly increase the firing rate above baseline.
-We used a bin-width of $3$ ms, calculated the baseline by taking the mean of the cross-correlogram for negative time lags, $-45$ to $-1.5$ ms, and applied a significance threshold of $0.01$ on the probability to obtain the observed or a higher count per bin of the poisson distribution with continuity correction \citep{abeles1982quantification}.
+We first binned spikes with $3$ ms bin-width and then calculated the baseline by taking the mean of the cross-correlogram for negative time lags, $-45$ to $-1.5$ ms.
+Finally, we applied a significance threshold of $0.01$ on the probability to obtain the observed or a higher count per bin of the poisson distribution with continuity correction \citep{abeles1982quantification}.
 For each labeled neuron, those intensities were selected that caused a significant increase in firing probability within $7.5$ ms.
@@
 For computing the IV estimate, we used a window size of $7.5 ms$ and a fixed synaptic delay of $1 ms$.
+
 IV and CCH estimates were separately bootstrapped with a sample size of $1000$.
 For CCH estimates, we employed the additive property of the CCH function.
-We divided spiketrains of a session in chunks of $100 s$, and calculated the CCH for each chunk.
-We then randomly selected chunks, with replacement, to match the full length of the session.
-CCH estimates were calculated on the sum of chunks.
-For the IV estimates, we randomly chose stimulations with replacement to match the original number of stimulations and calculated the estimate on those samples.
+We subdivided spiketrains of a session in in $100 s$ segments, and calculated the CCH for each segment individually.
+We then randomly selected segments with replacement, to match the full length of the session.
+Estimates from CCH were calculated on the sum of segments.
+For the IV estimates, we detected, as before, for each optogenetic stimulation event wether pre- and postsynaptic units were active.
+We then randomly selected stimulation events with replacement to match the original number of stimulations and calculated the IV estimate on the random sample.
 
@@
 \section{Acknowledgments}
-This research was partly funded by the Research Council of Norway Grant 231248 and the University of Oslo.
+This research was partly funded by the National Institutes of Health (NIH) grant  R01MH103910 and the University of Pennsylvania.
+In addition, the Research Council of Norway Grant 231248 and the University of Oslo.
 T.S (PhD fellow), M.F. (PI) are part of the Simula-UCSD-University of Oslo Research
-and PhD training (SUURPh) program, an international collaboration in computational biology and
-medicine funded by the Norwegian Ministry of Education and Research.
-We thank Sam McKenzie and Daniel Fine English for sharing and explaining the optotrode data.
+and PhD training (SUURPh) program, an international collaboration in computational biology and medicine funded by the Norwegian Ministry of Education and Research.
+We thank Sam McKenzie, Daniel Fine English from the lab of György Buzsáki for sharing and explaining the optotrode data.
 %-------------------------------------------------------------------------------
@@
 \bibliography{library}}
-%-------------------------------------------------------------------------------
-%                                 TODOS
-%-------------------------------------------------------------------------------
-%\newpage
-%\listoftodos
+
 \end{document}

### 2018-11-02 tools_experimentaldata.py
remove exana dependency

@@ -12,20 +12,15 @@ import collections
 import scipy as sc
 from method import IV
 from matplotlib import pyplot as plt
-from tools import despine
+from tools_plot import despine
 from tqdm import tqdm
 import statsmodels.api as sm
 import copy
 from skimage import measure
 import elephant.conversion as conv
 import elephant.spike_train_correlation as corr
-sys.path.append('../exana/'),
-from exana.statistics.tools import (ccg_significance,
-                                    ccg,
-                                    hollow_kernel,
-                                    correlogram)
-from exana.statistics.tools import poisson_continuity_correction as pcc
-from cross_correlation import cch_convolve
+from tools_analysis import cch_significance, hollow_kernel, correlogram, cch_convolve
+from tools_analysis import poisson_continuity_correction as pcc
 
 
 def loadmat(filename):
@@ -164,7 +159,7 @@ def load_unitlabels(path_to_optolabels, data_dir):
                                                 'cell_id'])
     units_db_tagged_pre['class'] = 'tagged_pre'
     units_db_tagged_pre['class_index'] = units_db_tagged_pre.index
-    
+
     units_db_tagged_post = pd.DataFrame(tagged_post,
                                         columns=['Sess_id',
                                                  'shank_id',
@@ -188,7 +183,7 @@ def load_unitlabels(path_to_optolabels, data_dir):
                                                     'cell_id'])
     units_db_nottagged_post['class'] = 'nottagged_post'
     units_db_nottagged_post['class_index'] = units_db_nottagged_post.index
-    
+
     units_db = pd.DataFrame(columns=['Sess_id',
                                      'shank_id',
                                      'unit_id',
@@ -255,7 +250,7 @@ def download_files_by_dict(file_dict,
 def download_files(units_db, link_db,
                    files_ext_general,
                    files_ext_by_shank):
-    
+
     for i, row in units_db.iterrows():
         dir_i = row['animal']
         date_i = row['date']
@@ -321,14 +316,14 @@ def create_neo_structure(file_dict,
                 clusters_i = clusters_i[1:]
                 # make sure that number of clusters match
                 assert n_clusters_i == len(np.unique(clusters_i))
-                
+
[…]

### 2018-11-02 manuscript/main.tex
update manuscript

@@
 Dashed line represents identity.
-Error bars are 95\% confidence intervals calculated by bootstrapping.
+Blue dots are median values and error bars indicate the $95$\% confidence intervals calculated by bootstrapping.
+For better visualization, we omitted one estimated connection, $CCH = 0.01$ and $IV = 0.036$.
 }
@@
 Estimates from CCH were calculated on the sum of segments.
-For the IV estimates, we detected, as before, for each optogenetic stimulation event wether pre- and postsynaptic units were active.
-We then randomly selected stimulation events with replacement to match the original number of stimulations and calculated the IV estimate on the random sample.
+For the IV estimates, we detected, for each optogenetic stimulation event, wether pre- and postsynaptic units are active.
+Further, we randomly selected stimulation events with replacement to match the original number of stimulations and calculated the IV estimate on the random sample.
 

### 2018-11-02 tools_experimentaldata.py
took pdb out of tests

@@ -1,4 +1,3 @@
-import sys
 import os
 import numpy as np
 import neo
@@ -17,15 +16,10 @@ from tqdm import tqdm
 import statsmodels.api as sm
 import copy
 from skimage import measure
-import elephant.conversion as conv
-import elephant.spike_train_correlation as corr
-sys.path.append('../exana/'),
-from exana.statistics.tools import (ccg_significance,
-                                    ccg,
-                                    hollow_kernel,
-                                    correlogram)
-from exana.statistics.tools import poisson_continuity_correction as pcc
-from cross_correlation import cch_convolve
+from tools_analysis import (cch_significance,
+                            hollow_kernel,
+                            correlogram,
+                            poisson_continuity_correction as pcc)
 
 
 def loadmat(filename):
@@ -1289,7 +1283,7 @@ def calculate_transmission_prob(blks,
                             cond_post and
                             len(spktr_pre) > 0 and
                             len(spktr_post) > 0):
-                        pcausal, pfast, bins, cch, cch_s = ccg_significance(
+                        pcausal, pfast, bins, cch, cch_s = cch_significance(
                             spktr_pre,
                             spktr_post,
                             limit=ccg_time_limit,
@@ -2169,3 +2163,32 @@ def bootstrap_iv(
                       ignore_index=True)
             
     return df
+
+
+def ccg(t1, t2=None, binsize=1*pq.ms, limit=100*pq.ms, auto=False, density=None,
+        **kwargs):
+    '''
+    Wrapper for cross_correlation_histogram from elephant.
+
+    Note
+    ----
+    returns bins at right edges
+    '''
+    import elephant.spike_train_correlation as corr
+    import elephant.conversion as conv
+    if auto: t2 = t1
+    cch, bin_ids =  corr.cross_correlation_histogram(
+        conv.BinnedSpikeTrain(t1, binsize),
+        conv.BinnedSpikeTrain(t2, binsize),
+        window=[-limit, limit],
+        **kwargs)
+    bins = cch.times
+    cch = cch.magnitude.flatten()
+    if auto:
+        # Compensate for the peak at time zero that results in autocorrelations
+        # by subtracting the total number of spikes from that bin. Note
+        # possible numerical issue here because 0.0 may fall at a bin edge.
+        c_temp, bins_temp = np.histogram([0.], bins=bins)
+        bin_containing_zero = np.nonzero(c_temp)[0][0]
+        cch[bin_containing_zero] = 0
+    return cch, bins
